Human-specific lncRNAs contributed critically to human evolution by distinctly regulating gene expression

Version 1

Abstract

Abstract

What genomic sequences make protein-coding genes generate divergent expression in closely related species, specifically, differentiate humans from apes, puzzle many researchers. Many studies examined species-specific gene birth, gene loss, and changes in promoters and transcription factor binding sites, but the identification and impact of human-specific lncRNAs remain unexplored. This study identified human-specific lncRNAs from GENCODE-annotated human lncRNAs, predicted their DNA binding sites (DBSs) genome-wide, and analyzed the DBSs and their counterparts in modern humans (CEU, CHB, and YRI), archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and chimpanzees. The results reveal how human-specific lncRNAs and their DBSs have transcriptionally regulated gene expression human-specifically. The rewiring of gene expression has undergone continuous evolution, significantly changed gene expression in the brain, promoted the adaptive evolution of humans, and influenced differences in modern humans. These results reveal the importance of human-specific lncRNAs (for human evolution) and highlight the importance of other species-specific lncRNAs.

Introduction

Genomic changes that drive evolution include gene birth (Kaessmann, 2010), gene loss (Albalat and Canestro, 2016), and changes in regulatory sequences (Prud’homme et al., 2007). Small genomic differences between humans and other hominids, but significant phenotypic and behavioral differences, make what genomic changes at what time critically drove human evolution an enduring puzzle (Anton et al., 2014; Pollen et al., 2023). Waves of studies have identified genes that promote human brain development (Evans et al., 2005; Fiddes et al., 2018; Mekel-Bobrov et al., 2005; Pinson et al., 2022; Suzuki et al., 2018). However, conclusions from gene birth-centered studies are controversial (Currat et al., 2006; Timpson et al., 2007; Yu et al., 2007). In particular, genes important for brain development (e.g., NOTCH2NL and ASPM) do not necessarily account for many human traits that preceded or coincided with brain enlargement. Recent studies revealed the importance of regulatory sequences for human evolution, including new sequences (Liu et al., 2021), lost sequences (McLean et al., 2011), sequences showing accelerated evolution (Dong et al., 2016; Mangan et al., 2022; Prabhakar et al., 2006; Whalen and Pollard, 2022), and the turnover of transcription factor binding sites (TFBS) (Krieger et al., 2022; Otto et al., 2009; Zhang et al., 2023). An unexplored question is how simian- and human-specific lncRNAs, with the sequences of their DNA binding sites, regulate conserved genes and drive human evolution.

On this question, experimental studies have generated multiple findings. (a) The chimpanzee and human genomes are almost 99% identical (Chimpanzee Sequencing and Analysis Consortium, 2005). (b) About one-third of human lncRNAs are primate-specific (Derrien et al., 2012). (c) Species-specific lncRNAs have distinct expression profiles in organs of different species (Sarropoulos et al., 2019). (d) Many lncRNAs can bind to DNA sequences by forming RNA:DNA triplexes (Abu Almakarem et al., 2012), thereby recruiting histone and DNA modification enzymes to binding sites to regulate nearby gene expression (Lee, 2009; Roadmap Epigenomics et al., 2015). (e) Approximately 40% of differentially expressed genes in humans result from interspecies epigenetic differences (Hernando-Herraez et al., 2015). Compared with gene-centered hypotheses of human evolution, these findings suggest an alternative hypothesis – distinctly rewired gene expression by human-specific (HS) lncRNAs and their DNA binding sites has critically promoted human evolution and determined human traits.

Orthologous lncRNA genes can be identified by sequence search by structure and sequence alignment (Nawrocki, 2014; Nawrocki and Eddy, 2013). HS lncRNAs can be identified upon orthologues of human lncRNAs in other mammals. RNA:DNA triplexes follow specific base-pairing rules (Abu Almakarem et al., 2012). These rules allow the prediction of lncRNAs’ DNA binding domains (DBD) and DNA binding sites (DBS) (DBDs and DBSs indicate predicted sequences of DNA binding domains and binding sites) (Lin et al., 2019; Wen et al., 2022). Gene expression in most human organs and tissues has been profiled (GTEx Consortium, 2017). Genomes of modern humans, archaic humans, and multiple apes have been sequenced (1000 Genomes Project Consortium et al., 2012; Chimpanzee Sequencing and Analysis Consortium, 2005; Meyer et al., 2012; Prufer et al., 2017; Pru fer et al., 2013). These results enable us to examine the above alternative hypothesis.

HS lncRNAs and their DBSs may have occurred in three situations – (a) HS lncRNAs occurred before their DBSs, (b) HS lncRNAs and their DBSs occurred simultaneously, (c) HS lncRNAs occurred after their DBSs. It is also unknown which situation the rewiring hypothesis corresponds to. In this study, we identified HS lncRNAs upon the orthologues of the GENCODE-annotated human lncRNA genes in 16 mammalian genomes and predicted HS lncRNAs’ DBDs and DBSs in the genome of modern humans (CEU, CHB, and YRI) and these DBSs’ counterparts in the genomes of archaic humans (Altai Neanderthals, Denisovans, Vindija Neanderthals), and chimpanzees (Figure 1A). Then, we analyzed the sequences and evolution of HS lncRNAs and their DBSs, functional enrichment of target genes, and impact of HS lncRNA regulation on gene expression in GTEx organs and tissues in archaic and modern humans. Our genome and transcriptome analyses suggest that HS lncRNAs and their DBSs have substantially and distinctly rewired gene expression during human evolution. The rewiring of gene expression also greatly determines differences between human populations.

Study Overview. (A) The relationships between the chimpanzees, the three archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and modern humans. The dashed lines indicate the phylogenetic distances from modern humans to archaic humans and chimpanzees. The genes and DBSs have the following implications – the DBS in B2M lacks a counterpart in chimpanzees; the DBS in ABL2 has great differences in archaic humans; the DBS in IRNAR1 is highly polymorphic in modern humans. Red characters in DBSs indicate tissue-specific eQTLs or population-specific favored mutations. (B) The distributions of DBSs’ length and binding affinity. (C) The distributions of the numbers of HS lncRNAs’ target genes and transcripts. (D) The targeting relationships between HS lncRNAs. (E) The distributions of sequence distances of DBSs from modern humans to chimpanzees and archaic humans. (F) The impacts of HS lncRNA regulation on gene expression in GTEx tissues.

Results

HS lncRNAs regulate diverse genes and transcripts

After searching for the orthologues of the 13562 GENCODE-annotated human lncRNA genes (v18) in 16 mammalian genomes by sequence and structure alignment using the Infernal program (Lin et al., 2019; Nawrocki, 2014; Nawrocki et al., 2009), we identified 66 HS lncRNA genes (Supplementary Table 1). We predicted HS lncRNAs’ DBSs in the 5000 bp promoter regions of the 179128 Ensembl-annotated transcripts (release 79) using the LongTarget program (Supplementary Note 1) (Lin et al., 2019). 105141 strong DBSs (binding affinity > 60, mean length > 147 bp) were identified in 96789 transcripts of 23396 genes, and 152836 weak DBSs (36 < binding affinity < 60) were identified in 127898 transcripts of 33185 genes (Figure 1B; Supplementary Table 2, 3). Genes and transcripts whose promoter regions contain the DBS of an HS lncRNA were assumed targets of the HS lncRNA. Several HS lncRNAs have DBSs in nearly all genes and transcripts (e.g., RP11-423H2.3 has strong DBSs in 73855 transcripts of 17941 genes), and only about 1.5% of target genes (0.6% of target transcripts) are human-specific (Figure 1C). Many transcripts have DBSs for multiple HS lncRNAs, many HS lncRNAs have multiple DBSs in a single transcript, and HS lncRNAs themselves show complex targeting relationships (Figure 1D; Supplementary Note 2). These findings suggest that HS lncRNAs substantially regulate gene expression.

DBS prediction was validated and supported by multiple methods and datasets (Supplementary Note 1; Supplementary Fig 1-5). (a) DBSs of HS lncRNAs overlap well with experimentally identified DNA methylation and histone modification signals in multiple cell lines (https://genome.UCSC.edu). (b) DBSs of NEAT1 and MALAT1 overlap well with experimentally identified DNA binding sites of NEAT1, MALAT1, and MEG3 (Mondal et al., 2015; West et al., 2014). (c) Many DBSs co-localize with annotated cis-regulatory elements (cCREs) in genes’ promoter regions (https://genome.UCSC.edu). (d) We used CRISPR/Cas9 to knock out the DBD (100-200 bp) of seven lncRNAs (three HS lncRNAs and four wrongly transcribed lncRNAs) in multiple cell lines and performed RNA-seq before and after DBD knockout. Gene expression analysis revealed that the |fold change| of target genes was significantly larger than the |fold change| of non-target genes (one-sided Mann-Whitney test, p = 3.1e-72, 1.49e-114, and 1.12e-206 for the HS lncRNA RP13-516M14.1, RP11-426L16.8, and SNORA59B, and p = 2.58e-09, 6.49e-41, 0.034, and 5.23e-07 for the four wrongly transcribed lncRNAs). The results suggest that the changed expression of target genes was caused by the knockout of DBDs. (e) Finally, the LongTarget program uses the Smith-Waterman local alignment to simultaneously identify DBDs and DBSs. The alignment algorithm determines that a DBS of 147 bp is extremely unlikely to be generated by chance (p < 8.2e-19 to 1.5e-48).

Target genes characterize human evolution and traits

Since strong and weak DBSs could occur early and late in human evolution, we examined whether their target genes were enriched in specific functions important for human evolution. We sorted target genes by DBS’s binding affinity, chose the top 2000 genes (DBS sequence > 252 bp, DBS binding affinity > 151) and bottom 2000 genes (DBS sequence < 60 bp, DBS binding affinity > 36), and performed over-representation analysis using the g:Profiler program and Gene Ontology (GO) database (Supplementary Table 4). The two target gene sets are co-enriched in GO terms important for human evolution, including “neuron projection development” (GO:0031175), “learning” (GO:0007612)), “regulation of anatomical structure size” (GO: GO:0090066), “head development” (GO:0060322), and “regulation of locomotion” (GO:0040012)) (Supplementary Note 3; Supplementary Table 4). 514 DBSs have binding affinity > 300 (DBS sequences > 500 bp) (Supplementary Table 5). Genes containing these strongest DBSs include IFNAR1 and NFATC1 (important for immune function), KIF21B and NTSR1 (critical for neural development), SLC2A11 and SLC2A1 (involved in glucose usage), BAIAP3 and TAS1R3 (a brain-specific angiogenesis inhibitor and a receptor for the sweet taste response), and some primate-specific noncoding RNA genes (e.g., CTD-3224I3.3 which is highly expressed in the cerebellum, lung, and testis) (Table 1). Interestingly, genes with weak DBSs are enriched in GO terms including “response to alcohol” (GO:0097305), “response to temperature stimulus” (GO:0009266), and “DNA methylation or demethylation” (GO:0044728), and “female pregnancy” (GO:0007565) (Supplementary Note 3; Supplementary Table 4).

The rewiring of gene expression has greatly evolved

The human genome is almost 99%, 99%, and 98% identical to the genomes of chimpanzees, bonobos, and gorillas (Chimpanzee Sequencing and Analysis Consortium, 2005; Prufer et al., 2012). Given these HS lncRNA genes, we assessed to what extent their DBS sequences are human-specific (whether DBSs occurred before, at the same time, or after HS lncRNAs). 97.81% of strong DBSs were found to have counterparts in chimpanzees, suggesting that most DBS sequences occurred before HS lncRNAs.

To examine the evolution of HS lncRNAs and their DBSs during human evolution, we identified their counterparts in Altai Neanderthals, Denisovans, and Vindija Neanderthals (Meyer et al., 2012; Prufer et al., 2017; Pru fer et al., 2013) (Figure 1A). HS lncRNAs and strong DBSs have counterparts in these archaic humans, but the distances of HS lncRNAs are small than the distances of DBSs between archaic and modern humans (all “distance” means distance per base). We computed DBS distances in two ways – from the human ancestor (downloaded from the EBI website) to chimpanzees, archaic humans, and modern humans (the first type of distances) and between the modern humans and chimpanzees and archaic humans (the second type of distances). In the first type of distances, many DBSs’ distances from the human ancestor to modern humans were smaller than to archaic humans. For example, DBSs in the top 20% (5033) genes in the human ancestor have distances > 0.015 to modern humans, but more DBSs have distances > 0.015 to Altai Neanderthals, Denisovans, and even chimpanzees (Supplementary Note 3). This defect may be caused by that the human ancestor was built using six primates without archaic humans. The second type of distances agree well with the phylogenetic distances between chimpanzees, archaic humans, and modern humans and suggest that greatly changed sequences occurred in very limited DBSs (Figure 1AE). For example, DBSs in the top 20% (4248) genes in chimpanzees have distances > 0.034, and DBSs having distances > 0.034 in genes in Altai Neanderthals, Denisovans, and Vindija Neanderthals are 1256, 2514, and 134 (Supplementary Table 6).

To examine whether the target genes of the mostly changed DBSs in chimpanzees and archaic humans are enriched in specific functions, we performed over-representation analysis for genes with using the g:Profiler program and Gene Ontology (GO) database. Since 4248, 1256, and 2514 genes in chimpanzees, Altai Neanderthals, and Denisovans have DBS distances > 0.034 (to modern humans), to make the statistical significance comparable, we examined the top 1256 genes in each analysis. The two sets of genes with largest DBS distances from modern humans to chimpanzees and Altai Neanderthals are enriched in many common GO terms, but “behavior” (GO:0007610) and “cognition” (GO:0050890) have higher significance in Altai Neanderthals than in chimpanzees (Figure 2). In addition, only genes in Altai Neanderthals are also enriched in such GO terms as “learning” (GO:0007612), “learning or memory” (GO:0007611), and “glucose homeostasis” (GO:0042593). The 134 genes whose DBS distances > 0.034 from Vindija Neanderthals to modern humans, as expected, are enriched only in two GO terms with low significance, including “smoothened signaling pathway” (GO:0007224) and “regulation of smoothened signaling pathway” (GO:0008589) (Supplementary Table 7). These results reveal that critical DBS changes occurred in genes involved in brain functions and some metabolism pathways in early archaic humans.

1256 target genes whose DBSs have largest distances between modern humans and chimpanzees and Altai Neanderthals are enriched in different Biological Processes GO terms. Upon significance threshold = 0.05 (Benjamini-Hochberg FDR), the two sets of genes in chimpanzees and Altai Neanderthals are enriched in 199 and 409 GO terms (50 < terms size < 1000), respectively. Pink color indicates shared GO terms. Left: Top GO terms enriched in genes in chimpanzees. Right: Top GO terms enriched in genes in Altai Neanderthals.

Gene regulation by HS lncRNAs shows differences in modern humans

We next used the 1000 Human Genome data to analyze single nucleotide polymorphisms (SNPs with minimal allele frequency (MAF) ≥ 0.1) in DBSs (1000 Genomes Project Consortium et al., 2012). Many DBSs were found to be highly polymorphic (Supplementary Table 8). The most SNP-rich DBSs (> 10 SNPs, 5 times the average number of SNPs for all DBSs) are in genes important for human evolution, including IFNAR1, NFATC1, TAS1R3, INS, ST3GAL4, and FN3KRP (Table 2). These SNPs affect DBS binding affinity, no matter they are favored, neutral, or recessive. A feature of modern human diets is high sugar intake, which causes non-enzymatic oxidation of proteins (i.e., glycation) and deactivates protein functions. Among proteins encoded by genes with the most SNP-rich DBSs, TAS1R3 recognizes diverse natural and synthetic sweeteners, Insulin decreases blood glucose concentration, ST3GAL4 regulates protein glycosylation, and FN3KRP deglycates proteins. It worth noting that DBSs in these genes have counterparts with mostly changed sequences in chimpanzees and archaic humans (Table 2), indicating human-specific evolution of the regulation of these genes.

We further used multiple statistical tests, including XP-CLR (Chen et al., 2010), iSAFE (Akbari et al., 2018), Tajima’s D (Tajima, 1989), Fay-Wu’s H (Fay and Wu, 2000), the fixation index (Fst) (Weir and Cockerham, 1984), and linkage disequilibrium (LD) (Slatkin, 2008), to detect selection signals in HS lncRNAs and strong DBSs in the human population CEU, CHB, and YRI (the representative populations of Europeans, Asians, and Africans). These tests detected selection signals in several HS lncRNA genes in CEU and CHB but not in YRI and selection signals in more DBSs in CEU and CHB than in YRI (Supplementary Note 4, 5; Supplementary Table 9, 10, 11). Examples include the DBSs in the two pigmentation-related genes MC1R and MFSD12 and the two odor reception-related genes OR6C1 and TAS1R3, which may reflect the adaptation of Eurasians to new environments and lifestyles.

SNPs in DBSs have cis-effects on gene expression

To examine the impact of DBS sequence differences on gene expression, we assessed the expression quantitative trait loci (eQTLs) in DBSs using the Genotype-Tissue Expression (GTEx) data (GTEx Consortium, 2017). 1727 SNPs with MAF ≥ 0.1 in at least one population and absolute values of cis-effect size (ES) ≥ 0.5 in at least one tissue were identified in DBSs in autosomal genes (Supplementary Table 12,13). These eQTLs include 372 “conserved” ones (in DBSs in the three archaic and three modern humans) and 1020 “novel” ones (in DBSs only in the three modern humans). Many conserved eQTLs are in the brain and have high derived allele frequencies (DAFs) in all three modern populations. A notable eQTL that has high DAFs in all three populations and positive ES ≥ 0.5 in 44 out of 48 GTEx tissues is rs2246577 in the DBS in FN3KRP, whose encoded protein deglycates proteins to restore their function (Table 1). On the other hand, many novel eQTLs are in a particular tissue and have population-specific DAFs (Supplementary Note 6).

We performed two analyses to examine that eQTLs in DBSs influence gene expression due to lncRNA-DBS interactions. First, we computed the expression correlation between HS lncRNAs and target transcripts with DBSs having eQTLs in specific tissues. Most (94%) HS lncRNA-target transcript pairs showed significant expression correlation in the tissues the eQTLs were identified (|Spearman’s rho| > 0.3 and FDR < 0.05). Second, we examined whether eQTLs are more enriched in DBSs than in promoter regions by computing and comparing the density of eQTLs in each DBS and each Ensembl-annotated promoter. The results confirm that eQTLs are more enriched in DBSs (one-sided Mann-Whitney test, p = 0.0) (Supplementary Note 6). Thus, many SNPs in DBSs have tissue-and population-specific cis-effects on gene expression.

HS lncRNAs contributed to brain evolution from archaic to modern humans

We finally examined the impact of HS lncRNAs and their DBSs on gene expression in the GTEx tissues (GTEx Consortium, 2017). 40 autosomal HS lncRNAs are expressed robustly (median TPM > 0.1) in at least one tissue and form 198876 pairs with target transcripts in all tissues. 45% of these pairs show a significant expression correlation in specific tissues (Spearman’s |rho| > 0.3 and FDR < 0.05). In contrast, when randomly sampling 10000 pairs of lncRNAs and protein-coding transcripts genome-wide, the percent of pairs showing expression correlation with the same significance is only 2.3%. Notably, 56% of HS lncRNA-target transcript pairs with significant correlation are present in at least one brain region (Figure 3A), indicating substantial gene expression regulation by HS lncRNAs in the brain.

The impact of HS lncRNAs on gene expression in GTEx tissues and organs. (A) Percentage distribution of correlated HS lncRNA-target transcript pairs across tissues and organs. Higher percentages of correlated pairs are in the brain than in other tissues and organs. (B) The distribution of significantly changed DBSs in HS lncRNA-target transcript pairs across tissues and organs between archaic and modern humans. Orange, red, and dark red indicate significant DBS changes from Denisovans (D), Altai Neanderthals and Denisovans (AD), and all of the three archaic humans (ADV). DBSs in HS lncRNA-target transcript pairs correlated in the seven brain regions have changed significantly and consistently since the Altai Neanderthals, Denisovans, and Vindija Neanderthals (one-sided two-sample Kolmogorov-Smirnov test, significant changes determined by FDR < 0.001).

To verify the above conclusion, we analyzed nine experimental datasets (Supplementary Note 7). First, we analyzed two datasets of epigenetic studies (one examined H3K27ac and H3K4me2 profiling of human, macaque, and mouse corticogenesis, and the other examined gene expression and H3K27ac modification in eight brain regions in humans and four other primates) (Reilly et al., 2015; Xu et al., 2018). 84% and 73% of genes in the two datasets have DBSs for HS lncRNAs, significantly more enriched in HS lncRNA targets than the genome-wide background (p = 1.21e-21 and 1.2e-56, two-sided Fisher’s exact test). Second, we analyzed two datasets of PsychENCODE studies (one examined spatiotemporally differentially expressed genes and spatiotemporally differentially methylated sites in 16 brain regions, and the other examined the spatiotemporal transcriptomic divergence across human and macaque brain development) (Li et al., 2018; Zhu et al., 2018). In the first dataset, 65 HS lncRNAs were expressed in all 16 brain regions, 109 transcripts had spatiotemporally differentially methylated sites in their promoters, and 56 of the 109 transcripts had DBSs for HS lncRNAs. In the second dataset, 8951 genes showed differential expression between the human and macaque brain, and 72% of differentially expressed genes in the human brain have DBSs of HS lncRNAs. Thus, both datasets show significant enrichment for HS lncRNA regulation. Third, three studies identified several genes critically regulating cortical expansion (Florio et al., 2018; Johnson et al., 2018; Suzuki et al., 2018). Of the 40 protein-coding genes reported by these studies, 29 have DBSs of HS lncRNAs in their promoter regions (Supplementary Table 14). Thus, these genes are enriched with DBSs of HS lncRNAs compared with the genome-wide background (p < 0.01, two-sided Fisher’s exact test). Finally, we analyzed two datasets of brain organoids studies. By establishing and comparing cerebral organoids between humans, chimpanzees, and macaques, Pollen et al. identified 261 human-specific gene expression changes (Pollen et al., 2019). By fusing human and chimpanzee iPS cells and differentiating the hybrid iPS cells into hybrid cortical spheroids, Agoglia et al. generated a panel of tetraploid human-chimpanzee hybrid iPS cells and identified thousands of genes with divergent expression between humans and chimpanzees (Agoglia et al., 2021). The 261 and 1102 genes in the two datasets are enriched with DBSs of HS lncRNAs compared with the genome-wide background (p = 1.2e-16 and 3.4e-74).

We next examined the evolution of the impact of HS lncRNAs and their DBSs on gene expression in the GTEx tissues. For each DBS in an HS lncRNA-target transcript pair correlated in a GTEx tissue, we computed the sequence distance of the DBS from modern humans to archaic humans. Then, we compared the distribution of DBS sequence distances in each tissue with the distribution of DBS sequence distances in all tissues (one-sided two-sample Kolmogorov-Smirnov test). DBSs in HS lncRNA-target transcript pairs correlated in brain regions have significantly changed sequence distances since the Altai Neanderthals and Denisovans (Figure 3B), suggesting that gene expression regulation by HS lncRNAs in the brain has experienced significant evolution and that HS lncRNAs contributed greatly to rapid brain evolution. Results from other GTEx tissues, including Kidney_Cortex, Ovary, and Skin, also suggest that the rewired gene expression has influenced human evolution.

Discussion

Modern humans differ from archaic humans and great apes little in genome sequences but greatly in phenotypes and behaviors, making what genomic differences critically determine the phenotypic and behavioral differences a compelling question. Several HS protein-coding genes promote the rapid expansion of neocortex (Fiddes et al., 2018; Florio et al., 2018; Pinson et al., 2022). Variations in TFBS cause gene expression changes (Franchini and Pollard, 2017). The rewiring of gene expression by TFs and TFBSs evolves (Johnson, 2017), so should the rewiring of gene expression by lncRNAs and their DBSs. However, the rewiring and evolution of the second case have been barely examined and poorly understood. This study identified HS lncRNAs and their DBSs, analyzed the rewiring and evolution of gene expression, and presents a new answer to the question of what genomic changes at what time critically drove human evolution. We performed seven CRISPR/Cas9 experiments to validate DBS prediction and analyzed multiple experimental datasets to validate our genome and transcriptome analyses. These results support the hypothesis that HS lncRNAs have distinctly rewired gene expression for adaptive evolution of humans.

Several findings are noteworthy. First, Neanderthal-inherited sequences have measurable impacts on gene expression, but the impacts are least detectable in the brain (McCoy et al., 2017). Here, we found that HS lncRNAs have more significantly influenced gene expression in the brain than in other tissues. Second, the regulation of genes by HS lncRNAs, making humans adapt to high sugar intake, has undergone great evolution. Third, genes regulated by HS lncRNAs are enriched in different functions at different stages. Fourth, old SNPs in DBSs are in genes regulating neural development, and new SNPs in DBSs are in genes regulating such processes as glucose metabolism.

The above findings help reveal the functions of HS lncRNAs and the adaptive evolution of humans and highlight the important functions of other species-specific lncRNAs. They also raise three questions. First, the protein-coding regions in the mouse and human genomes are about 85% identical, but many mouse lncRNAs are rodent-specific (Yue et al., 2014). Thus, a question is how mouse-specific lncRNAs specifically rewire gene expression in mice and how the rewiring influences the transcriptional differences between humans and mice (Breschi et al., 2017; Hodge et al., 2019; Zhu et al., 2018). Second, whether mice and other mammals have a lncRNA like RP11-423H2.3 that substantially rewires gene expression species-specifically. Third, the extent clade- and species-specific lncRNAs would make speciation and evolution preordained. These questions are of great interest and have practical importance.

Material and methods

Data sources

The sequences of human (GRCh38/hg38, GRCh37/hg19) and chimpanzee genomes (panTro5) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). Three high-quality archaic human genomes were obtained from the Max Planck Institute for Evolutionary Anthropology (https://www.eva.mpg.de/genetics/index.html), which include an Altai Neanderthal which lived approximately 122 thousand years ago (kya) (Pru fer et al., 2013), a Denisovan (an Asian relative of Neanderthals) which lived approximately 72 kya (Meyer et al., 2012), and a Vindija Neanderthal which lived approximately 52 kya (Prufer et al., 2017). The ancestral sequences for the human genome (GRCh37) (which were generated upon the EPO alignments of six primate species human, chimpanzee, gorilla, orangutan, macaque, and marmoset) were downloaded from the EBI website (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignm ents/). The SNP data of modern humans were obtained from the 1000 Genomes Project (phase I) (1000 Genomes Project Consortium et al., 2012). The three modern human populations are CEU (Utah residents with Northern and Western European ancestry), CHB (Han Chinese in Beijing, China), and YRI (Yoruba in Ibadan, Nigeria), which contain 100, 99, and 97 individuals, respectively. The genetic map files of modern humans were downloaded from the 1000 Genomes Project website (ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/technical/working/20110106_recombination_hotspots/). The 1000 Genomes phased genotype data in VCF format were downloaded from the International Genome Sample Resource website (http://www.internationalgenome.org/). Annotated human genes and transcripts were obtained from the Ensembl website (http://www.Ensembl.org). FASTA-format nucleotide sequences were converted from VCFs using the vcf-consensus program in the VCFtools package when necessary (Danecek et al., 2011). Data in the GeneCards Human Gene Database (www.genecards.org) were used to annotate genes and transcripts.

RNA-sequencing (RNA-seq) data (transcripts per million [TPM]) and eQTL data of human tissues were obtained from the Genotype-Tissue Expression Project (GTEx, v8) website (https://gtexportal.org/) (GTEx Consortium, 2017). Histone modification and DNA methylation signals in multiple cell lines and ENCODE Candidate Cis-Regulatory Elements (cCRE) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). The genome-wide DNA binding sites of lncRNA NEAT1, MALAT1, and MEG3 were obtained from experimental studies (Mondal et al., 2015; West et al., 2014).

Identifying HS lncRNA genes

We used the Infernal program (Nawrocki et al., 2009), which searches orthologous RNA sequences upon sequence and structure alignment, to identify orthologous exons in 16 mammalian genomes for each exon in each of the 13562 GENCODE (v18)-annotated human lncRNA genes (Lin et al., 2019). The 16 mammals were chimpanzee (CSAC 2.1.4/panTro4), macaque (BGI CR_1.0/rheMac3), marmoset (WUGSC 3.2/calJac3), tarsier (Broad/tarSyr1), mouse lemur (Broad/micMur1), tree shrew (Broad/tupBel1), mouse (GRCm38/mm10), rat (Baylor3.4/rn4, RGSC6.0/rn6), guinea pig (Broad/cavPor3), rabbit (Broad/oryCun2), dog (Broad CanFam3.1/canFam3), cow (Baylor Btau_4.6.1/bosTau7), elephant(Broad/loxAfr3), hedgehog (EriEur2.0/eriEur2), opossum (Broad/monDom5), and platypus (WUGSC 5.0.1/ornAna1) (http://genome.UCSC.edu). If the number of orthologous exons of a human lncRNA gene in a genome exceeded half the exon number of the human lncRNA gene, these orthologous exons were assumed to form an orthologous lncRNA gene. If a human lncRNA gene had no orthologous gene in all of the 16 mammals, it was assumed to be human-specific.

Identifying DBSs of HS lncRNAs

LncRNAs bind to DNA sequences by forming RNA:DNA triplexes. Each triplex comprises triplex-forming oligonucleotides (TFO) in the lncRNA and a triplex-targeting site (TTS) in the DNA sequence. We used the LongTarget program to predict HS lncRNAs’ DNA binding domains (DBD) and binding sites (DBS) with the default parameters (Ruleset = all, TT penalty = -1000, CC penalty = 0, Offset = 15, Identity ≥ 60, Nt ≥ 50) (Lin et al., 2019; Wen et al., 2022). LongTarget program predicts DBDs and DBSs simultaneously, with a DBD comprising a set of densely overlapping TTOs and a DBS comprising a set of densely overlapping TTSs. For each HS lncRNAs, we predicted its DBSs in the 5000 bp promoter regions (-3500 bp upstream and +1500 bp downstream the transcription start site) of the 179128 Ensembl-annotated transcripts (release 79). For each DBS, a binding affinity was calculated by multiplexing the DBS length and the averaged Identity score of all TTSs (the Identity score is the percentage of paired nucleotides). Strong and weak DBSs were classified upon binding affinity ≥ 60 and < 60. A transcript whose promoter region contains a strong DBS of an HS lncRNA was assumed to be a target transcript of the HS lncRNA, and the gene containing this transcript was assumed to be a target gene of the HS lncRNA. As the 1000 Genomes Project (phase I) data and the archaic human genomes are based on GRCh37/hg19, the coordinates of DBSs were converted from GRCh38/hg38 to GRCh37/hg19 using the liftover program from the UCSC Genome Browser (Kuhn et al., 2013).

Experimentally validating DBS prediction

A 157 bp sequence (chr17:80252565-80252721, hg19) containing the DBD of RP13-516M14.1, a 202 bp sequence (chr1:113392603-113392804, hg19) containing the DBD of RP11-426L16.8, and a 198 bp sequence (chr17:19460524-19460721, hg19) containing the DBD of SNORA59B, were knocked out in the HeLa cell line, RKO cell line, and SK-MES-1 cell line, respectively. Two sequences (chr1:156643524-156643684, chr10:52445649-52445740, hg38) containing the DBD of two wrongly transcribed noncoding sequences were knocked out in the HCT-116 and A549 cell lines, respectively. The seven cases of knocked out were performed by UBIGENE, Guangzhou, China, using CRISPR-UTM, a revised version of CRISPR/Cas9 technology (http://www.ubigene.com). The RNA-sequencing (RNA-seq) was performed by Novogene, Beijing, China (https://cn.novogene.com) and HaploX, Shenzhen, China (https://www.haplox.cn/), respectively.

Mapping DBSs of HS lncRNAs in the chimpanzee and archaic human genomes

We used the liftover program from the UCSC Genome Browser (http://genome.UCSC.edu) to map loci of DBSs in the human genome hg38 to the chimpanzee genome Pan_tro 3.0 (panTro5). The mapping results were examined by checking the human-chimpanzee pairwise alignment (hg38 vs panTro5) in the UCSC Genome Browser (http://genome.UCSC.edu). 97.81% (100403/102651) of DBSs were mapped, and the 2248 unmapped DBSs in 429 genes were considered human-specific. We used vcf-consensus in the VCFtools package to extract the DBSs of HS lncRNAs from the VCF files of Altai Neanderthals, Denisovans, and Vindija Neanderthals. The variant with the highest quality score was selected whenever multiple variant calls were observed at a given locus. The obtained DBS sequences in the chimpanzees and three archaic humans are called counterparts of DBSs in these genomes.

Estimating sequence distances of DBSs between different genomes

We first aligned DBS sequences in the genomes of humans, chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals using the MAFFT7 program for measuring sequence distances from modern humans to chimpanzees and archaic humans (Katoh and Standley, 2013). We then computed sequence distances using the dnadist program with the Kimura 2-parameter model in the PHYLIP (3.6) package (http://evolution.genetics.washington.edu/phylip.html) and using the Tamura-Nei model in the MEGA7 package (Kumar et al., 2016). The two methods generated equivalent results. The largest distance between DBSs in humans and their counterparts in chimpanzees is 5.383. Since 2248 DBSs in 429 human genes do not have counterparts in chimpanzees, we assumed that the sequence distances of these DBSs between humans and chimpanzees were 10.0. Since the top 20% of DBSs in chimpanzees (including these 2248 DBSs) have distances ≥ 0.034, we used 0.034 as the threshold to identify the “mostly changed” DBSs in the chimpanzees and three archaic humans.

We determined human ancestral sequences of DBSs upon the human ancestor sequences from the EBI website, which were generated upon the EPO alignments of six primate species. We used the above-mentioned methods to compute DBSs’ sequence distances from the human ancestor to chimpanzees, archaic humans, and modern humans. We found that when the human-chimpanzee ancestral sequence has the ancestral sequence (which means the inferring of ancestral allele is of high confidence), DBS distances from the human ancestor to modern humans are larger than to archaic humans, but this situation only accounts for about 63.8%. For many DBSs, the distances from the human ancestor to modern humans are smaller than to archaic humans (especially Neanderthals and Denisovans) and even to chimpanzees. This defect may be caused by the absence of archaic humans in building the human ancestral sequence.

Detecting positive selection signals in HS lncRNA genes and DBSs

We used multiple tests to detect positive selection signals in HS lncRNA genes. First, we used the XP-CLR test (parameters = -w1 0.001 300 100 -p0 0.95, window size = 0.1 cM, grid size = 100 bp) to perform six pairwise genome-wide scans (i.e., CEU-CHB, CEU-YRI, CHB-CEU, CHB-YRI, YRI-CEU, and YRI-CHB) (Chen et al., 2010). The upper 1% of scores across the entire genome in each pairwise scan were 34.6 in the CEU-YRI scan, 16.8 in the CEU-CHB scan, 45.0 in the CHB-YRI scan, 26.9 in the CHB-CEU scan, 14.1 in the YRI-CEU scan, and 14.1 in the YRI-CHB scan. These scores were used as the thresholds of positive selection signals in these populations. Second, we used the iSAFE program to scan each genomic region that contains a HS lncRNA gene and its 500-kb upstream and downstream sequences (Akbari et al., 2018). Strongly selected loci were detected only in CEU and CHB. Third, we used the VCFtools program to calculate Tajima’s D values for each HS lncRNA gene in CEU, CHB, and YRI (Danecek et al., 2011). The calculation was performed using a 1500-bp non-overlapping sliding window because the lengths of these genes exceed 1500 bp. To generate the background reference for judging the significant increase or decrease of Tajima’s D for HS lncRNA genes in a population, we calculated Tajima’s D values across the whole genome using a sliding window of 1500 bp. As values of Tajima’s D were compared with the background reference, a significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively, instead of population demography dynamics (Tajima, 1989). Fourth, we used the integrated Fst to detect positive selection signals in HS lncRNA genes. The Fst of each HS lncRNA gene was computed for three comparisons, i.e., CEU-YRI, CHB-YRI, and CHB-CEU. Extreme Fst values of SNPs were detected in HS lncRNA genes in the comparisons of CEU-YRI and CHB-YRI. Since allele frequencies for different loci vary in a genome and genetic drift may have different effects at different loci, we used a sliding window of 1500 for comparing Fst values of HS lncRNA genes with the genome-wide background. Extreme Fst values indicate positive selection. Finally, we applied linkage disequilibrium (LD) analysis to each HS lncRNA gene. We computed the pairwise LD (r²) in CEU, CHB, and YRI for common SNPs (with minimal allele frequency (MAF) ≥ 0.05 in at least one population) in HS lncRNA genes and DBSs using the PLINK program (Purcell et al., 2007). Significantly increased LD was detected in SNPs in HS lncRNA genes in CEU and CHB. The LD patterns were portrayed using the Haploview program (Barrett et al., 2005).

Next, we used the above tests to detect positive selection signals in DBSs. First, the 100 bp grid size of the XP-CLR test also allowed the detection of selection signals in DBSs. Second, we performed the Tajima’s D and Fay-Wu’s H tests (Fay and Wu, 2000; Tajima, 1989). We calculated Tajima’s D values for each DBS in CEU, CHB, and YRI using the VCFtools program (Danecek et al., 2011), with a sliding window of 147 bp (the mean length of strong DBSs). To generate the background reference for judging the significant increase or decrease of Tajima’s D in a population, we calculated Tajima’s D values across the whole genome using the same sliding window. As values of Tajima’s D were compared with the background reference, a significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively. Fay-Wu’s H values were calculated similarly using the VariScan program (Vilella et al., 2005). Calculating Fay-Wu’s H demands the ancestral sequences as the outgroup. We extracted the ancestral sequences of DBSs from the human ancestral sequence which was generated upon the EPO alignments of six primate species (human, chimpanzee, gorilla, orangutan, macaque, and marmoset) (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignm ents/). Third, we computed the Fst to measure the frequency differences of alleles in DBSs between populations. For the CEU-YRI, CHB-YRI, and CHB-CEU pairwise comparisons, we used the revised VCFtools program to compute the weighted Fst for all SNPs in each DBS (Weir and Cockerham, 1984). Fourth, we integrated the weighted Fst values in the three populations into an “integrated Fst” which indicated whether the DBS locus was under selection in a certain population (Nielsen and Slatkin, 2013). We used sliding windows of 147 bp for comparing Fst values of DBSs with the genome-wide background. We empirically denoted the values of integrated Fst in the upper 10% of scores across the entire genome as statistically significant. To determine positive selection more reliably, we used Tajima’s D and the integrated Fst to jointly determine if a DBS was under positive selection in a population. The thresholds which determined the upper 10% of Tajima’s D values across the entire genome in CEU, CHB, and YRI were -0.97, -0.96, and -0.97, respectively, and the threshold which determined the upper 10% of integrated Fst values across the entire genome in the three populations was 0.22. For example, a DBS was assumed to be under positive selection in CEU if (a) the DBS had a Tajima’s D<-0.97 in CEU and Tajima’s D>0.0 in the two other populations and (b) the DBS had an integrated Fst>0.22. Analogously, a DBS was assumed to be under positive selection in both CEU and CHB if the DBS had a Tajima’s D <-0.97 in CEU, <-0.96 in CHB, and >0.0 in YRI, and had an integrated Fst>0.22.

Functional enrichment analysis of genes

We used the g:Profiler program (with the parameter settings: Organism=Homo sapiens, Ordered query = No, Significance threshold = Benjamini-Hochberg FDR, User threshold = 0.05, 50 < terms size < 1000) and the Gene Ontology (GO) database to perform over-representation analysis (Raudvere et al., 2019). Over-representation analysis is used to determine which a priori defined gene sets are more present (over-represented) in a subset of “interesting” genes than what would be expected by chance. Here, we used the analysis to examine different target gene sets are enriched in what GO terms (Biological Process). The target gene sets included genes with strongest DBSs in humans, genes with weakest DBSs in humans, and genes with the mostly changed DBSs in chimpanzees and three archaic humans. To choose genes with strongest DBSs, we ranked genes in descending order based on DBS binding affinity and chose the top 2000 genes for over-representation analysis. To choose genes with weakest DBSs, we first ranked genes in ascending order based on DBS binding affinity, then ranked genes in descending order based on the number of regulatory HS lncRNAs, and finally chose the top 2000 genes with binding affinity ≤ 40 and regulatory HS lncRNAs ≥ 5. We used the distance threshold of 0.034 to choose genes with mostly changed DBSs. Genes with mostly changed DBSs in chimpanzees include the 429 genes in which 2248 DBSs lack counterparts. 4248, 1256, 2514, and 134 genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals have DBSs or their counterparts with distance ≥ 0.034. To make the statistical significance of enrichment analysis comparable across chimpanzees, Altai Neanderthals, and Denisovans, we used the same number of genes (the top 1256 genes whose DBSs have the largest DBS distances) as the input to the g:Profiler program.

Analysing SNP frequencies in human populations

The frequency of common SNPs (MAF ≥ 0.05) in DBSs in the three modern human populations was computed using the VCFtools package (Danecek et al., 2011). The ancestral/derived states of SNPs were inferred upon the human ancestor sequences and were used to determine derived allele frequency.

Analysing the cis-effect of SNPs in DBSs on target gene expression

SNPs with MAF > 0.1 in DBSs in any of the three modern human populations and absolute values of cis-effect size > 0.5 (FDR < 0.05) in any of the GTEx tissues were examined for an influence on the expression of the target genes (GTEx Consortium, 2017). SNPs that are eQTLs in the GTEx tissues and have biased derived allele frequencies (DAF) in the three modern human populations were examined to estimate whether the eQTL is population-specific.

Examining the tissue-specific impact of HS lncRNA-regulated gene expression

First, we examined the expression of HS lncRNA genes across the GTEx tissues. HS lncRNA genes with a median TPM value > 0.1 in a tissue were considered robustly expressed in that tissue. Upon this criterion, 40 HS lncRNA genes were expressed in at least one tissue and were used to examine the impact of HS lncRNA regulation on gene and transcript expression (other HS lncRNAs may function in the cytoplasm). Since a HS lncRNA gene may have multiple transcripts, we adopted the transcript which contains the predicted DNA binding domain (DBD) and has the largest TPM value as the representative transcript of the HS lncRNA. We calculated the pairwise Spearman’s correlation coefficient between the expression of a HS lncRNA (the representative transcript) and the expression of each of its target transcripts using the scipy.stats.spearmanr program in the scipy package. The expression of a HS lncRNA gene and a target transcript was considered to be significantly correlated if the |Spearman’s rho| > 0.3, with Benjamini-Hochberg FDR < 0.05. We examined the percentage distribution of correlated HS lncRNA-target transcript pairs across GTEx tissues and organs (Figure 3A).

We examined all GTEx tissues to determine in which tissues HS lncRNA-regulated gene expression may have undergone significant changes from archaic to modern humans. If a pair of HS lncRNA and target transcript are robustly expressed in a tissue and their expression shows a significant correlation (|Spearman’s rho| > 0.3, with Benjamini-Hochberg FDR < 0.05) in the tissue, we computed the sequence distance of the HS lncRNA’s DBS in the transcript from the three archaic humans to modern humans. We compared the sequence distances of all DBSs in each tissue with the sequence distances of all DBSs in all GTEx tissues (i.e. the distribution of all DBS distances was taken as the background). A one-sided two-sample Kolmogorov-Smirnov test was used to examine whether the sequence distances of all DBSs in a specific tissue deviate from the background distribution (which reflects the “neutral evolution” of gene expression). For a tissue, if the test result was Benjamini-Hochberg FDR < 0.001, the tissue was considered to have significantly changed gene expression regulated by the HS lncRNA. We used different colors to mark tissues with significantly changed gene expression regulation since Altai Neanderthals, Denisovans, and Vindija Neanderthals, and used “D”, “AD”, and “ADV” to indicate changes since Denisovans, since Altai Neanderthals and Denisovans, and since Altai Neanderthals, Denisovans, and Vindija Neanderthals, respectively (Figure 3B).

Declaration of competing interest

The authors declare no competing interest.

Author contributions

H.Z. designed the study and drafted the manuscript. J.L. performed most data analyses. Y.W. and J.T. performed XP-CLR and iSAFE analyses. All authors have read the manuscript and consent to its publication.

Supporting information

eLife Assessment

In this valuable manuscript, the authors attempt to examine the role of long non-coding RNAs (lncRNAs) in human evolution, through a set of population genetics and functional genomics analyses that leverage existing datasets and tools. Although the methods are at times inadequate - for example, suitable methods and/or relevant controls are lacking at many points, and selection is inferred sometimes too quickly - the results nonetheless point towards a possible contribution of long non-coding RNAs to the evolution of human biology and they suggest clear directions for future, more rigorous study.

Reviewer #1 (Public Review):

Summary

While DNA sequence divergence, differential expression, and differential methylation analysis have been conducted between humans and the great apes to study changes that "make us human", the role of lncRNAs and their impact on the human genome and biology has not been fully explored. In this study, the authors computationally predict HSlncRNAs as well as their DNA Binding sites using a method they have developed previously and then examine these predicted regions with different types of enrichment analyses. Broadly, the analysis is straightforward and after identifying these regions/HSlncRNAs the authors examined their effects using different external datasets.

Strengths/weaknesses

By and large, the analysis performed is dependent on their ability to identify HSlncRNAs and their DBS. I think that they have done a good job of showing the performance metrics of their methods in previous publications. Thereafter, they perform a series of enrichment-type analyses that have been used in the field for quite a while now to look at tissue-specific enrichment, or region-specific enrichment, or functional enrichment, and I think these have been carried out well. The authors achieved the aims of their work. I think one of the biggest contributions that this paper brings to the field is their annotation of these HSlncRNAs. Thus a major revisionary effort could be spent on applying their method to the latest genomes that have been released so that the community could get a clean annotation of newly identified HSlncRNAs (see comment 2).

Comments

1. Though some of their results about certain HSlncRNAs having DBSs in all genes is rather surprising/suspicious, I think that broadly their process to identify and validate DBSs is robust, they have multiple lines of checks to identify such regions, including functional validation. These predictions are bound to have some level of false positive/negative rate and it might be nice to restate those here and on what experiment/validation data these were conducted. However, the rest of their analysis comprises different types of enrichment analysis which shouldn't be affected by outlier HSlncRNAs if indeed their FPR/FNR are low.

2. There are now several new genomes available as part of the Zoonomia consortium and 240 Primate consortium papers released. These papers have re-examined some annotations such as Human Accelerated Regions (HARs) and found with a larger dataset as well as better reference genomes, that a large fraction of HARs were actually incorrectly annotated - that is that they were also seen in other lineages outside of just the great apes. If these papers have not already examined HSlncRNAs, the authors should try and re-run the computational predictions with this updated set and then identify HSlncRNAs there. This might help to clarify their signal and remove lncRNAs that might be present in other primates but are somehow missing in the great apes. This might also help to mitigate some results that they see in section 3 of their paper in comparing DBS distances between archaics and humans.

3. The differences between the archaic hominins in their DBS distances to modern humans are a bit concerning. At some level, we expect these to be roughly similar when examining African modern humans and perhaps the Denisovan being larger when examining Europeans and Asians, but they seem to have distances that aren't expected given the demography. In addition, from their text for section 3, they begin by stating that they are computing two types of distances but then I lost track of which distance they were discussing in paragraph 3 of section 3. Explicitly stating which of the two distances in the text would be helpful for the reader.

4. Isn't the correct control to examine whether eQTLs are more enriched in HSlncRNA DBSs a set of transcription factor binding sites? I don't think using just promoter regions is a reasonable control here. This does not take away from the broader point however that eQTLs are found in DBSs and I think they can perform this alternate test.

5. In the discussion, they highlight the evolution of sugar intake, which I'm not sure is appropriate. This comes not from GO enrichment but rather from a few genes that are found at the tail of their distribution. While these signals may be real, the evolution of traits is often highly polygenic and they don't see this signal in their functional enrichment. I suggest removing that line. Moreover, HSlncRNAs are ones that are unique across a much longer time frame than the transition to agriculture which is when sugar intake rose greatly. Thus, it's unlikely to see enrichment for something that arose in the past 6000-7000 years would in the annotation that is designed to detect human-chimp or human-neanderthal level divergence.

Reviewer #2 (Public Review):

Lin et al attempt to examine the role of lncRNAs in human evolution in this manuscript. They apply a suite of population genetics and functional genomics analyses that leverage existing data sets and public tools, some of which were previously built by the authors, who clearly have experience with lncRNA binding prediction. However, I worry that there is a lack of suitable methods and/or relevant controls at many points and that the interpretation is too quick to infer selection. While I don't doubt that lnc RNAs contribute to the evolution of modern humans, and certainly agree that this is a question worth asking, I think this paper would benefit from a more rigorous approach to tackling it.

At this point, my suggestions are mostly focused on tightening and strengthening the methods; it is hard for me to predict the consequence of these changes on the results or their interpretation, but as a general rule I also encourage the authors to not over-interpret their conclusions in terms of what phenotype was selected for when as they do at certain points (eg glucose metabolism).

I note some specific points that I think would benefit from more rigorous approaches, and suggest possible ways forward for these.

1. Much of this work is focused on comparing DNA binding domains in human-unique long-noncoding RNAs and DNA binding sites across the promoters of genes in the human genome, and I think the authors can afford to be a bit more methodical/selective in their processing and filtering steps here. The article begins by searching for orthologues of human lncRNAs to arrive at a set of 66 human-specific lncRNAs, which are then characterised further through the rest of the manuscript. Line 99 describes a binding affinity metric used to separate strong DBS from weak DBS; the methods (line 432) describe this as being the product of the DBS or lncRNA length times the average Identity of the underlying TTSs. This multiplication, in fact, undoes the standardising value of averaging and introduces a clear relationship between the length of a region being tested and its overall score, which in turn is likely to bias all downstream inference, since a long lncRNA with poor average affinity can end up with a higher score than a short one with higher average affinity, and it's not quite clear to me what the biological interpretation of that should be. Why was this metric defined in this way?

2. There is also a strong assumption that identified sites will always be bound (line 100), which I disagree is well-supported by additional evidence (lines 109-125). The authors show that predicted NEAT1 and MALAT1 DBS overlap experimentally validated sites for NEAT1, MALAT1, and MEG3, but this is not done systematically, or genome-wide, so it's hard to know if the examples shown are representative, or a best-case scenario.

It's also not quite clear how overlapping promoters or TSS are treated - are these collapsed into a single instance when calculating genome-wide significance? If, eg, a gene has five isoforms, and these differ in the 3' UTR but their promoter region contains a DBS, is this counted five times, or one? Since the interaction between the lncRNA and the DBS happens at the DNA level, it seems like not correcting for this uneven distribution of transcripts is likely to skew results, especially when testing against genome-wide distributions, eg in the results presented in sections 5 and 6. I do not think that comparing genes and transcripts putatively bound by the 40 HS lncRNAs to a random draw of 10,000 lncRNA/gene pairs drawn from the remaining ~13500 lncRNAs that are not HS is a fair comparison. Rather, it would be better to do many draws of 40 non-HS lncRNAs and determine an empirical null distribution that way, if possible actively controlling for the overall number of transcripts (also see the following point).

3. Thresholds for statistical testing are not consistent, or always well justified. For instance, in line 142 GO testing is performed on the top 2000 genes (according to different rankings), but there's no description of the background regions used as controls anywhere, or of why 2000 genes were chosen as a good number to test? Why not 1000, or 500? Are the results overall robust to these (and other) thresholds? Then line 190 the threshold for downstream testing is now the top 20% of genes, etc. I am not opposed to different thresholds in principle, but they should be justified.

Likewise, comparing Tajima's D values near promoters to genome-wide values is unfair, because promoters are known to be under strong evolutionary constraints relative to background regions; as such it is not surprising that the results of this comparison are significant. A fairer comparison would attempt to better match controls (eg to promoters without HS lncRNA DBS, which I realise may be nearly impossible), or generate empirical p-values via permutation or simulation.

4. There are huge differences in the comparisons between the Vindija and Altai Neanderthal genomes that to me suggest some sort of technical bias or the such is at play here. e.g. line 190 reports 1256 genes to have a high distance between the Altai Neanderthal and modern humans, but only 134 Vindija genes reach the same cutoff of 0.034. The temporal separation between the two specimens does not seem sufficient to explain this difference, nor the difference between the Altai Denisovan and Neanderthal results (2514 genes for Denisovan), which makes me wonder if it is a technical artefact relating to the quality of the genome builds? It would be worth checking.

5. Inferring evolution: There are some points of the manuscript where the authors are quick to infer positive selection. I would caution that GTEx contains a lot of different brain tissues, thus finding a brain eQTL is a lot easier than finding a liver eQTL, just because there are more opportunities for it. Likewise, claims in the text and in Tables 1 and 2 about the evolutionary pressures underlying specific genes should be more carefully stated. The same is true when the authors observe high Fst between groups (line 515), which is only one possible cause of high Fst - population differentiation and drift are just as capable of giving rise to it, especially at small sample sizes.

Author Response

The following is the authors’ response to the original reviews.

We thank the two reviewers very much for their careful review and valuable comments. Upon these comments, the following revisions have been made. First, we have performed a new analysis on human accelerated regions (HARs) recently reported by the Zoonomia Project. Second, we have presented more data on experimentally detected and computationally predicted DBSs of MALAT1, NEAT1, and MEG3. Third, we have added details on the RNA-seq data processing and subsequent differential expression testing to the Materials and Methods section. Fourth, we have clarified some details on the human ancestor sequence and the use of parameters and thresholds. Six new citations are added. In addition, we have also carefully polished the main text. We hope these revisions, together with the Responses-to-Reviewers, would help the reader better get the information from the paper.

eLife assessment

In this valuable manuscript, the authors attempt to examine the role of long non-coding RNAs (lncRNAs) in human evolution, through a set of population genetics and functional genomics analyses that leverage existing datasets and tools. Although the methods are at times inadequate - for example, suitable methods and/or relevant controls are lacking at many points, and selection is inferred sometimes too quickly - the results nonetheless point towards a possible contribution of long non-coding RNAs to the evolution of human biology and they suggest clear directions for future, more rigorous study.

Public Reviews:

Reviewer #1 (Public Review):

Summary

While DNA sequence divergence, differential expression, and differential methylation analysis have been conducted between humans and the great apes to study changes that "make us human", the role of lncRNAs and their impact on the human genome and biology has not been fully explored. In this study, the authors computationally predict HSlncRNAs as well as their DNA Binding sites using a method they have developed previously and then examine these predicted regions with different types of enrichment analyses. Broadly, the analysis is straightforward and after identifying these regions/HSlncRNAs the authors examined their effects using different external datasets.

Strengths/weaknesses

By and large, the analysis performed is dependent on their ability to identify HSlncRNAs and their DBS. I think that they have done a good job of showing the performance metrics of their methods in previous publications. Thereafter, they perform a series of enrichment-type analyses that have been used in the field for quite a while now to look at tissue-specific enrichment, or region-specific enrichment, or functional enrichment, and I think these have been carried out well. The authors achieved the aims of their work. I think one of the biggest contributions that this paper brings to the field is their annotation of these HSlncRNAs. Thus a major revisionary effort could be spent on applying their method to the latest genomes that have been released so that the community could get a clean annotation of newly identified HSlncRNAs (see comment 2).

Comments

1. Though some of their results about certain HSlncRNAs having DBSs in all genes is rather surprising/suspicious, I think that broadly their process to identify and validate DBSs is robust, they have multiple lines of checks to identify such regions, including functional validation. These predictions are bound to have some level of false positive/negative rate and it might be nice to restate those here and on what experiment/validation data these were conducted. However, the rest of their analysis comprises different types of enrichment analysis which shouldn't be affected by outlier HSlncRNAs if indeed their FPR/FNR are low.

1. There are now several new genomes available as part of the Zoonomia consortium and 240 Primate consortium papers released. These papers have re-examined some annotations such as Human Accelerated Regions (HARs) and found with a larger dataset as well as better reference genomes, that a large fraction of HARs were actually incorrectly annotated - that is that they were also seen in other lineages outside of just the great apes. If these papers have not already examined HSlncRNAs, the authors should try and re-run the computational predictions with this updated set and then identify HSlncRNAs there. This might help to clarify their signal and remove lncRNAs that might be present in other primates but are somehow missing in the great apes. This might also help to mitigate some results that they see in section 3 of their paper in comparing DBS distances between archaics and humans.

Responses:

(1) Thanks for the good suggestion. We have checked the Zoonomia reported genomes and found that new primate genomes are monkeys and lemurs but not apes (Zoonomia Consortium. Nature 2023. https://doi.org/10.1038/s41586-020-2876-6), and the phylogenetic relationships between monkeys and humans are much more remote than those between apes and humans. In addition, the Zoonomia project did target identifying new lncRNA genes.

(2) We have examined the Zoonomia-reported HARs (Keough et al. Science 2023. DOI: 10.1126/science.abm1696). Of the 312 HARs reported by Keough et al, 8 overlap 26 DBSs of 14 HS lncRNAs; moreover, DBSs greatly outnumber HARs, suggesting that HAR and DBS are different sequences with different functions.

(3) In the revised manuscript, a new paragraph (the second one) has been added to the section “HS lncRNAs regulate diverse genes and transcripts” to describe the HAR analysis result.

1. The differences between the archaic hominins in their DBS distances to modern humans are a bit concerning. At some level, we expect these to be roughly similar when examining African modern humans and perhaps the Denisovan being larger when examining Europeans and Asians, but they seem to have distances that aren't expected given the demography. In addition, from their text for section 3, they begin by stating that they are computing two types of distances but then I lost track of which distance they were discussing in paragraph 3 of section 3. Explicitly stating which of the two distances in the text would be helpful for the reader.

Responses:

(1) Upon the archaic human genomes, the genomic distances from the three modern humans are shorter to Denisovan than to Altai Neanderthal; however, upon the related studies we cite, the phylogenetic relationship between the three modern humans is more remote to Denisovan than to Altai Neanderthal. Thus, the finding that 2514 and 1256 DBSs have distances >0.034 in Denisovans and Altai Neanderthals is not unreasonable. The numbers of DBSs, of course, depend on the cutoff of 0.034, which is somewhat subjective but not unreasonable.

(2) The second paragraph is added to the Discussion, discussing parameters and cutoffs.

(3) Regarding the two types of distance, the distances computed in the first way were not further analyzed because, as we note, “This anomaly may be caused by that the human ancestor was built using six primates without archaic humans”.

1. Isn't the correct control to examine whether eQTLs are more enriched in HSlncRNA DBSs a set of transcription factor binding sites? I don't think using just promoter regions is a reasonable control here. This does not take away from the broader point however that eQTLs are found in DBSs and I think they can perform this alternate test.

Responses:

Indeed, TFBSs are more comparable to DBSs than promoters. However, many more methods have been developed to predict TFBSs than to predict DBSs, making us concerned about TFBS prediction's reliability. Since most QTLs in DBSs are mQTLs (Supplementary Table 13), but many QTLs in TFBSs are eQTLs (Flynn et al. PLoS Genetics 2021. DOI: 10.1371/journal.pgen.1009719), it is pretty safe to conclude that DBSs are enriched in mQTLs.

1. In the Discussion, they highlight the evolution of sugar intake, which I'm not sure is appropriate. This comes not from GO enrichment but rather from a few genes that are found at the tail of their distribution. While these signals may be real, the evolution of traits is often highly polygenic and they don't see this signal in their functional enrichment. I suggest removing that line. Moreover, HSlncRNAs are ones that are unique across a much longer time frame than the transition to agriculture which is when sugar intake rose greatly. Thus, it's unlikely to see enrichment for something that arose in the past 6000-7000 years would in the annotation that is designed to detect human-chimp or human-neanderthal level divergence.

Responses:

(1) The Discussion on human adaptation to high sugar intake is based on both enriched GO terms (Supplementary Table 4, 7) and a set of genes in modern humans with the most SNP-rich DBSs (Table 2). These glucose-related GO terms are not at the tail of the list because, of the 614 enriched GO terms (enriched in genes with strongest DBSs), glucose metabolism-related ones are ranked 208, 212, 246, 264, 504, 522, 591, and of the 409 enriched GO terms (enriched in the top 1256 genes in Altai Neanderthals), glucose metabolism-related ones are ranked 152 and 217.

(2) Indeed, there are other top-ranked enriched GO terms; some (e.g., neuron projection development (GO:0031175) and cell projection morphogenesis (GO:0048858)) have known impact on human evolution, but the impact of others (e.g., cell junction organization (GO:0034330)) remain unclear. We specifically report human adaptation to high sugar intake because the DBSs in related genes show differences in modern humans (Table 2).

Reviewer #2 (Public Review):

Lin et al attempt to examine the role of lncRNAs in human evolution in this manuscript. They apply a suite of population genetics and functional genomics analyses that leverage existing data sets and public tools, some of which were previously built by the authors, who clearly have experience with lncRNA binding prediction. However, I worry that there is a lack of suitable methods and/or relevant controls at many points and that the interpretation is too quick to infer selection. While I don't doubt that lnc RNAs contribute to the evolution of modern humans, and certainly agree that this is a question worth asking, I think this paper would benefit from a more rigorous approach to tackling it.

At this point, my suggestions are mostly focused on tightening and strengthening the methods; it is hard for me to predict the consequence of these changes on the results or their interpretation, but as a general rule I also encourage the authors to not over-interpret their conclusions in terms of what phenotype was selected for when as they do at certain points (eg glucose metabolism).

Responses:

(1) Now, we use more cautious wording to describe the results.

(2) A paragraph (the second one) is added to Discussion to explain parameters and cutoffs.

(3) We make the caution at the end of the third paragraph that “We note that these are findings instead of conclusions, and they indicate, suggest, or support something revealing the primary question of what genomic differences critically determine the phenotypic differences between humans and apes and between modern and archaic humans”.

I note some specific points that I think would benefit from more rigorous approaches, and suggest possible ways forward for these.

1. Much of this work is focused on comparing DNA binding domains in human-unique long-noncoding RNAs and DNA binding sites across the promoters of genes in the human genome, and I think the authors can afford to be a bit more methodical/selective in their processing and filtering steps here. The article begins by searching for orthologues of human lncRNAs to arrive at a set of 66 human-specific lncRNAs, which are then characterised further through the rest of the manuscript. Line 99 describes a binding affinity metric used to separate strong DBS from weak DBS; the methods (line 432) describe this as being the product of the DBS or lncRNA length times the average Identity of the underlying TTSs. This multiplication, in fact, undoes the standardising value of averaging and introduces a clear relationship between the length of a region being tested and its overall score, which in turn is likely to bias all downstream inference, since a long lncRNA with poor average affinity can end up with a higher score than a short one with higher average affinity, and it's not quite clear to me what the biological interpretation of that should be. Why was this metric defined in this way?

Responses:

(1) Binding affinity and length of all DBSs of HS lncRNAs are given in Supplementary Table 2 and 3. Since a triplex (say, 100 bp in length) may have 50% or 70% of nucleotides bound, it is necessary to differentiate binding affinity and length, and the two measures can differentiate DBSs of the same length but with different binding affinity and DBSs with the same binding affinity but different length.

(2) Differentiating DBSs into strong and weak ones is somewhat subjective, accurately differentiating them demands experimental data that are currently unavailable, and it is advisable to separately analyze strong and weak DBSs because they may likely influence different aspects of human evolution.

1. There is also a strong assumption that identified sites will always be bound (line 100), which I disagree is well-supported by additional evidence (lines 109-125). The authors show that predicted NEAT1 and MALAT1 DBS overlap experimentally validated sites for NEAT1, MALAT1, and MEG3, but this is not done systematically, or genome-wide, so it's hard to know if the examples shown are representative, or a best-case scenario.

Responses:

(1) We do not assume/think that identified sites will always be bound. Instead, lncRNA/DBS binding is highly context-dependent (including tissue-specific).

(2) An extra supplementary table (Supplementary Table 15) is added to show what predicted DBSs overlap experimentally detected DBSs for NEAT1, MALAT1, and MEG3. By the way, it is more accurate to say “experimentally detected” than “experimentally validated”, because experimental data have true/false positives and true/false negatives, and different sequencing protocols (for detecting lncRNA/DNA binding) may generate somewhat different results.

It's also not quite clear how overlapping promoters or TSS are treated - are these collapsed into a single instance when calculating genome-wide significance? If, eg, a gene has five isoforms, and these differ in the 3' UTR but their promoter region contains a DBS, is this counted five times, or one? Since the interaction between the lncRNA and the DBS happens at the DNA level, it seems like not correcting for this uneven distribution of transcripts is likely to skew results, especially when testing against genome-wide distributions, eg in the results presented in sections 5 and 6. I do not think that comparing genes and transcripts putatively bound by the 40 HS lncRNAs to a random draw of 10,000 lncRNA/gene pairs drawn from the remaining ~13500 lncRNAs that are not HS is a fair comparison. Rather, it would be better to do many draws of 40 non-HS lncRNAs and determine an empirical null distribution that way, if possible actively controlling for the overall number of transcripts (also see the following point).

Responses:

(1) We analyzed each and every GENCODE-annotated transcript (Supplementary Table 2). For example, if a gene has N TSS and N transcripts, DBSs are predicted in N promoter regions. When analyzing gene expression in tissues, each and every transcript is analyzed.

(2) Ideally, it would be better to do many draws, but statistically, a huge number is needed due to the number of total genes in the human genome.

(3) We feel that doing many draws of 40 non-HS lncRNAs and determining an empirical null distribution is not as straightforward as comparing HS lncRNA-target transcript pairs (45% show significant expression correlation) with random lncRNA-random transcript pairs (2.3% show significant expression correlation).

1. Thresholds for statistical testing are not consistent, or always well justified. For instance, in line 142 GO testing is performed on the top 2000 genes (according to different rankings), but there's no description of the background regions used as controls anywhere, or of why 2000 genes were chosen as a good number to test? Why not 1000, or 500? Are the results overall robust to these (and other) thresholds? Then line 190 the threshold for downstream testing is now the top 20% of genes, etc. I am not opposed to different thresholds in principle, but they should be justified.

Responses:

(1) The over-representation analysis using g:Profiler was applied to the top and bottom 2000 genes with the whole genome as the background. The number “2000” was chosen somewhat subjectively. If more or fewer genes were chosen, more or fewer enriched GO terms would be identified, but GO terms with adjusted P-values <0.05 would be quite stable.

(2) A paragraph (the second one) is added to the Discussion to explain parameters and cutoffs.

Likewise, comparing Tajima's D values near promoters to genome-wide values is unfair, because promoters are known to be under strong evolutionary constraints relative to background regions; as such it is not surprising that the results of this comparison are significant. A fairer comparison would attempt to better match controls (eg to promoters without HS lncRNA DBS, which I realise may be nearly impossible), or generate empirical p-values via permutation or simulation.

Responses:

We examined Tajima’s D in DBSs (Supplementary Figure 9) and in HS lncRNA genes (Supplementary Figure 18). We compared the Tajima’s D values with the genome-wide background in both cases.

1. There are huge differences in the comparisons between the Vindija and Altai Neanderthal genomes that to me suggest some sort of technical bias or the such is at play here. e.g. line 190 reports 1256 genes to have a high distance between the Altai Neanderthal and modern humans, but only 134 Vindija genes reach the same cutoff of 0.034. The temporal separation between the two specimens does not seem sufficient to explain this difference, nor the difference between the Altai Denisovan and Neanderthal results (2514 genes for Denisovan), which makes me wonder if it is a technical artefact relating to the quality of the genome builds? It would be worth checking.

Responses:

(1) The cutoff of 0.034 was chosen upon that DBSs in the top 20% (4248) genes in chimpanzees have distances larger than this cutoff, and accordingly, 4248, 1256, 2514, and 134 genes have DBSs distances >0.034 in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals. These numbers of genes qualitatively agree with the phylogenetic distances from chimpanzees, archaic humans to modern humans. If a percentage larger or smaller than 20% (e.g., 10% or 30%) is chosen, and so is a cutoff X, the numbers of genes with DBSs distance >X would not be 4248, 1256, 2514, and 134, but could still qualitatively agree with the phylogenetic distances from chimpanzees, archaic humans to modern humans.

(2) The second paragraph in the Discussion now explains the parameters and cutoffs.

1. Inferring evolution: There are some points of the manuscript where the authors are quick to infer positive selection. I would caution that GTEx contains a lot of different brain tissues, thus finding a brain eQTL is a lot easier than finding a liver eQTL, just because there are more opportunities for it. Likewise, claims in the text and in Tables 1 and 2 about the evolutionary pressures underlying specific genes should be more carefully stated. The same is true when the authors observe high Fst between groups (line 515), which is only one possible cause of high Fst - population differentiation and drift are just as capable of giving rise to it, especially at small sample sizes.

Responses:

(1) We analyzed brain tissues separately instead of taking the whole brain as a tissue, see Supplementary Table 12 and Figure 3.

(2) We make the caution at the end of the third paragraph that “We note that these are findings instead of conclusions, and they indicate, suggest, or support something revealing the primary question of what genomic differences critically determine the phenotypic differences between humans and apes and between modern and archaic humans”.

Reviewer #1 (Recommendations For The Authors):

Some figures are impossible to see/read so I wasn't able to evaluate them - Fig, 1B, 1E, 1F are small and blurry.

Responses:

High-quality figures are provided.

Typo in line 178: in these archaic humans, the distances of HS lncRNAs are smaller than the distances of DBSs.

Responses:

This is not a typo. We use “distance per base” to measure whether HS lncRNAs or their DBSs have evolved more from archaic humans to modern humans. See also Supplementary Note 4 and 5.

Reviewer #2 (Recommendations For The Authors):

1. There's some inconsistency in the genome builds and the database versions used, eg, sometimes panTro4 is used and sometimes panTro5 (line 456). Likewise, the version of GENCODE used is very old (18), the current version is 43. The current version contains 19928 lncRNAs, which is a big difference relative to what is being tested!

Responses:

(1) panTro4 was used to search orthologues of human lncRNAs; this time-consuming work started several years ago when the version of GENCODE was V18 (see Lin et al., 2019).

(2) Regarding “the version of GENCODE used is very old (V18)”, we have later examined the 4396 human lncRNAs reported in GENCODE V36 and found that the set of 66 HS lncRNAs remains the same.

(3) The counterparts of HS lncRNAs’ DBSs in chimpanzees were predicted recently using panTro5.

1. Table 1: What does 'mostly' mean in this context? I understand that it refers to sequence differences between humans and the other genomes, but what is the actual threshold, and how is it defined?

Responses:

The title of Table 1 is “Genes with strongest DBSs and mostly changed sequence distances from modern humans to archaic humans and chimpanzees”. Instead of using two cutoffs, choosing genes with the two features seems easy and sensible.

1. Line 117: The methods do not include information on the RNA-seq data processing and subsequent DE testing.

Responses:

The details are added to the section “Experimentally validating DBS predictiom” (The reads were aligned to the human GRCh38 genome using Hiasat2 (Kim et al., 2019), and the resulting sam files were converted to bam files using Samtools (Li et al., 2009). Stringtie was used to quantify gene expression level (Pertea et al., 2015). Fold change of gene expression was computed using the edgeR package (Robinson et al., 2010), and significant up- and down-regulation of target genes after DBD knockout was determined upon |log2(fold change)| > 1 with FDR < 0.1).

1. Line 180: I looked at the EPO alignment and it's not clear to me what 'human ancestor' means, but it may well explain the issues the authors have with calculating distances (I agree those numbers are weird). Is it the reconstructed ancestral state of humans at around 300-200,000 years ago (coalescence of most human uniparental lineages), or the inferred sequence of the human-chimpanzee most recent common ancestor? If it's the former, it's not surprising it skews results towards shorter distances for modern humans, since the tree distance from that point to archaic hominins is significantly larger than to modern humans.

Responses:

The “human ancestor” is constructed by the EBI team upon the genomes of six primates in the Ensembl website. We find that the reconstructed ancestral state of humans may be unlikely around 300,000-200,000 years, and may be much earlier. We also find that many DNA sequences of the “human ancestor” are low-confidence calls (i.e., the ancestral states are supported by only one primate’s sequence).

1. Line 221: SNP-rich DBS: Is this claim controlled for the length of the DBS?

Responses:

No. Long DBSs tend to have more SNPs. When comparing the same DBS in modern humans, archaic humans, and chimpanzees, both the length and SNP number reflect evolution, so it is not necessary to control for the length.

1. Given that GTEx is primarily built off short-read data and it is impossible to link binding of a lncRNA to a DBS with its impact with a specific transcript

Responses:

As written in the section “Examining the tissue-specific impact of HS lncRNA-regulated gene expression”, we calculated the pairwise Spearman's correlation coefficient between the expression of an HS lncRNA (the representative transcript, median TPM value > 0.1) and the expression of each of its target transcripts (median TPM value > 0.1) using the scipy.stats.spearmanr program in the scipy package. The expression of an HS lncRNA gene and a target transcript was considered to be significantly correlated if the |Spearman's rho| > 0.3, with Benjamini-Hochberg FDR < 0.05.

1. Line 429: should TTO be TFO?

Responses:

Here TTO should be TFO; the typo is corrected.

1. Methods, section 7: Some of the text in this section should perhaps be moved to the results section?

Responses:

Each of the two paragraphs in Methods’ section 7 is quite large, and some contents in Supplementary Notes are also very relevant. Thus, moving them to the Results section could make the Results too lengthy and specific.

1. Line 587: GTEx is built from samples of primarily European ancestry and has poor representation of African ancestry and negligible representation of Asian ancestry (see the GTEx v8 paper supplement). This means that it is basically impossible to find a non-European population-specific eQTL in GTEx, which in turn impacts these results.

Responses:

(1) Indeed, this is a serious issue of data analysis, and this issue cannot be solved until more Africans are sequenced.

(2) Anyway, one can still find considerable African-specific eQTLs in GTEx, such as rs28540058 (with frequency of 0, 0, 0.13 in CEU, CHB, YRI) and rs58772997 (with frequency of 0, 0, 0.12 in CEU, CHB, YRI (see Supplementary Table12 and Supplementary Figure 22).

Version 2

Abstract

Abstract

What genomic sequences make conserved genes generate divergent expression in closely related species, which may have critically driven human evolution, has puzzled researchers for decades. Genomic studies have examined species-specific gene birth, gene loss, and changes in promoters and transcription factor binding sites, but species-specific epigenetic regulation remains barely explored. This study identified human-specific long noncoding RNAs (lncRNAs) from GENCODE-annotated human lncRNAs, predicted their DNA binding sites (DBSs) genome-wide, analyzed these DBSs and their counterparts in modern humans (CEU, CHB, and YRI), archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and chimpanzees, and analyzed the impact of DBSs on gene expression in modern and archaic humans. The results suggest that human-specific lncRNAs and their DBSs have substantially rewired gene expression human-specifically and that the rewiring has evolved continuously from archaic to modern humans. Rewired gene expression promotes brain development, makes humans adapt to new environments and lifestyles, and causes differences in modern humans. These results uncover a critical dimension of human evolution and underscore the diverse functions of species-specific lncRNAs.

Introduction

Genomic changes that drive evolution include gene birth (Kaessmann, 2010), gene loss (Albalat and Canestro, 2016), and changes in regulatory sequences (Prud’homme et al., 2007). The limited genomic differences but substantial phenotypic and behavioral divergence between humans and other hominids make what genomic changes at what time have critically driven human evolution an enduring puzzle (Anton et al., 2014; Pollen et al., 2023). Although waves of studies have identified genes promoting human brain development (Evans et al., 2005; Fiddes et al., 2018; Mekel-Bobrov et al., 2005; Pinson et al., 2022; Suzuki et al., 2018), conclusions from gene birth-centered studies are controversial (Currat et al., 2006; Timpson et al., 2007; Yu et al., 2007). Moreover, genes important for brain development (e.g., NOTCH2NL and ASPM) do not necessarily regulate other human traits that precede or coincide with brain enlargement. Recent studies revealed the importance of regulatory sequences for human evolution, including new sequences (Liu et al., 2021), lost sequences (McLean et al., 2011), human accelerated regions (HARs) (Dong et al., 2016; Mangan et al., 2022; Prabhakar et al., 2006; Whalen and Pollard, 2022), and the turnover of transcription factor binding sites (TFBS) (Krieger et al., 2022; Otto et al., 2009; Zhang et al., 2023). An unexplored question is how epigenetic regulation contributes to human evolution.

Experimental studies have generated substantial findings on epigenetic regulation. (a) The chimpanzee and human genomes are almost 99% identical (Chimpanzee Sequencing and Analysis Consortium, 2005). (b) About one-third of human long-noncoding RNAs (lncRNA) are primate-specific (Derrien et al., 2012). (c) Species-specific lncRNAs have distinct expression profiles in organs of different species (Sarropoulos et al., 2019). (d) Many lncRNAs can bind to DNA sequences by forming RNA:DNA triplexes (Abu Almakarem et al., 2012), thereby recruiting histone and DNA modification enzymes to binding sites to regulate nearby gene expression (Lee, 2009; Roadmap Epigenomics et al., 2015). (e) Approximately 40% of differentially expressed human genes result from interspecies epigenetic differences (Hernando-Herraez et al., 2015). Compared with protein-coding gene-centered hypotheses of human evolution, the alternative one is plausible – human-specific (HS) lncRNAs and their DNA binding sites have human-specifically rewired gene expression to critically promote human evolution and determine human traits.

RNA sequence search by structure and sequence alignment can identify orthologous lncRNA genes in genomes (Nawrocki, 2014; Nawrocki and Eddy, 2013). Thus, HS lncRNAs can be identified upon human lncRNAs’ orthologues in other mammals. Many lncRNA can bind to DNA sequences following specific base-pairing rules to form RNA:DNA triplexes (Abu Almakarem et al., 2012). Thus, these rules allow computational prediction of lncRNAs’ DNA binding domains (DBD) and DNA binding sites (DBS) (Lin et al., 2019; Wen et al., 2022). Gene expression in most human organs and tissues has been profiled (GTEx Consortium, 2017). Further, cenomes of modern humans, archaic humans, and multiple apes have been sequenced (1000 Genomes Project Consortium et al., 2012; Chimpanzee Sequencing and Analysis Consortium, 2005; Meyer et al., 2012; Prufer et al., 2017; Pru fer et al., 2013). Together, these results make it feasible to explore the above alternative hypothesis.

In this study, we identified HS lncRNAs upon the orthologues of the GENCODE-annotated human lncRNA genes in 16 mammalian genomes, predicted HS lncRNAs’ DBDs and DBSs in the genomes of modern humans (CEU, CHB, and YRI), predicted these DBSs’ counterparts in the genomes of archaic humans (Altai Neanderthals, Denisovans, Vindija Neanderthals) and chimpanzees (Figure 1A), and experimentally validated DBS prediction. Then, we applied genome analyses and evolutionary/population genetics analyses to HS lncRNAs and their DBSs, applied transcriptome analysis to modern and archaic humans using the GTEx project data, and analyzed multiple experimental datasets. Our genome, transcriptome, and evolutionary analyses suggest that HS lncRNAs and their DBSs have continuously and distinctly rewired gene expression during human evolution. The rewired gene expression influences or determines diverse human traits, from an enlarged brain to modified metabolism pathways.

Study Overview. (A) The relationships between the chimpanzees, the three archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and the three modern human populations. Dashed lines indicate the phylogenetic distances from modern humans to archaic humans and chimpanzees. Genes and DBSs have the following implications – the DBS in B2M lacks a counterpart in chimpanzees; the DBS in ABL2 has great differences between archaic and modern humans; the DBS in IRNAR1 is highly polymorphic in modern humans. Red letters in DBSs indicate tissue-specific eQTLs or population-specific favored mutations. (B) DBSs’ mean length and binding affinity. (C) The target genes and transcripts of 66 HS lncRNAs. (D) The targeting relationships between HS lncRNAs. (E) The sequence distances of DBSs (the top 40%) from modern humans to chimpanzees and archaic humans. (F) The impacts of interactions between HS lncRNAs and DBSs on gene expression in GTEx tissues.

Results

HS lncRNAs regulate diverse genes and transcripts

After searching for orthologues of the 13562 GENCODE-annotated human lncRNA genes (v18) in 16 mammalian genomes and the genomes of archaic humans by RNA sequence and structure alignment using the Infernal program (Figure 1A) (Lin et al., 2019; Nawrocki, 2014; Nawrocki et al., 2009), we identified 66 HS lncRNA genes (Supplementary Table 1). We predicted HS lncRNAs’ DBSs in the 5000 bp promoter regions of the 179128 Ensembl-annotated transcripts (release 79) using the LongTarget program (Supplementary Note 1) (Lin et al., 2019). 105141 strong DBSs (mean binding affinity >60, mean length >147 bp) were identified in 96789 transcripts of 23396 genes, and 152836 weak DBSs (36< mean binding affinity <60) were identified in 127898 transcripts of 33185 genes (Figure 1B; Supplementary Table 2, 3). Genes and transcripts whose promoter regions contain the DBS of an HS lncRNA were assumed targets of the HS lncRNA. Several HS lncRNAs have DBSs in nearly all genes and transcripts (e.g., RP11-423H2.3 has strong DBSs in 73855 transcripts of 17941 genes), and only about 1.5% of target genes (0.6% of target transcripts) are human-specific genes and transcripts (Figure 1C), indicating that HS lncRNAs substantially rewire conserved genes and supporting the above rewiring hypothesis. Many transcripts have DBSs for multiple HS lncRNAs, many HS lncRNAs have multiple DBSs in a single transcript, and HS lncRNAs themselves show complex targeting relationships (Figure 1D; Supplementary Note 2), indicating the complexity of the rewiring.

HARs are conserved genomic loci but have evolved acceleratedly in the human lineage. Multiple researchers assume that HARs critically influence human evolution (especially brain evolution); however, results obtained from HAR studies are controversial (Levchenko et al., 2018). The Zoonomia Project recently identified 312 HARs important for 3D genome rewiring and neurodevelopment (Keough et al., 2023). We found that only 8 HARs overlap 26 DBSs of 14 HS lncRNAs, suggesting that DBS and HAR are sequences of different functions subject to separate examination.

We validated DBS prediction using multiple methods and datasets (Supplementary Note 1; Supplementary Fig 1-5). (a) HS lncRNA DBSs overlap well with experimentally identified DNA methylation and histone modification signals in multiple cell lines (https://genome.UCSC.edu). (b) Many DBSs co-localize with annotated cis-regulatory elements (cCREs) in genes’ promoter regions (https://genome.UCSC.edu). (c) predicted and experimentally detected DBSs of NEAT1, MALAT1, MEG3 overlap well (Mondal et al., 2015; West et al., 2014). (d) We used CRISPR/Cas9 to knock out predicted DBDs (100-200 bp) of seven lncRNAs (three HS lncRNAs and four wrongly transcribed lncRNAs) in multiple cell lines, performed RNA-seq before and after DBD knockout, and analyzed differential gene expression The |fold change| of target genes (genes with DBS) was significantly larger than the |fold change| of non-target genes (genes without DBS) (one-sided Mann-Whitney test, p = 3.1e-72, 1.49e-114, and 1.12e-206 for the HS lncRNA RP13-516M14.1, RP11-426L16.8, and SNORA59B, and p = 2.58e-09, 6.49e-41, 0.034, and 5.23e-07 for the four wrongly transcribed lncRNAs), suggesting that changed expression of target genes is caused by DBD knockout. (e) Finally, the LongTarget program uses the Smith-Waterman local alignment to identify DBD/DBS binding; the alignment algorithm determines that a DBS of 147 bp is extremely unlikely to be generated by chance (p < 8.2e-19 to 1.5e-48).

HS lncRNAs’ target genes characterize human evolution and traits

Since strong and weak DBSs may occur at different times in human evolution, we separately examined whether their target genes are enriched in functions important for human evolution. We sorted target genes by DBS binding affinity, chose the top 2000 genes (DBS length >252 bp and DBS binding affinity >151) and bottom 2000 genes (DBS length <60 bp but DBS binding affinity >36), and applied over-representation analysis using the g:Profiler program and Gene Ontology (GO) database to the two gene sets (Supplementary Table 4). The top and bottom 2000 genes are co-enriched in GO terms important for human evolution (instead of just brain evolution), including “neuron projection development” (GO:0031175), “learning” (GO:0007612), “regulation of anatomical structure size” (GO: GO:0090066), “head development” (GO:0060322), and “regulation of locomotion” (GO:0040012)) (Supplementary Note 3; Supplementary Table 4). 514 strongest DBSs have binding affinity >300 (DBS length >500 bp) (Supplementary Table 5), and genes containing these strongest DBSs include IFNAR1 and NFATC1 (important for immune function), KIF21B and NTSR1 (critical for neural development), SLC2A11 and SLC2A1 (involved in glucose usage), BAIAP3 and TAS1R3 (a brain-specific angiogenesis inhibitor and a receptor for the sweet taste response), and some primate-specific noncoding RNA genes (e.g., CTD-3224I3.3 which is highly expressed in the cerebellum, lung, and testis) (Table 1). Of note, genes with weak DBSs are enriched in GO terms including “response to alcohol” (GO:0097305), “response to temperature stimulus” (GO:0009266), “DNA methylation or demethylation” (GO:0044728), and “female pregnancy” (GO:0007565) (Supplementary Note 3; Supplementary Table 4). These findings help reveal how epigenetic regulation reshapes different human traits at different times.

The rewiring of gene expression has greatly evolved

The human genome is almost 99%, 99%, and 98% identical to the genomes of chimpanzees, bonobos, and gorillas (Chimpanzee Sequencing and Analysis Consortium, 2005; Prufer et al., 2012). Since where sequence differences are in these genomes is a question, we assessed to what extent HS lncRNAs’ DBS sequences are human-specific and found that 97.81% of strong DBSs have counterparts in chimpanzees. This suggests that many DBSs are not human-specific.

To examine the evolution of HS lncRNAs and their DBSs during human evolution, we identified their counterparts in Altai Neanderthals, Denisovans, and Vindija Neanderthals (Meyer et al., 2012; Prufer et al., 2017; Prüfer et al., 2013). HS lncRNAs and strong DBSs have counterparts in these archaic humans, but the distances of HS lncRNAs are smaller than the distances of DBSs between archaic and modern humans (“distance” means distance per base), supporting that many DBSs occur in ape genomes instead of human-specific. We computed DBS distances in two ways. The first is from the reconstructed human ancestor (downloaded from the EBI website) to chimpanzees, archaic humans, and modern humans; the second is from chimpanzees and archaic humans to modern humans. The first type of distances has the anomaly that many DBSs’ distances from the human ancestor to modern humans are smaller than to archaic humans (for example, DBSs in the top 20% (5033) genes in the human ancestor have distances >0.015 to modern humans, but more DBSs have distances >0.015 to Altai Neanderthals, Denisovans, and even chimpanzees) (Supplementary Note 3). This anomaly may be caused by the human ancestor being built using six primates without archaic humans; therefore, no analysis was followed. In contrast, the second type of distances agrees well with the phylogenetic distances between chimpanzees, archaic humans, and modern humans. Notable, large sequence changes occurred in limited DBSs (Figure 1E), and genes containing these greatly changed DBSs may critically characterize human evolution. Given the high similarity between human and chimpanzee genomes, the top 20% (4248) genes in chimpanzees may sufficiently characterize the human-chimpanzee differences, and these genes have DBS distances >0.034. Taking this as a threshold, 1256, 2514, and 134 genes in Altai Neanderthals, Denisovans, and Vindija Neanderthals have DBS distance >0.034, and these numbers agree with the phylogenetic relationships between these archaic and modern humans (Figure 1A; Supplementary Table 6).

To examine target genes with mostly changed DBSs are enriched in what functions, we applied over-representation analysis using the g:Profiler program and Gene Ontology (GO) database to target genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals. To make statistical significance comparable, we used the top 1256 genes in chimpanzees and three archaic humans as the inputs. The 1256 genes in chimpanzees and Altai Neanderthals are enriched in many common GO terms; however, “behaviour” (GO:0007610) and “cognition” (GO:0050890) have much higher significance in Altai Neanderthals than in chimpanzees (Figure 2). In addition, enrichment of such GO terms as “learning” (GO:0007612), “learning or memory” (GO:0007611), and “glucose homeostasis” (GO:0042593) is specific to Altai Neanderthals. Agreeing with the close relationship between Vindija Neanderthals and modern humans, the 134 genes in Vindija Neanderthals are enriched only in two GO terms with low significance (“smoothened signalling pathway” (GO:0007224) and “regulation of smoothened signalling pathway” (GO:0008589)) (Supplementary Table 7). These results help reveal the periods and positions of epigenetic regulation changes and help infer the periods of related phenotypic changes.

1256 target genes whose DBSs have largest distances from modern humans to chimpanzees and Altai Neanderthals are enriched in different Biological Processes GO terms. Upon significance threshold = 0.05 (Benjamini-Hochberg FDR), the two gene sets in chimpanzees and Altai Neanderthals are enriched in 199 and 409 GO terms (50 < terms size < 1000), respectively. Pink color indicates shared GO terms. Left: Top GO terms enriched in genes in chimpanzees. Right: Top GO terms enriched in genes in Altai Neanderthals.

Epigenetic regulation by HS lncRNAs shows differences in modern humans

To check whether the rewiring is still going on, we analysed single nucleotide polymorphisms (SNPs, with minimal allele frequency (MAF) ≥0.1) in DBSs using the 1000 Human Genome data (1000 Genomes Project Consortium et al., 2012). Many DBSs are highly polymorphic (>10 SNPs, 5 times the average number of SNPs for all DBSs); many DBSs have mostly changed sequence distances from modern humans to archaic humans and chimpanzees; and the intersection of the above two sets are in genes important for human evolution (Supplementary Table 8), including IFNAR1, NFATC1, TAS1R3, INS, ST3GAL4, and FN3KRP (Table 2). SNPs inside these DBSs, whether favored or recessive, affect DBS binding affinity in modern humans. A notable case is sugar metabolism. A feature of modern human diets is high sugar intake, which causes non-enzymatic oxidation of proteins (i.e., glycation) and deactivates protein functions. Among proteins encoded by genes whose DBSs are most SNP-rich and have mostly changed sequence distances, TAS1R3 recognizes diverse natural and synthetic sweeteners, Insulin decreases blood glucose concentration, ST3GAL4 regulates protein glycosylation, and FN3KRP deglycates proteins. DBSs in these genes not only have counterparts with mostly changed sequences in chimpanzees and archaic humans, but also are SNP-rich, indicating continuous evolution of these genes’ regulation by HS lncRNAs.

We next used multiple statistical tests, including XP-CLR (Chen et al., 2010), iSAFE (Akbari et al., 2018), Tajima’s D (Tajima, 1989), Fay-Wu’s H (Fay and Wu, 2000), the fixation index (Fst) (Weir and Cockerham, 1984), and linkage disequilibrium (LD) (Slatkin, 2008), to detect selection signals in HS lncRNAs and strong DBSs in the modern human CEU, CHB, and YRI (the representative populations of Europeans, Asians, and Africans). These tests detected selection signals in several HS lncRNA genes in CEU and CHB but not in YRI and selection signals in more DBSs in CEU and CHB than in YRI (Supplementary Note 4, 5; Supplementary Table 9, 10, 11). Examples include the DBSs in the two pigmentation-related genes MC1R and MFSD12 and in the two odor reception-related genes OR6C1 and TAS1R3. The enriched SNPs, detected selection signals, and annotated gene functions suggest specific and ongoing influence of rewired gene expression on human evolution.

SNPs in DBSs have cis-effects on gene expression

To examine the impact of DBSs on gene expression, we assessed the expression quantitative trait loci (eQTLs) in DBS sequences using the Genotype-Tissue Expression (GTEx) data (GTEx Consortium, 2017). 1727 SNPs with MAF ≥0.1 in at least one population and with absolute values of cis-effect size (ES) ≥0.5 in at least one tissue were identified in DBSs in autosomal genes (Supplementary Table 12,13). These eQTLs include 372 “conserved” ones (also in DBSs in the three archaic humans) and 1020 “novel” ones (in DBSs only in modern humans). Many conserved eQTLs are in brain tissues and have high derived allele frequencies (DAFs) in all three modern populations. A notable eQTL that has high DAFs in all three populations and positive ES ≥0.5 in 44 out of 48 GTEx tissues is rs2246577 in the DBS in FN3KRP, whose encoded protein deglycates proteins to restore their function (Table 2). In contrast, many novel eQTLs are in a particular tissue and have population-specific DAFs (Supplementary Note 6).

We performed two analyses to examine how eQTLs in DBSs influence gene expression via lncRNA-DBS interactions. First, we computed the expression correlation between HS lncRNAs and target transcripts with DBSs having eQTLs in specific tissues. Most (94%) HS lncRNA-target transcript pairs showed significant expression correlation in tissues in which the eQTLs were identified (|Spearman’s rho| >0.3 and FDR <0.05). Second, we examined whether eQTLs are more enriched in DBSs than in promoter regions by computing and comparing the density of eQTLs in each DBS and each Ensembl-annotated promoter. The results suggest that eQTLs are more enriched in DBSs than in promoters (one-sided Mann-Whitney test, p = 0.0) and that these eQTLs influence gene expression via lncRNA-DBS interactions (Supplementary Note 6). Tissue-specific eQTLs and population-specific SNPs in DBSs confirm the cis-effects of HS lncRNA-mediated epigenetic regulation on gene expression.

HS lncRNAs contributed to brain evolution from archaic to modern humans

We finally examined the impact of HS lncRNA-DBS interaction on gene expression in the GTEx tissues (GTEx Consortium, 2017). 40 autosomal HS lncRNAs are expressed robustly (median TPM > 0.1) in at least one tissue, and they form 198876 pairs with target transcripts in all tissues. 45% of these pairs show a significant expression correlation in specific tissues (Spearman’s |rho| >0.3 and FDR <0.05). In contrast, when randomly sampling 10000 pairs of lncRNAs and protein-coding transcripts genome-wide, the percent of pairs showing this level of expression correlation (Spearman’s |rho| >0.3 and FDR <0.05) is only 2.3%. Equally notable is that 56% of HS lncRNA-target transcript pairs with a significant correlation are present in at least one brain region (Figure 3A), indicating substantial gene expression regulation by HS lncRNAs in the brain.

The impact of HS lncRNA-DBS interaction on gene expression in GTEx tissues and organs. (A) The percentage distribution of correlated HS lncRNA-target transcript pairs across GTEx tissues and organs. Higher percentages of correlated pairs are in the brain than in other tissues and organs. (B) The distribution of significantly changed DBSs in HS lncRNA-target transcript pairs across GTEx tissues and organs between archaic and modern humans. Orange, red, and dark red indicate significant DBS changes from Denisovans (D), Altai Neanderthals and Denisovans (A.D.), and all of the three archaic humans (ADV). DBSs in HS lncRNA-target transcript pairs correlated in seven brain regions (in dark red) have changed significantly and consistently since the Altai Neanderthals, Denisovans, and Vindija Neanderthals (one-sided two-sample Kolmogorov-Smirnov test, significant changes determined by FDR <0.001).

We analysed nine experimental datasets to verify the above conclusion substantial gene expression regulation by HS lncRNAs in the brain (Supplementary Note 7). First, we analysed two datasets of epigenetic studies (one examined H3K27ac and H3K4me2 profiling of human, macaque, and mouse corticogenesis, and the other examined gene expression and H3K27ac modification in eight brain regions in humans and four other primates) (Reilly et al., 2015; Xu et al., 2018). 84% and 73% of genes in the two datasets have DBSs for HS lncRNAs, significantly more enriched in HS lncRNA targets than the genome-wide background (p = 1.21e-21 and 1.2e-56, two-sided Fisher’s exact test). Second, we analysed two datasets of PsychENCODE studies (one examined spatiotemporally differentially expressed genes and spatiotemporally differentially methylated sites in 16 brain regions, and the other examined the spatiotemporal transcriptomic divergence across human and macaque brain development) (Li et al., 2018; Zhu et al., 2018). In the first dataset, 65 HS lncRNAs are expressed in all 16 brain regions, 109 transcripts have spatiotemporally differentially methylated sites in their promoters, and 56 of the 109 transcripts have DBSs for HS lncRNAs. In the second dataset, 8951 genes show differential expression between the human and macaque brains, and 72% of differentially expressed genes in the human brain have DBSs of HS lncRNAs. Thus, both datasets show significant enrichment for HS lncRNA regulation. Third, three studies identified several genes critically regulating cortical expansion (Florio et al., 2018; Johnson et al., 2018; Suzuki et al., 2018). Of the 40 protein-coding genes reported by these studies, 29 have DBSs of HS lncRNAs in their promoter regions (Supplementary Table 14). Thus, these genes are enriched with DBSs of HS lncRNAs compared with the genome-wide background (p < 0.01, two-sided Fisher’s exact test). Finally, we analysed two datasets of brain organoids studies. By establishing and comparing cerebral organoids between humans, chimpanzees, and macaques, Pollen et al. identified 261 human-specific gene expression changes (Pollen et al., 2019). By fusing human and chimpanzee iPS cells and differentiating the hybrid iPS cells into hybrid cortical spheroids, Agoglia et al. generated a panel of tetraploid human-chimpanzee hybrid iPS cells and identified thousands of genes with divergent expression between humans and chimpanzees (Agoglia et al., 2021). The 261 and 1102 genes in the two datasets are enriched with DBSs of HS lncRNAs compared with the genome-wide background (p = 1.2e-16 and 3.4e-74).

We next examined the evolution of the impact of HS lncRNAs and their DBSs on gene expression in the GTEx tissues. For each DBS in an HS lncRNA-target transcript pair correlated in a GTEx tissue, we computed the sequence distance of the DBS from modern humans to archaic humans. Then, we compared the distribution of DBS sequence distances in each tissue with the distribution of DBS sequence distances in all tissues (one-sided two-sample Kolmogorov-Smirnov test). DBSs in HS lncRNA-target transcript pairs correlated in brain regions have significantly changed sequence distances since the Altai Neanderthals and Denisovans (Figure 3B), suggesting that gene expression regulation by HS lncRNAs in the brain has experienced significant evolution and that HS lncRNAs contributed greatly to recent brain evolution (Ponce de Leon et al., 2021). Results from other GTEx tissues, including Kidney-Cortex, Ovary, and Skin-Sun Expressed (Figure 3B), also suggest continuous influences of rewired gene expression on human evolution.

Discussion

Modern humans differ from archaic humans and great apes trivially in genomes but greatly in phenotypes, making what genomic differences critically determine the phenotypic differences a profound question. This question has been addressed by many researchers along multiple approaches, and attention has been especially paid to new protein-coding genes, HARs, and evolved TF-TFBS interactions. Findings include HS protein-coding genes promote rapid neocortex expansion (Fiddes et al., 2018; Florio et al., 2018; Pinson et al., 2022), HARs are significantly enriched in domains important for human-specific 3D genome organization (Keough et al., 2023), and gene expression is rewired by TFs and TFBSs (Johnson, 2017). These findings suggest that species-specific lncRNAs and their DBSs may also substantially rewire gene expression. This study identified HS lncRNAs, their DBSs, and counterparts of these DBSs in ancient humans and multiple apes. Our analyses of the evolution of and interaction between HS lncRNAs and their DBSs generate evidence suggesting that HS lncRNAs and their DBSs have contributed greatly to human evolution by distinctly rewiring gene expression. Our analyses of multiple experimental datasets support this conclusion. The rewiring can be dated back to chimpanzees, has significantly evolved from archaic to modern humans, and still influences differences in modern humans. The occurrence of HS lncRNAs and their DBSs may have three situations – (a) HS lncRNAs preceded their DBSs, (b) HS lncRNAs and their DBSs co-occurred, (c) HS lncRNAs succeeded their DBSs. Our results support the third situation and the rewiring hypothesis.

We performed seven CRISPR/Cas9 experiments to validate DBS prediction and analysed multiple experimental datasets to support the main conclusion. Parameters and thresholds may influence data analysis results. When running Infernal, LongTarget, MAFFT7, and Phylip, default parameters were used. When analysing enrichment of GO terms in gene lists with strong and weak DBSs, the top and bottom 2000 genes were used, with the whole genome as the background. When analyzing enrichment of GO terms in gene lists with DBSs with large sequence distances, the threshold of 0.034 was used. DBSs in 4248, 1256, 2514, and 134 genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals have distance >0.034 to modern humans, and these numbers sufficiently characterize the human-chimpanzee differences and qualitatively agree with the phylogenetic relationships. Other parameter and threshold values would generate somewhat different results, but enriched GO terms with high significance will be quite stable. We note that most of the reported are findings rather than conclusions, indicative or suggestive of something that answer the primary question of what genomic differences critically determine the phenotypic differences between humans and apes and between modern and archaic humans.

Several findings are noteworthy. First, Neanderthal-inherited sequences have measurable impacts on gene expression in modern humans, and the impacts are least detectable in the brain (McCoy et al., 2017). Here we find HS lncRNAs have more significantly influenced gene expression in the brain than in other tissues. Second, the regulation of many genes by HS lncRNAs has undergone great evolution, specific examples including those making humans adapt to high sugar intake. Third, intensively regulated genes are enriched in different functions at different periods of human evolution. Specifically, “old” SNPs in DBSs are in genes regulating neural development, and “new” SNPs in DBSs are in genes regulating such processes as glucose metabolism. These findings reveal the diverse, continuous, and subtle contribution of HS lncRNA to human evolution.

The above findings not only reveal HS lncRNAs’ functions and adaptive human evolution, but also raise multiple new questions. First, the protein-coding regions in the mouse and human genomes are about 85% identical; however, many mouse lncRNAs are rodent-specific (Yue et al., 2014). Thus, two questions are how mouse-specific lncRNAs specifically rewire gene expression in mice and how human- and mouse-specific rewiring influences the cross-species transcriptional differences (Breschi et al., 2017; Hodge et al., 2019; Zhu et al., 2018). Second, whether mice and other mammals have lncRNAs such as RP11-423H2.3 that substantially rewires gene expression species-specifically. Third, to what extent clade- and species-specific lncRNAs would make speciation and evolution preordained. These interesting questions have practical importance.

Material and methods

Data sources

The sequences of human (GRCh38/hg38, GRCh37/hg19) and chimpanzee genomes (panTro5) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). Three high-quality archaic human genomes were obtained from the Max Planck Institute for Evolutionary Anthropology (https://www.eva.mpg.de/genetics/index.html), which include an Altai Neanderthal which lived approximately 122 thousand years ago (kya) (Pru fer et al., 2013), a Denisovan (an Asian relative of Neanderthals) which lived approximately 72 kya (Meyer et al., 2012), and a Vindija Neanderthal which lived approximately 52 kya (Prufer et al., 2017). The ancestral sequences for the human genome (GRCh37) (which were generated upon the EPO alignments of six primate species human, chimpanzee, gorilla, orangutan, macaque, and marmoset) were downloaded from the EBI website (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/). The SNP data of modern humans were obtained from the 1000 Genomes Project (phase I) (1000 Genomes Project Consortium et al., 2012). The three modern human populations are CEU (Utah residents with Northern and Western European ancestry), CHB (Han Chinese in Beijing, China), and YRI (Yoruba in Ibadan, Nigeria), which contain 100, 99, and 97 individuals, respectively. The genetic map files of modern humans were downloaded from the 1000 Genomes Project website (ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/technical/working/20110106_recombination_hotspots/). The 1000 Genomes phased genotype data in VCF format were downloaded from the International Genome Sample Resource website (http://www.internationalgenome.org/). Annotated human genes and transcripts were obtained from the Ensembl website (http://www.Ensembl.org). FASTA-format nucleotide sequences were converted from VCFs using the vcf-consensus program in the VCFtools package when necessary (Danecek et al., 2011). Data in the GeneCards Human Gene Database (www.genecards.org) were used to annotate genes and transcripts.

RNA-sequencing (RNA-seq) data (transcripts per million [TPM]) and eQTL data of human tissues were obtained from the Genotype-Tissue Expression Project (GTEx, v8) website (https://gtexportal.org/) (GTEx Consortium, 2017). Histone modification and DNA methylation signals in multiple cell lines and ENCODE Candidate Cis-Regulatory Elements (cCRE) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). The genome-wide DNA binding sites of lncRNA NEAT1, MALAT1, and MEG3 were obtained from experimental studies (Mondal et al., 2015; West et al., 2014). The predicted and experimentally detected DBSs in target genes of MALAT1, NEAT1, and MEG3 are given in supplementary tables (Supplementary Table 15).

Identifying HS lncRNA genes

We used the Infernal program (Nawrocki et al., 2009), which searches orthologous RNA sequences upon sequence and structure alignment, to identify orthologous exons in 16 mammalian genomes for each exon in each of the 13562 GENCODE (v18)-annotated human lncRNA genes (Lin et al., 2019). The 16 mammals were chimpanzee (CSAC 2.1.4/panTro4), macaque (BGI CR_1.0/rheMac3), marmoset (WUGSC 3.2/calJac3), tarsier (Broad/tarSyr1), mouse lemur (Broad/micMur1), tree shrew (Broad/tupBel1), mouse (GRCm38/mm10), rat (Baylor3.4/rn4, RGSC6.0/rn6), guinea pig (Broad/cavPor3), rabbit (Broad/oryCun2), dog (Broad CanFam3.1/canFam3), cow (Baylor Btau_4.6.1/bosTau7), elephant(Broad/loxAfr3), hedgehog (EriEur2.0/eriEur2), opossum (Broad/monDom5), and platypus (WUGSC 5.0.1/ornAna1) (http://genome.UCSC.edu). If the number of orthologous exons of a human lncRNA gene in a genome exceeded half the exon number of the human lncRNA gene, these orthologous exons were assumed to form an orthologous lncRNA gene. If a human lncRNA gene had no orthologous gene in all of the 16 mammals, it was assumed to be human-specific.

Identifying DBSs of HS lncRNAs

LncRNAs bind to DNA sequences by forming RNA:DNA triplexes. Each triplex comprises triplex-forming oligonucleotides (TFO) in the lncRNA and a triplex-targeting site (TTS) in the DNA sequence. We used the LongTarget program to predict HS lncRNAs’ DNA binding domains (DBD) and binding sites (DBS) with the default parameters (Ruleset = all, T.T. penalty = −1000, CC penalty = 0, Offset = 15, Identity ≥ 60, Nt ≥ 50) (Lin et al., 2019; Wen et al., 2022). LongTarget program predicts DBDs and DBSs simultaneously, with a DBD comprising a set of densely overlapping TFOs and a DBS comprising a set of densely overlapping TTSs. For each HS lncRNA, we predicted its DBSs in the 5000 bp promoter regions (−3500 bp upstream and +1500 bp downstream the transcription start site) of the 179128 Ensembl-annotated transcripts (release 79). For each DBS, a binding affinity was calculated by multiplexing the DBS length and the averaged Identity score of all TTSs (the Identity score is the percentage of paired nucleotides). Strong and weak DBSs were classified upon binding affinity ≥ 60 and < 60. A transcript whose promoter region contains a strong DBS of an HS lncRNA was assumed to be a target transcript of the HS lncRNA, and the gene containing this transcript was assumed to be a target gene of the HS lncRNA. As the 1000 Genomes Project (phase I) data and the archaic human genomes are based on GRCh37/hg19, the coordinates of DBSs were converted from GRCh38/hg38 to GRCh37/hg19 using the liftover program from the UCSC Genome Browser (Kuhn et al., 2013).

Experimentally validating DBS prediction

A 157 bp sequence (chr17:80252565-80252721, hg19) containing the DBD of RP13-516M14.1, a 202 bp sequence (chr1:113392603-113392804, hg19) containing the DBD of RP11-426L16.8, and a 198 bp sequence (chr17:19460524-19460721, hg19) containing the DBD of SNORA59B, were knocked out in the HeLa cell line, RKO cell line, and SK-MES-1 cell line, respectively. Two sequences (chr1:156643524-156643684, chr10:52445649-52445740, hg38) containing the DBD of two wrongly transcribed noncoding sequences were knocked out in the HCT-116 and A549 cell lines, respectively. The seven knockouts were performed by UBIGENE, Guangzhou, China (http://www.ubigene.com) using CRISPR-UTM, a revised version of CRISPR/Cas9 technology. Before and after the seven DBD knockouts, RNA-sequencing (RNA-seq) was performed by Novogene, Beijing, China (https://cn.novogene.com) and HaploX, Shenzhen, China (https://www.haplox.cn/). The reads were aligned to the human GRCh38 genome using the Hiasat2 program (Kim et al., 2019), and the resulting Sam files were converted to bam files using Samtools (Li et al., 2009). The Stringtie program was used to quantify gene expression levels (Pertea et al., 2015). Fold change of gene expression was computed using the edgeR package (Robinson et al., 2010), and significant up- and down-regulation of target genes after DBD knockout was determined upon |log2(fold change)| > 1 with FDR < 0.1.

Genome-wide DBSs of NEAT1, MALAT1, and MEG3 were experimentally detected (Mondal et al., 2015; West et al., 2014). We also used these data to validate DBS prediction by predicting DBSs of the three lncRNAs and checking the overlap between predicted and experimentally detected DBSs (Supplementary Table 15).

Mapping DBSs of HS lncRNAs in the chimpanzee and archaic human genomes

We used the liftover program from the UCSC Genome Browser (http://genome.UCSC.edu) to map loci of DBSs in the human genome hg38 to the chimpanzee genome Pan_tro 3.0 (panTro5). The mapping results were examined by checking the human-chimpanzee pairwise alignment (hg38 vs panTro5) in the UCSC Genome Browser (http://genome.UCSC.edu). 97.81% (100403/102651) of DBSs were mapped, and the 2248 unmapped DBSs in 429 genes were considered human-specific. We used vcf-consensus in the VCFtools package to extract the DBSs of HS lncRNAs from the VCF files of Altai Neanderthals, Denisovans, and Vindija Neanderthals. The variant with the highest quality score was selected whenever multiple variant calls were observed at a given locus. The obtained DBS sequences in the chimpanzees and three archaic humans are called counterparts of DBSs in these genomes.

Estimating sequence distances of DBSs between different genomes

We first aligned DBS sequences in the genomes of humans, chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals using the MAFFT7 program for measuring sequence distances from modern humans to chimpanzees and archaic humans (Katoh and Standley, 2013). We then computed sequence distances using the dnadist program with the Kimura 2-parameter model in the PHYLIP (3.6) package (http://evolution.genetics.washington.edu/phylip.html) and using the Tamura-Nei model in the MEGA7 package (Kumar et al., 2016). The two methods generated equivalent results. The largest distance between DBSs in humans and their counterparts in chimpanzees is 5.383. Since 2248 DBSs in 429 human genes do not have counterparts in chimpanzees, we assumed that the sequence distances of these DBSs between humans and chimpanzees were 10.0. Since the top 20% of DBSs in chimpanzees (including these 2248 DBSs) have distances ≥ 0.034, we used 0.034 as the threshold to identify the “mostly changed” DBSs in the chimpanzees and three archaic humans.

We determined human ancestral sequences of DBSs upon the human ancestor sequences from the EBI website, which were generated upon the EPO alignments of six primate species. We used the above-mentioned methods to compute DBSs’ sequence distances from the human ancestor to chimpanzees, archaic humans, and modern humans. We found that when the human-chimpanzee ancestral sequence has the ancestral sequence (which means the inferring of ancestral allele is of high confidence), DBS distances from the human ancestor to modern humans are larger than to archaic humans, but this situation only accounts for about 63.8%. For many DBSs, the distances from the human ancestor to modern humans are smaller than to archaic humans (especially Neanderthals and Denisovans) and even to chimpanzees. This defect may be caused by the absence of archaic humans in building the human ancestral sequence.

Detecting positive selection signals in HS lncRNA genes and DBSs

We used multiple tests to detect positive selection signals in HS lncRNA genes. First, we used the XP-CLR test (parameters = -w1 0.001 300 100 -p0 0.95, window size = 0.1 cM, grid size = 100 bp) to perform six pairwise genome-wide scans (i.e., CEU-CHB, CEU-YRI, CHB-CEU, CHB-YRI, YRI-CEU, and YRI-CHB) (Chen et al., 2010). The upper 1% of scores across the entire genome in each pairwise scan were 34.6 in the CEU-YRI scan, 16.8 in the CEU-CHB scan, 45.0 in the CHB-YRI scan, 26.9 in the CHB-CEU scan, 14.1 in the YRI-CEU scan, and 14.1 in the YRI-CHB scan. These scores were used as the thresholds of positive selection signals in these populations. Second, we used the iSAFE program to scan each genomic region that contains a HS lncRNA gene and its 500-kb upstream and downstream sequences (Akbari et al., 2018). Strongly selected loci were detected only in CEU and CHB. Third, we used the VCFtools program to calculate Tajima’s D values for each HS lncRNA gene in CEU, CHB, and YRI (Danecek et al., 2011). The calculation was performed using a 1500-bp non-overlapping sliding window because the lengths of these genes exceed 1500 bp. To generate the background reference for judging the significant increase or decrease of Tajima’s D for HS lncRNA genes in a population, we calculated Tajima’s D values across the whole genome using a sliding window of 1500 bp. As values of Tajima’s D were compared with the background reference, a significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively, instead of population demography dynamics (Tajima, 1989). Fourth, we used the integrated Fst to detect positive selection signals in HS lncRNA genes. The Fst of each HS lncRNA gene was computed for three comparisons, i.e., CEU-YRI, CHB-YRI, and CHB-CEU. Extreme Fst values of SNPs were detected in HS lncRNA genes in the comparisons of CEU-YRI and CHB-YRI. Since allele frequencies for different loci vary in a genome and genetic drift may have different effects at different loci, we used a sliding window of 1500 for comparing Fst values of HS lncRNA genes with the genome-wide background. Extreme Fst values indicate positive selection. Finally, we applied linkage disequilibrium (LD) analysis to each HS lncRNA gene. We computed the pairwise LD (r²) in CEU, CHB, and YRI for common SNPs (with minimal allele frequency (MAF) ≥ 0.05 in at least one population) in HS lncRNA genes and DBSs using the PLINK program (Purcell et al., 2007). Significantly increased LD was detected in SNPs in HS lncRNA genes in CEU and CHB. The LD patterns were portrayed using the Haploview program (Barrett et al., 2005).

Next, we used the above tests to detect positive selection signals in DBSs. First, the 100 bp grid size of the XP-CLR test also allowed the detection of selection signals in DBSs. Second, we performed the Tajima’s D and Fay-Wu’s H tests (Fay and Wu, 2000; Tajima, 1989). We calculated Tajima’s D values for each DBS in CEU, CHB, and YRI using the VCFtools program (Danecek et al., 2011), with a sliding window of 147 bp (the mean length of strong DBSs). To generate the background reference for judging the significant increase or decrease of Tajima’s D in a population, we calculated Tajima’s D values across the whole genome using the same sliding window. As values of Tajima’s D were compared with the background reference, a significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively. Fay-Wu’s H values were calculated similarly using the VariScan program (Vilella et al., 2005). Calculating Fay-Wu’s H demands the ancestral sequences as the outgroup. We extracted the ancestral sequences of DBSs from the human ancestral sequence, which was generated upon the EPO alignments of six primate species (human, chimpanzee, gorilla, orangutan, macaque, and marmoset) (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignm ents/). Third, we computed the Fst to measure the frequency differences of alleles in DBSs between populations. For the CEU-YRI, CHB-YRI, and CHB-CEU pairwise comparisons, we used the revised VCFtools program to compute the weighted Fst for all SNPs in each DBS (Weir and Cockerham, 1984). Fourth, we integrated the weighted Fst values in the three populations into an “integrated Fst” which indicated whether the DBS locus was under selection in a certain population (Nielsen and Slatkin, 2013). We used sliding windows of 147 bp for comparing Fst values of DBSs with the genome-wide background. We empirically denoted the values of integrated Fst in the upper 10% of scores across the entire genome as statistically significant. To determine positive selection more reliably, we used Tajima’s D and the integrated Fst to jointly determine if a DBS was under positive selection in a population. The thresholds that determined the upper 10% of Tajima’s D values across the entire genome in CEU, CHB, and YRI were −0.97, −0.96, and −0.97, respectively, and the threshold that determined the upper 10% of integrated Fst values across the entire genome in the three populations was 0.22. For example, a DBS was assumed to be under positive selection in CEU if (a) the DBS had a Tajima’s D<-0.97 in CEU and Tajima’s D>0.0 in the two other populations and (b) the DBS had an integrated Fst>0.22. Analogously, a DBS was assumed to be under positive selection in both CEU and CHB if the DBS had a Tajima’s D <-0.97 in CEU, <-0.96 in CHB, and >0.0 in YRI, and had an integrated Fst>0.22.

Functional enrichment analysis of genes

We used the g:Profiler program (with the parameter settings: Organism=Homo sapiens, Ordered query = No, Significance threshold = Benjamini-Hochberg FDR, User threshold = 0.05, 50 < terms size < 1000) and the Gene Ontology (GO) database to perform over-representation analysis (Raudvere et al., 2019). This analysis is used to determine what pre-defined gene sets (GO terms) are more present (over-represented) in a list of “interesting” genes than what would be expected by chance. The lists of genes included genes with strongest DBSs in humans, genes with weakest DBSs in humans, and genes with mostly changed DBSs in chimpanzees and three archaic humans. To choose genes with strong DBSs, we ranked genes in descending order based on binding affinity and chose the top 2000 genes. To choose genes with weak DBSs, we first ranked genes in ascending order based on binding affinity, then ranked genes in descending order based on the number of regulatory HS lncRNAs, and finally chose the top 2000 genes with binding affinity ≤40 and regulatory HS lncRNAs ≥5. We used the distance threshold of 0.034 to choose genes with mostly changed DBSs. Genes with mostly changed DBSs in chimpanzees include the 429 genes in which 2248 DBSs lack counterparts. 4248, 1256, 2514, and 134 genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals have DBSs with distance ≥0.034. To make the statistical significance comparable across chimpanzees, Altai Neanderthals, and Denisovans, we used the same number of genes (the top 1256 genes whose DBSs have the largest DBS distances) as the input to the g:Profiler program.

Analysing SNP frequencies in human populations

The frequency of common SNPs (MAF ≥0.05) in DBSs in the three modern human populations was computed using the VCFtools package (Danecek et al., 2011). The ancestral/derived states of SNPs were inferred upon the human ancestor sequences and were used to determine derived allele frequency.

Analysing the cis-effect of SNPs in DBSs on target gene expression

SNPs with MAF >0.1 in DBSs in any of the three modern human populations and absolute values of cis-effect size >0.5 (FDR <0.05) in any of the GTEx tissues were examined for an influence on the expression of the target genes (GTEx Consortium, 2017). SNPs that are eQTLs in the GTEx tissues and have biased derived allele frequencies (DAF) in the three modern human populations were examined to estimate whether the eQTL is population-specific.

Examining the tissue-specific impact of HS lncRNA-regulated gene expression

First, we examined the expression of HS lncRNA genes across the GTEx tissues. HS lncRNA genes with a median TPM value >0.1 in a tissue were considered robustly expressed in that tissue. Upon this criterion, 40 HS lncRNA genes were expressed in at least one tissue and were used to examine the impact of HS lncRNA regulation on gene and transcript expression (other HS lncRNAs may function in the cytoplasm). Since a HS lncRNA gene may have multiple transcripts, we adopted the transcript, which contains the predicted DNA binding domain (DBD) and has the largest TPM value as the representative transcript of the HS lncRNA. We calculated the pairwise Spearman’s correlation coefficient between the expression of a HS lncRNA (the representative transcript) and the expression of each of its target transcripts using the scipy.stats.spearmanr program in the scipy package. The expression of a HS lncRNA gene and a target transcript was considered to be significantly correlated if the |Spearman’s rho| > 0.3, with Benjamini-Hochberg FDR <0.05. We examined the percentage distribution of correlated HS lncRNA-target transcript pairs across GTEx tissues and organs (Figure 3A).

We examined all GTEx tissues to determine in which tissues HS lncRNA-regulated gene expression may have undergone significant changes from archaic to modern humans. If a pair of HS lncRNA and target transcript are robustly expressed in a tissue and their expression shows a significant correlation (|Spearman’s rho| >0.3, with Benjamini-Hochberg FDR <0.05) in the tissue, we computed the sequence distance of the HS lncRNA’s DBS in the transcript from the three archaic humans to modern humans. We compared the sequence distances of all DBSs in each tissue with the sequence distances of all DBSs in all GTEx tissues (i.e. the distribution of all DBS distances was taken as the background). A one-sided two-sample Kolmogorov-Smirnov test was used to examine whether the sequence distances of all DBSs in a specific tissue deviate from the background distribution (which reflects the “neutral evolution” of gene expression). For a tissue, if the test result was Benjamini-Hochberg FDR < 0.001, the tissue was considered to have significantly changed gene expression regulated by the HS lncRNA. We used different colors to mark tissues with significantly changed gene expression regulation since Altai Neanderthals, Denisovans, and Vindija Neanderthals, and used “D”, “A.D.”, and “ADV” to indicate changes since Denisovans, since Altai Neanderthals and Denisovans, and since Altai Neanderthals, Denisovans, and Vindija Neanderthals, respectively (Figure 3B).

Supporting information

eLife Assessment

This valuable study uses population and functional genomics to examine long non-coding RNAs (lncRNAs) in the context of human evolution. Computational prediction of human-specific lncRNAs and their DNA binding sites and analyses of these loci lead to the development of hypotheses regarding the potential roles of these genetic elements in human biology. The evidence supporting the conclusions is, however, still incomplete, as key details regarding the methodology and analyses are lacking.

Reviewer #1 (Public Review):

Summary

While DNA sequence divergence, differential expression, and differential methylation analysis have been conducted between humans and the great apes to study changes that "make us human", the role of lncRNAs and their impact on the human genome and biology has not been fully explored. In this study, the authors computationally predict HSlncRNAs as well as their DNA Binding sites using a method they have developed previously and then examine these predicted regions with different types of enrichment analyses. Broadly, the analysis is straightforward and after identifying these regions/HSlncRNAs the authors examined their effects using different external datasets.

I no longer have any concerns about the manuscript as the authors have addressed my comments in the first round of review.

Reviewer #2 (Public Review):

Lin et al attempt to examine the role of lncRNAs in human evolution in this manuscript. They apply a suite of population genetics and functional genomics analyses that leverage existing data sets and public tools, some of which were previously built by the authors, who clearly have experience with lncRNA binding prediction. However, I worry that there is a lack of suitable methods and/or relevant controls at many points and that the interpretation is too quick to infer selection. While I don't doubt that lnc RNAs contribute to the evolution of modern humans, and certainly agree that this is a question worth asking, I think this paper would benefit from a more rigorous approach to tackling it.

I thank the authors for their revisions to the manuscript; however, I find that the bulk of my comments have not been addressed to my satisfaction. As such, I am afraid I cannot say much more than what I said last time, emphasising some of my concerns with regards to the robustness of some of the analyses presented. I appreciate the new data generated to address some questions, but think it could be better incorporated into the text - not in the discussion, but in the results.

Author Response:

The following is the authors’ response to the previous reviews

Reviewer #1 (Public Review):

Summary

While DNA sequence divergence, differential expression, and differential methylation analysis have been conducted between humans and the great apes to study changes that "make us human", the role of lncRNAs and their impact on the human genome and biology has not been fully explored. In this study, the authors computationally predict HSlncRNAs as well as their DNA Binding sites using a method they have developed previously and then examine these predicted regions with different types of enrichment analyses. Broadly, the analysis is straightforward and after identifying these regions/HSlncRNAs the authors examined their effects using different external datasets.

I no longer have any concerns about the manuscript as the authors have addressed my comments in the first round of review.

We thank the reviewer for the valuable comments, which have helped us improve the manuscript.

Reviewer #2 (Public Review):

Lin et al attempt to examine the role of lncRNAs in human evolution in this manuscript. They apply a suite of population genetics and functional genomics analyses that leverage existing data sets and public tools, some of which were previously built by the authors, who clearly have experience with lncRNA binding prediction. However, I worry that there is a lack of suitable methods and/or relevant controls at many points and that the interpretation is too quick to infer selection. While I don't doubt that lncRNAs contribute to the evolution of modern humans, and certainly agree that this is a question worth asking, I think this paper would benefit from a more rigorous approach to tackling it.

I thank the authors for their revisions to the manuscript; however, I find that the bulk of my comments have not been addressed to my satisfaction. As such, I am afraid I cannot say much more than what I said last time, emphasising some of my concerns with regards to the robustness of some of the analyses presented. I appreciate the new data generated to address some questions, but think it could be better incorporated into the text - not in the discussion, but in the results.

We thank the reviewer for the careful reading and valuable comments. In this round of revision, we address the two main concerns: (1) there is a lack of suitable methods and/or relevant controls at many points, and (2) the interpretation is too quick to infer selection. Based on these comments, we have carefully revised all sections of the manuscript, including the Introduction, Results, Discussion, and Materials and Methods.

In addition, we have performed two new analyses. Based on the two analyses, we have added one figure and two sections to Results, two sections to Materials and Methods, one figure to Supplementary Notes, and two tables to Supplementary Tables. These results were obtained using new methods and provided more support to the main conclusion.

To be more responsible, we re-look into the comments made in the first round and respond to them further. The following are point-to-point responses to comments.

Since many of the details in the Responses-To-Comments are available in published papers and eLife publishes Responses-To-Comments, we do not greatly revise supplementary notes to avoid ostensibly repeating published materials.

“lack of suitable methods and/or relevant controls”.

We carefully chose the methods, thresholds, and controls in the study; now, we provide clearer descriptions and explanations.

(1) We have expanded the last paragraph in Introduction to briefly introduce the methods, thresholds, and controls.

(2) In many places in Results and Materials and Methods, revisions are made to describe and justify methods, thresholds, and controls.

(3) Some methods, thresholds, and controls have good consensus, such as FDR and genome-wide background, but others may not, such as the number of genes that greatly differ between humans and chimpanzees. Now, we describe our reasons for the latter situation. For example, we explain that “About 5% of genes have significant sequence differences in humans and chimpanzees, but more show expression differences due to regulatory sequences. We sorted target genes by their DBS affinity and, to be prudential, chose the top 2000 genes (DBS length>252 bp and binding affinity>151) and bottom 2000 genes (DBS length<60 bp but binding affinity>36) to conduct over-representation analysis”.

(4) We also carefully choose proper words to make descriptions more accurate.

Responses to the suggestion “new data generated could be better incorporated into the text”.

(1) We think that this sentence “The occurrence of HS lncRNAs and their DBSs may have three situations – (a) HS lncRNAs preceded their DBSs, (b) HS lncRNAs and their DBSs co-occurred, (c) HS lncRNAs succeeded their DBSs. Our results support the third situation and the rewiring hypothesis”, previously in Discussion, should be better in section 2.3. We have revised it and moved it into the second paragraph of section 2.3.

(2) Our two new analyses generated new data, and we describe them in Results.

(3) It is possible to move more materials from Supplementary Notes to the main text, but it is probably unnecessary because the main text currently has eight sub-sections, two tables, and four figures.

Responses to the comment “the interpretation is too quick to infer selection”.

(1) When using XP-CLR, iSAFE, Tajima's D, Fay-Wu's H, the fixation index (Fst), and linkage disequilibrium (LD) to detect selection signals, we used the widely adopted parameters and thresholds but did not mention this clearly in the original manuscript. Now, in the first sentence of the second paragraph of section 2.4, we add the phrase “with widely-used parameters and thresholds” (more details are available in section 4.7 and Supplementary Notes).

(2) It is not the first time we used these tests. Actually, we used these tests in two other studies (Tang et al. Uncovering the extensive trade-off between adaptive evolution and disease susceptibility. Cell Rep. 2022; Tang et al. PopTradeOff: A database for exploring population-specificity of adaptive evolution, disease susceptibility, and drug responsiveness. Comput Struct Biotechnol J. 2023). In this manuscript, section 2.5 and section 4.12 describe how we use these tests to detect signals and infer selection. We also cite the above two published papers from which the reader can obtain more details.

(3) Also, in section 2.4, we stress that “Signals in considerable DBSs were detected by multiple tests, indicating the reliability of the analysis”.

To further respond to the comments of “lack of suitable methods” and “this paper would benefit from a more rigorous approach to tackling it”, we have performed two new analyses. The results of the new analyses agree well with previous results and provide new support for the main conclusion. The result of section 2.5 is novel and interesting.

We write in Discussion “Two questions are how mouse-specific lncRNAs specifically rewire gene expression in mice and how human- and mouse-specific rewiring influences the cross-species transcriptional differences”. To investigate whether the rewiring of gene expression by HS lncRNA in humans is accidental in evolution, we have made further genomic and transcriptomic analyses (Lin et al. Intrinsically linked lineage-specificity of transposable elements and lncRNAs reshapes transcriptional regulation species- and tissue-specifically. doi: https://doi.org/10.1101/2024.03.04.583292). To verify the obtained conclusions, we analyzed the spermatogenesis data from multiple species and obtained supporting evidence (not published).

I note some specific points that I think would benefit from more rigorous approaches, and suggest possible ways forward for these.

Much of this work is focused on comparing DNA binding domains in human-unique long-noncoding RNAs and DNA binding sites across the promoters of genes in the human genome, and I think the authors can afford to be a bit more methodical/selective in their processing and filtering steps here. The article begins by searching for orthologues of human lncRNAs to arrive at a set of 66 human-specific lncRNAs, which are then characterised further through the rest of the manuscript. Line 99 describes a binding affinity metric used to separate strong DBS from weak DBS; the methods (line 432) describe this as being the product of the DBS or lncRNA length times the average Identity of the underlying TTSs. This multiplication, in fact, undoes the standardising value of averaging and introduces a clear relationship between the length of a region being tested and its overall score, which in turn is likely to bias all downstream inference, since a long lncRNA with poor average affinity can end up with a higher score than a short one with higher average affinity, and it's not quite clear to me what the biological interpretation of that should be. Why was this metric defined in this way?

(1) Using RNA:DNA base-pairing rules, other DBS prediction programs return just DBSs with lengths. Using RNA:DNA base-pairing rules and a variant of Smith-Waterman local alignment, LongTarget returns DBSs with lengths and identity values together with DBDs (local alignment makes DBDs and DBSs predicted simultaneously). Thus, instead of measuring lncRNA/DNA binding based on DBS length, we measure lncRNA/DNA binding based on both DBS length and DBD/DBS identity (simply called identity, which is the percentage of paired nucleotides in the RNA and DNA sequences). This allows us to define “binding affinity”. One may think that binding affinity is a more complex function of length and identity. But, according to in vitro studies (see the review Abu Almakarem et al. 2012 and citations therein, and see He et al. 2015 and citations therein), the strength of a triplex is determined by all paired nucleotides (i.e., triplet). Thus, binding affinity=length * identity is biologically reasonable.

(2) Further, different from predicting DBS upon individual base-pairing rules such as AT-G and CG-C, LongTarget integrates base-pairing rules into rulesets, each covering A, T, C, and G (see the two figures below, which are from He et al 2015). This makes every nucleotide in the RNA and DNA sequences comparable and allows the computation of identity.

(3) On whether LongTarget may predict unreasonably long DBSs. Three technical features of LongTarget make this highly unlikely (and more unlikely than other programs). The three features are (a) local alignment, (b) gap penalty, and (c) TT penalty (He et al. 2015).

(4) Some researchers may think that a higher identity threshold (e.g., 0.8 or even higher) makes the predicted DBSs more reliable. This is not true. To explore plausible identity values, we analyzed the distribution of Kcnq1ot1’s DBSs in the large Kcnq1 imprinting region (which contains many known imprinted genes). We found that a high threshold for identity (e.g., 0.8) will make DBSs in many known imprinted genes fail to be predicted. Upon our analysis of many lncRNAs and upon early in vitro experiments, plausible identity values range from 0.4 to 0.8.

(5) Is it necessary or advisable to define an identity threshold? Since identity values from 0.4 to 0.8 are plausible and identity is a property of a DBS but does not reflect the strength of the whole triplex, it is more reasonable to define a threshold for binding affinity to control predicted DBSs. As explained above, binding affinity = length*identity is a reasonable measure of the strength of a triplex. The default threshold is 60, and given an identity of 0.6 in many triplexes, a DBS with affinity=60 is about 100 bp. Compared with TF binding sites (TFBS), 100 bp is quite long. As we explain in the main text, “taking a DBS of 147 bp as an example, it is extremely unlikely to be generated by chance (p < 8.2e-19 to 1.5e-48)”.

(6) How to validate predicted DBSs? Validation faces these issues. (a) DBDs are predicted on the genome level, but target transcripts are expressed in different tissues and cells. So, no single transcriptomic dataset can validate all predicted DBSs of a lncRNA. No matter using what techniques and what cells, only a small portion of predicted DBSs can be experimentally captured (validated). (b) The resolution of current experimental techniques is limited; thus, experimentally identified DBSs (i.e., “peaks”) are much longer than computationally predicted DBSs. (c) Experimental results contain false positives and false negatives. So, validation (or performance evaluation) should also consider the ROC curves (Wen et al. 2022).

(7) As explained above, a long DBS may have a lower binding affinity than a short DBS. A biological interpretation is that the long DBS may accumulate mutations that decrease its binding ability gradually.

There is also a strong assumption that identified sites will always be bound (line 100), which I disagree is well-supported by additional evidence (lines 109-125). The authors show that predicted NEAT1 and MALAT1 DBS overlap experimentally validated sites for NEAT1, MALAT1, and MEG3, but this is not done systematically, or genome-wide, so it's hard to know if the examples shown are representative, or a best-case scenario.

(1) We did not make this assumption. Apparently, binding depends on multiple factors, including co-expression of genes and specific cellular context.

(2) On the second issue, “this is not done systematically, or genome-wide”. We did genome-wide but did not show all results (supplementary fig 2 shows three genomic regions, which are impressively good). In Wen et al. 2022, we describe the overall results.

It's also not quite clear how overlapping promoters or TSS are treated - are these collapsed into a single instance when calculating genome-wide significance? If, eg, a gene has five isoforms, and these differ in the 3' UTR but their promoter region contains a DBS, is this counted five times, or one? Since the interaction between the lncRNA and the DBS happens at the DNA level, it seems like not correcting for this uneven distribution of transcripts is likely to skew results, especially when testing against genome-wide distributions, eg in the results presented in sections 5 and 6. I do not think that comparing genes and transcripts putatively bound by the 40 HS lncRNAs to a random draw of 10,000 lncRNA/gene pairs drawn from the remaining ~13500 lncRNAs that are not HS is a fair comparison. Rather, it would be better to do many draws of 40 non-HS lncRNAs and determine an empirical null distribution that way, if possible actively controlling for the overall number of transcripts (also see the following point).

(1) We predicted DBSs in the promoter region of 179128 Ensembl-annotated transcripts and did not merge DBSs (there is no need to merge them). If multiple transcripts share the same TSS, they may share the same DBS, which is natural.

(2) If the DBSs of multiple transcripts of a gene overlap, the overlap does not raise a problem for lncRNA/DNA binding analysis in specific tissues because usually only one transcript is expressed in a tissue. Therefore, there is no such situation “If, e.g., a gene has five isoforms, and these differ in the 3' UTR but their promoter region contains a DBS, is this counted five times, or one?”

(3) It is unclear to us what “it seems like not correcting for this uneven distribution of transcripts is likely to skew results” means. Regarding testing against genome-wide distributions, statistically, it is beneficial to make many rounds of random draws genome-wide, but this will take a huge amount of time. Since more variables demand more rounds of drawing, to our knowledge, this is not widely practiced in large-scale transcriptomic data analyses.

(4) If the difference (result) is small thus calls for rigorous statistical testing, making many rounds of random draws genome-wide is necessary. In our results, “45% of these pairs show a significant expression correlation in specific tissues (Spearman's |rho| >0.3 and FDR <0.05). In contrast, when randomly sampling 10000 pairs of lncRNAs and protein-coding transcripts genome-wide, the percent of pairs showing this level of expression correlation (Spearman's |rho| >0.3 and FDR <0.05) is only 2.3%”.

Thresholds for statistical testing are not consistent, or always well justified. For instance, in line 142 GO testing is performed on the top 2000 genes (according to different rankings), but there's no description of the background regions used as controls anywhere, or of why 2000 genes were chosen as a good number to test? Why not 1000, or 500? Are the results overall robust to these (and other) thresholds? Then line 190 the threshold for downstream testing is now the top 20% of genes, etc. I am not opposed to different thresholds in principle, but they should be justified.

(1) We used the g:Profiler program to perform over-representation analysis to identify enriched GO terms. This analysis is used to determine what pre-defined gene sets (GO terms) are more present (over-represented) in a list of “interesting” genes than what would be expected by chance. Specifically, this analysis is often used to examine whether the majority of genes in a pre-defined gene set fall in the extremes of a list: the top and bottom of the list, for example, may correspond to the largest differences in expression between the two cell types. g:Profiler always takes the whole genome as the reference; that is why we did not mention the whole genome reference. We now add in section 2.2 “(with the whole genome as the reference)”.

(2) Why choosing 2000 but not 2500 genes is somewhat subjective. We now explain that “About 5% of genes have significant sequence differences in humans and chimpanzees, but more show expression differences due to regulatory sequences. We sorted target genes by their DBS affinity and, to be prudential, chose the top 2000 genes (DBS length>252 bp and binding affinity>151) and bottom 2000 genes (DBS length<60 bp but binding affinity>36) to conduct over-representation analysis”.

Likewise, comparing Tajima's D values near promoters to genome-wide values is unfair, because promoters are known to be under strong evolutionary constraints relative to background regions; as such it is not surprising that the results of this comparison are significant. A fairer comparison would attempt to better match controls (eg to promoters without HS lncRNA DBS, which I realise may be nearly impossible), or generate empirical p-values via permutation or simulation.

We used these tests to detect selection signals in DBSs but not in the whole promoter regions. Using promoters without HS lncRNA DBS as the control also has risks because promoter regions contain other kinds of regulatory sequences.

There are huge differences in the comparisons between the Vindija and Altai Neanderthal genomes that to me suggest some sort of technical bias or the such is at play here. e.g. line 190 reports 1256 genes to have a high distance between the Altai Neanderthal and modern humans, but only 134 Vindija genes reach the same threshold of 0.034. The temporal separation between the two specimens does not seem sufficient to explain this difference, nor the difference between the Altai Denisovan and Neanderthal results (2514 genes for Denisovan), which makes me wonder if it is a technical artefact relating to the quality of the genome builds? It would be worth checking.

We feel it is hard to know whether or not the temporal separation between these specimens is sufficient to explain the differences because many details of archaic humans and their genomes remain unknown and because mechanisms determining genotype-phenotype relationships remain poorly known. After 0.034 was determined, these numbers of genes were determined accordingly. We chose parameters and thresholds that best suit the most important requirements, but these parameters and thresholds may not best suit other requirements; this is a problem for all large-scale studies.

Inferring evolution: There are some points of the manuscript where the authors are quick to infer positive selection. I would caution that GTEx contains a lot of different brain tissues, thus finding a brain eQTL is a lot easier than finding a liver eQTL, just because there are more opportunities for it. Likewise, claims in the text and in Tables 1 and 2 about the evolutionary pressures underlying specific genes should be more carefully stated. The same is true when the authors observe high Fst between groups (line 515), which is only one possible cause of high Fst - population differentiation and drift are just as capable of giving rise to it, especially at small sample sizes.

(1) We add in Discussion that “Finally, not all detected signals reliably indicate positive selection”.

(2) Our results are that more signals are detected in CEU and CHB than in YRI; this agrees all population genetics studies and implies that our results are not wrongly biased because more samples and larger samples were obtained from CEU and CHB.

Version 3

Abstract

Abstract

What genes and regulatory sequences make humans and apes that share substantial genes but show distinct phenotypes has puzzled researchers for decades. Genomic studies have examined species-specific genes and regulatory sequences (e.g., transcription factor binding sites, TFBS); birth, loss, and changes in these genes and sequences can greatly drive speciation and evolution. However, those involved in epigenetic regulation - species-specific lncRNA genes and regulatory sequences - remain poorly explored. We identified human-specific (HS) lncRNAs from GENCODE-annotated human lncRNAs, predicted their DNA binding domains (DBDs) and binding sites (DBSs) genome-wide, analyzed DBSs in modern humans (CEU, CHB, and YRI), archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and their counterparts in chimpanzees, and analyzed how DBSs influence gene expression in modern and archaic humans. Our results suggest that HS lncRNAs and their DBSs have substantially rewired gene expression human-specifically, the rewiring has evolved continuously from archaic to modern humans, and rewired gene expression has promoted brain development, made humans adapt to new environments and lifestyles, and caused differences in modern humans.

Introduction

Genomic changes that drive evolution include gene birth (Kaessmann, 2010), gene loss (Albalat and Canestro, 2016), and changes in regulatory sequences (Prud’homme et al., 2007). The limited genomic differences but the substantial phenotypic and behavioral divergence between humans and other hominids make what genomic changes have critically driven human evolution an enduring puzzle (Anton et al., 2014; Pollen et al., 2023). Waves of studies have identified genes promoting human brain development (Evans et al., 2005; Fiddes et al., 2018; Mekel-Bobrov et al., 2005; Pinson et al., 2022; Suzuki et al., 2018), but conclusions from gene birth-centric studies are controversial (Currat et al., 2006; Timpson et al., 2007; Yu et al., 2007). Moreover, genes important for brain development (e.g., NOTCH2NL and ASPM) may not regulate other human traits that coincide with or even precede brain enlargement. Studies revealed regulatory sequence changes important for human evolution, including new sequences (Liu et al., 2021), lost sequences (McLean et al., 2011), human accelerated regions (HARs) (Dong et al., 2016; Mangan et al., 2022; Prabhakar et al., 2006; Whalen and Pollard, 2022), and the turnover of transcription factor binding sites (TFBS) (Krieger et al., 2022; Otto et al., 2009; Zhang et al., 2023). However, how epigenetic regulatory sequences contribute to human evolution remains barely explored.

Experiments have generated substantial findings on epigenetic regulation. (a) About one-third of human lncRNA are primate-specific (Derrien et al., 2012), and the chimpanzee and human genomes are almost 99% identical (Chimpanzee Sequencing and Analysis Consortium, 2005). (b) Species-specific lncRNAs have distinct expression profiles in organs of different species (Sarropoulos et al., 2019). (c) Many lncRNAs can bind to DNA sequences by forming RNA:DNA triplexes (Abu Almakarem et al., 2012), and recruit histone and DNA modification enzymes to binding sites (Lee, 2009; Roadmap Epigenomics et al., 2015). (d) Approximately 40% of differentially expressed human genes result from interspecies epigenetic differences (Hernando-Herraez et al., 2015). Thus, besides human-specific (HS) protein-coding genes and related regulatory sequences (e.g., TFBS), HS lncRNAs and their DNA binding sites may also critically promote human evolution and determine human traits by human-specifically rewiring gene expression.

Since RNA sequence search by structure and sequence alignment can reliably identify orthologous lncRNA genes in genomes (Nawrocki, 2014; Nawrocki and Eddy, 2013), HS lncRNAs can be identified upon human lncRNAs’ orthologues in other mammals. Many lncRNAs contain DNA binding domains (DBDs) and can use DBDs to bind to DNA binding sites (DBSs) in the genome. Since RNA:DNA triplexes follow specific base-pairing rules (Abu Almakarem et al., 2012), DBDs and DBSs can be computationally predicted (Lin et al., 2019; Wen et al., 2022). Detected gene expression in organs and tissues, especially in human organs and tissues reported by the GTEx project (GTEx Consortium, 2017), provides data for examining whether the expression of lncRNAs and their potential targets (genes with DBSs) are correlated. Further, genomes of modern humans, archaic humans, and multiple apes have been sequenced (1000 Genomes Project Consortium et al., 2012; Chimpanzee Sequencing and Analysis Consortium, 2005; Meyer et al., 2012; Prufer et al., 2017; Prüfer et al., 2013), making the above analyses performed across species. Thus, it is time to examine whether HS lncRNAs and their DBSs have substantially and human-specifically rewired gene expression and determined human traits.

This study identified HS lncRNAs upon the orthologues of the GENCODE-annotated human lncRNA genes in 16 mammalian genomes, predicted HS lncRNAs’ DBDs and DBSs in the genomes of modern humans (CEU, CHB, and YRI), archaic humans (Altai Neanderthals, Denisovans, Vindija Neanderthals), and these DBSs’ counterparts in the genome of chimpanzees (Figure 1A). We experimentally validated DBS prediction using CRISPR/Cas9 experiment and multiple cell lines. Cross-population and cross-species analyses allowed us to exploit reasonable methods, controls, gene numbers, and data sources. To analyze DBSs accurately, which have different lengths, unclear biochemical properties, and significant variations across populations and species, we computed the binding affinity for each DBS upon its length and identity (percent of paired nucleotides) and classified DBSs into strong and weak ones. We also classified DBSs into old and young ones upon when most of their sequence changes occurred. These strategies ensure robust genomic analyses, which include evolutionary/population genetics analyses, and transcriptomic analyses, which cover multiple datasets, including the GTEx data and data of multiple experiments. Our analyses reveal that HS lncRNAs and their DBSs have distinctly and continuously rewired gene expression during human evolution, and the rewired gene expression determines or greatly influences diverse human traits from a rapidly evolved brain to a modified metabolism system.

Study Overview.

(A) The relationships between the chimpanzees, the three archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and the three modern human populations. Dashed lines indicate the phylogenetic distances from modern humans to archaic humans and chimpanzees. Genes and DBSs have the following implications – the DBS in B2M lacks a counterpart in chimpanzees; the DBS in ABL2 has great differences between archaic and modern humans; the DBS in IRNAR1 is highly polymorphic in modern humans. Red letters in DBSs indicate tissue-specific eQTLs or population-specific favored mutations. (B) DBSs’ mean length and binding affinity. (C) The target genes and transcripts of 66 HS lncRNAs. (D) The targeting relationships between HS lncRNAs. (E) The sequence distances of DBSs (the top 40%) from modern humans to chimpanzees and archaic humans. (F) The impacts of interactions between HS lncRNAs and DBSs on gene expression in GTEx tissues.

Results

HS lncRNAs regulate diverse genes and transcripts

How many human lncRNA genes are human-specific is an important question still without a clear answer. After searching for orthologues of the 13562 GENCODE-annotated human lncRNA genes (v18) in 16 mammalian genomes and in the genomes of archaic humans using the Infernal program that performs RNA sequence and structure alignment (Figure 1A) (Lin et al., 2019; Nawrocki, 2014; Nawrocki et al., 2009), we identified 66 HS lncRNA genes (Supplementary Table 1). Then, we predicted DBSs of HS lncRNAs in the 5000 bp promoter regions of the 179128 Ensembl-annotated transcripts (release 79) using the LongTarget program (Supplementary Note 1) (Lin et al., 2019). The binding affinity of a pair of DBD and DBS depends on their length and identity (i.e., the percent of paired nucleotides). Analyzing DBSs at known imprinted genes predicted using different LongTarget parameters and comparing DBSs with CHART-seq-reported binding sequences (“peaks”) reveal the reasonable affinity threshold of 60 (He et al., 2015) (Supplementary Fig. 2). Since weak DBSs could be generated recently and play subtle functions, we used affinity>=60 and 36<affinity<60 as the thresholds for identifying strong and weak DBSs. 105141 strong DBSs (mean length>147 bp) were identified in 96789 transcripts of 23396 genes; 152836 weak DBSs were identified in 127898 transcripts of 33185 genes (Figure 1B; Supplementary Table 2, 3). These genes and transcripts are HS lncRNAs’ potential targets. Several HS lncRNAs (e.g., RP11-423H2.3) have abundant targets; many targets have DBSs for multiple HS lncRNAs; many HS lncRNAs have multiple DBSs in a single target; and about 1.5% of target genes (0.6% of target transcripts) are human-specific (Figure 1C). Further, HS lncRNAs themselves show complex targeting relationships (Figure 1D; Supplementary Note 2). These results indicate that HS lncRNAs have distinctly rewired the expression of substantial conserved genes instead of regulating evolutionarily new genes.

We validated DBS prediction using multiple methods and datasets (Supplementary Note 1; Supplementary Fig 1-5). (a) HS lncRNA DBSs overlap well with experimentally identified DNA methylation and histone modification signals in multiple cell lines (https://genome.UCSC.edu). (b) Many DBSs co-localize with annotated cis-regulatory elements (cCREs) in genes’ promoter regions (https://genome.UCSC.edu). (c) Predicted DBSs overlap well with experimentally detected DBSs of NEAT1, MALAT1, and MEG3 (Mondal et al., 2015; West et al., 2014). (d) We used CRISPR/Cas9 to knock out predicted DBDs (100-200 bp) of seven lncRNAs (three HS lncRNAs and four wrongly transcribed lncRNAs) in multiple cell lines, performed RNA-seq before and after DBD knockout, analyzed differential gene expression, and found that the |fold change| of target genes was significantly larger than the |fold change| of non-target genes (one-sided Mann-Whitney test, p = 3.1e-72, 1.49e-114, 1.12e-206, 2.58e-09, 6.49e-41, 0.034, and 5.23e-07, respectively). This result suggests that the changed expression of target genes is caused by DBD knockout. (e) The LongTarget program uses the Smith-Waterman local alignment to identify DBD/DBS binding; thus, taking a DBS of 147 bp as an example, it is extremely unlikely to be generated by chance (p < 8.2e-19 to 1.5e-48).

HARs are conserved genomic loci yet show accelerated evolution in the human lineage. Researchers assume that HARs critically influence human evolution, but results obtained from HAR studies are controversial (Levchenko et al., 2018). The Zoonomia Project recently identified 312 HARs important for 3D genome rewiring and neurodevelopment (Keough et al., 2023). Since only 8 HARs overlap 26 DBSs of 14 HS lncRNAs, we infer that DBS and HAR are functionally different sequences.

HS lncRNAs’ target genes characterize human evolution and traits

Since strong and weak DBSs may occur and function in different periods in human evolution, it is interesting and necessary to examine whether their target genes are enriched in distinct functions. About 5% of genes have significant sequence differences in humans and chimpanzees, but more show expression differences due to regulatory sequences. We sorted target genes by their DBS affinity and, to be prudential, chose the top 2000 genes (DBS length>252 bp and binding affinity>151) and bottom 2000 genes (DBS length#x003C;60 bp but binding affinity>36) to conduct over-representation analysis using the g:Profiler program and Gene Ontology (GO) database to the two gene sets (Supplementary Table 4). The top and bottom 2000 genes (with the whole genome as the reference) are co-enriched in GO terms important for human evolution, including "neuron projection development" (GO:0031175), "learning" (GO:0007612), "regulation of anatomical structure size" (GO: GO:0090066), "head development" (GO:0060322), and "regulation of locomotion" (GO:0040012) (Supplementary Note 3; Supplementary Table 4). The genes containing strongest DBSs (514 DBSs have length >500 bp) include IFNAR1 and NFATC1 (important for immune function), KIF21B and NTSR1 (critical for neural development), SLC2A11 and SLC2A1 (involved in glucose usage), BAIAP3 (a brain-specific angiogenesis inhibitor) and TAS1R3 (a receptor for the sweet taste response), and some primate-specific noncoding RNA genes (e.g., CTD-3224I3.3 which is highly expressed in the cerebellum, lung, and testis) (Table 1; Supplementary Table 4, 5). On the other hand, genes with weak DBSs are enriched in such GO terms as “response to alcohol” (GO:0097305), “response to temperature stimulus” (GO:0009266), “DNA methylation or demethylation” (GO:0044728), and “female pregnancy” (GO:0007565) (Supplementary Note 3). These findings justify the classification of DBSs into strong and weak ones and indicate that strong and weak DBSs have different functions.

The rewiring of gene expression has greatly evolved

The human genome is almost 99%, 99%, and 98% identical to the genomes of chimpanzees, bonobos, and gorillas (Chimpanzee Sequencing and Analysis Consortium, 2005; Prufer et al., 2012), but what sequence variants critically differentiate these species remains unknown. We assessed whether HS lncRNAs’ DBSs are conserved in these species or human-specific. Notably, 97.81% of the 105141 strong DBSs have counterparts in chimpanzees, suggesting that these DBSs are similar to HARs in evolution and have undergone human-specific evolution.

HS lncRNAs may be generated before, with, or after their DBSs. To analyze the evolution of HS lncRNAs and their DBSs, we identified their counterparts in not only chimpanzees but also Altai Neanderthals, Denisovans, and Vindija Neanderthals (Meyer et al., 2012; Prufer et al., 2017; Prüfer et al., 2013). Counterparts of HS lncRNAs and strong DBSs were identified in these archaic humans, but the distances (distance per base) of HS lncRNAs are smaller than the distances of DBSs between archaic and modern humans, suggesting that many DBSs are older than HS lncRNAs and HS lncRNAs were generated after most of their DBSs. This suggests that HS lncRNAs make conversed genes in apes expressed specifically in humans.

We computed DBS distances in two ways to examine DBS evolution. The first is from the reconstructed human ancestor (downloaded from the EBI website) to chimpanzees, archaic humans, and modern humans; the second is from chimpanzees and archaic humans to modern humans. The first set of distances has the anomaly that many distances from the human ancestor to modern humans are smaller than to archaic humans. For example, DBSs in the top 20% genes in the human ancestor have distances >0.015 to modern humans, but more DBSs have distances >0.015 to Altai Neanderthals, Denisovans, and even chimpanzees (Supplementary Note 3). This anomaly is caused by the human ancestor being built using six primates without archaic humans. The second set of distances agrees well with the phylogenetic distances between chimpanzees, archaic humans, and modern humans. Large distances are in a small portion of DBSs (Figure 1E), thus targets containing these DBSs may greatly characterize human evolution. Based on a recent study that identified 5,984 genes differentially expressed between human-only and chimpanzee-only iPSC lines (Song et al., 2021), we estimated that the top 20% (4248) genes in chimpanzees may well characterize the human-chimpanzee differences, and we found that these genes have DBS distances >0.034. Taking 0.034 as the threshold, we identified 1256, 2514, and 134 genes in Altai Neanderthals, Denisovans, and Vindija Neanderthals with DBS distance >0.034. These numbers agree with the phylogenetic relationships between these archaic and modern humans (Figure 1A; Supplementary Table 6).

To reveal whether target genes with mostly changed DBSs are enriched in specific functions, we applied over-representation analysis to target genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals using the g:Profiler program and Gene Ontology (GO) database. To make statistical significance comparable, we used just the top 1256 genes as the inputs. The 1256 genes in chimpanzees and Altai Neanderthals are enriched in many common GO terms. However, “behavior” (GO:0007610) and "cognition" (GO:0050890) have much higher significance in Altai Neanderthals than in chimpanzees (Figure 2), and some GO terms such as "learning" (GO:0007612), “learning or memory” (GO:0007611), and “glucose homeostasis” (GO:0042593) is enriched only in Altai Neanderthals. In addition, agreeing with the closer relationship between Vindija Neanderthals and modern humans, the 134 genes in Vindija Neanderthals are enriched only in two GO terms with low significance (“smoothened signaling pathway” (GO:0007224) and “regulation of smoothened signaling pathway” (GO:0008589)) (Supplementary Table 7). These results reveal when and how HS lncRNA-mediated epigenetic regulation influences human evolution.

1256 target genes whose DBSs have the largest distances from modern humans to chimpanzees and Altai Neanderthals are enriched in different Biological Processes GO terms.

Upon significance threshold = 0.05 (Benjamini-Hochberg FDR), the two gene sets in chimpanzees and Altai Neanderthals are enriched in 199 and 409 GO terms (50<terms size<1000), respectively. Shared GO terms are color-marked. Left: Top GO terms enriched in genes in chimpanzees. Right: Top GO terms enriched in genes in Altai Neanderthals.

Epigenetic regulation by HS lncRNAs shows differences in modern humans

To check whether the rewiring has been population-specifically revised in modern humans, we analyzed single nucleotide polymorphisms (SNPs, with minimal allele frequency (MAF)≥0.1) in strong DBSs using the 1000 Human Genome data (1000 Genomes Project Consortium et al., 2012). We computed the average number of SNPs for all DBSs, identified highly polymorphic DBSs (SNP numbers in CEU, CHB, YRI are 5 times larger than the average), and identified DBSs with mostly changed sequence distances from modern humans to chimpanzees, Altai Neanderthals, and Denisovans (sequence distance>0.034 to chimpanzees, Altai Neanderthals, and Denisovans which are indicated by C, A, D in Table 1,2). The intersection of the two classes of DBSs (highly polymorphic DBSs and mostly changed DBSs), wherein SNPs may greatly change binding affinity, are found in genes important for human evolution (Supplementary Table 8), including IFNAR1, NFATC1, TAS1R3, INS, ST3GAL4, and FN3KRP (Table 2). Genes involved in sugar metabolism are especially notable. A key feature of modern human diets is high sugar intake which causes non-enzymatic oxidation of proteins (i.e., glycation) and deactivates protein functions. Among proteins encoded by genes that have these two classes of DBSs, TAS1R3 recognizes diverse natural and synthetic sweeteners, Insulin decreases blood glucose concentration, ST3GAL4 regulates protein glycosylation, and FN3KRP deglycates proteins. These DBSs indicate continuous evolution of gene regulation human-specifically and population-specifically by HS lncRNAs.

We next used multiple statistical tests, including XP-CLR (Chen et al., 2010), iSAFE (Akbari et al., 2018), Tajima’s D (Tajima, 1989), Fay-Wu’s H (Fay and Wu, 2000), the fixation index (Fst) (Weir and Cockerham, 1984), and linkage disequilibrium (LD) (Slatkin, 2008), with widely-used parameters and thresholds (Supplementary Note 4, 5), to detect selection signals in HS lncRNAs and strong DBSs in CEU, CHB, and YRI which are the representative populations of Europeans, Asians, and Africans. These tests detected selection signals in several HS lncRNA genes in CEU and CHB but not in YRI and selection signals in more DBSs in CEU and CHB than in YRI. Selection signals in YRI may be underestimated due to fewer samples and smaller sample sizes (than CEU and CHB), yet the overall results agree well with features of human evolution. Signals in considerable DBSs were detected by multiple tests, indicating the reliability of the analysis (Supplementary Note 4, 5; Supplementary Table 9, 10, 11). DBSs in the two pigmentation-related genes MC1R and MFSD12 and the two odor reception-related genes OR6C1 and TAS1R3 are notable examples. The rich SNPs, population-specific selection signals, and specific functions of related genes together indicate the ongoing influence of rewired gene expression on human evolution.

Young weak DBSs may have greatly promoted recent human evolution

To further examine whether the rewiring has been revised recently, we classified strong and weak DBSs with significant sequence changes into “old” ones, “young” ones, and others. We defined old DBSs as having “Human-Chimp distance>0.034 AND Human-Altai Neanderthals distance==0” and young DBSs as having “Human-Altai Neanderthals distance>0.034 OR Human-Denisovan distance>0.034”, which demark early and late sequence changes in human evolution. We identified favored and hitchhiking mutations in 17 human populations (Tang et al., 2022; Tang et al., 2023), classified DBSs into six classes (strong old, strong young, strong other, weak old, weak young, weak others), and examined the numbers of favored and hitchhiking mutations in each class (Supplementary Table 16). Favored and hitchhiking mutations are most enriched in young weak DBSs, suggesting that these DBSs may have greatly promoted recent human evolution.

SNPs in DBSs have cis-effects on gene expression

To verify the above genomic and population genetics analysis, we examined SNPs in DBSs and their impact on gene expression using transcriptomic data. Based on the Genotype-Tissue Expression (GTEx) data (GTEx Consortium, 2017), we identified expression quantitative trait loci (eQTLs) in DBSs using the widely adopted criteria of MAF≥0.1 and cis-effect size (ES)≥0.5. 1727 SNPs, with MAF ≥0.1 in at least one population and absolute values of ES≥0.5 in at least one tissue, were identified in DBSs in autosomal genes (Supplementary Table 12,13). These eQTLs include 372 “conserved” ones (i.e., also in DBSs in the three archaic humans) and 1020 “novel” ones (i.e., only in DBSs in modern humans). Many conserved eQTLs are in brain tissues and have high derived allele frequencies (DAFs) in all three modern populations. A notable eQTL that has high DAFs in all three populations and has positive ES≥0.5 in 44 out of 48 GTEx tissues is rs2246577 in the DBS in FN3KRP, whose encoded protein deglycates proteins to restore their function (Table 2). In contrast, many novel eQTLs are in a particular tissue and have population-specific DAFs (Supplementary Note 6).

Next, we performed two analyses to examine how eQTLs in DBSs influence gene expression via lncRNA-DBS interactions. First, we computed the expression correlation between HS lncRNAs and target transcripts with DBSs having eQTLs in specific tissues. Most (94%) HS lncRNA-target transcript pairs showed significant expression correlation in tissues in which the eQTLs were identified (|Spearman’s rho| >0.3 and FDR <0.05). Second, we examined whether eQTLs are more enriched in DBSs than in promoter regions by computing and comparing the density of eQTLs in each DBS and each Ensembl-annotated promoter. Using promoters as the control for the comparison is reasonable as they contain and also are longer than DBSs. The results indicate that eQTLs are more enriched in DBSs than in promoters (one-sided Mann-Whitney test, p = 0.0) and that these eQTLs influence gene expression via lncRNA-DBS interactions (Supplementary Note 6). Thus, tissue-specific eQTLs and population-specific SNPs in DBSs support the genomic and population genetics analysis.

HS lncRNAs promote brain evolution from archaic to modern humans

To further confirm that SNPs in DBSs have cis-effects on gene expression, we examined the impact of HS lncRNA-DBS on gene expression in the GTEx tissues (GTEx Consortium, 2017). 40 autosomal HS lncRNAs are expressed in at least one tissue (median TPM>0.1) and form 198876 HS lncRNA-DBS pairs with target transcripts in all tissues. Of these pairs, 45% show a significant expression correlation in specific tissues (Spearman’s |rho|>0.3 and FDR<0.05). To check how likely such correlation could be obtained by chance, we randomly sampled 10000 pairs of lncRNAs and protein-coding transcripts genome-wide and found that only 2.3% of pairs show significant expression correlation (Spearman’s |rho|>0.3 and FDR<0.05). Moreover, a higher percentage (56%) of HS lncRNA-target transcript pairs with significant correlation is detected in at least one brain region, indicating more extensive gene expression regulation by HS lncRNAs in the brain (Figure 3A).

The impact of HS lncRNA-DBS interaction on gene expression in GTEx tissues and organs.

(A) The percentage distribution of correlated HS lncRNA-target transcript pairs across GTEx tissues and organs. Higher percentages of correlated pairs are in the brain than in other tissues and organs. (B) The distribution of significantly changed DBSs in HS lncRNA-target transcript pairs across GTEx tissues and organs between archaic and modern humans. Orange, red, and dark red indicate significant DBS changes from Denisovans (D), Altai Neanderthals and Denisovans (AD), and all of the three archaic humans (ADV). DBSs in HS lncRNA-target transcript pairs correlated in seven brain regions (in dark red) have changed significantly and consistently since the Altai Neanderthals, Denisovans, and Vindija Neanderthals (one-sided two-sample Kolmogorov-Smirnov test, significant changes determined by FDR <0.001).

To obtain detailed supporting evidence, we analyzed nine experimental datasets (Supplementary Note 7). First, we analyzed two datasets of epigenetic studies, one examining H3K27ac and H3K4me2 profiling of human, macaque, and mouse corticogenesis, and the other examining gene expression and H3K27ac modification in eight brain regions in humans and four other primates (Reilly et al., 2015; Xu et al., 2018). Compared with the genome-wide background, 84% and 73% of genes in the two datasets have DBSs for HS lncRNAs, indicating significantly higher enrichment in HS lncRNA targets (p = 1.21e-21 and 1.2e-56, two-sided Fisher’s exact test). When taking gene expression in chimpanzees as the control, 1851 genes show human-specific transcriptomic differences in one or multiple brain regions, but only 240 genes show chimpanzee-specific transcriptomic differences. Second, we analyzed two datasets of PsychENCODE studies, one examining spatiotemporally differentially expressed genes and spatiotemporally differentially methylated sites in 16 brain regions, and the other examining the spatiotemporal transcriptomic divergence across human and macaque brain development (Li et al., 2018; Zhu et al., 2018). In the first dataset, 65 HS lncRNAs are expressed in all 16 brain regions, 109 transcripts have spatiotemporally differentially methylated sites in their promoters, and 56 of the 109 transcripts have DBSs for HS lncRNAs. In the second dataset, 8951 genes show differential expression between the human and macaque brains, and 72% of differentially expressed genes in the human brain have DBSs of HS lncRNAs. Thus, both datasets show significant enrichment for HS lncRNA regulation. Third, three studies identified genes critically regulating cortical expansion (Florio et al., 2018; Johnson et al., 2018; Suzuki et al., 2018). Of the 40 reported protein-coding genes, 29 have DBSs of HS lncRNAs in promoter regions (Supplementary Table 14). Thus, these genes are enriched with DBSs of HS lncRNAs compared with the genome-wide background (p<0.01, two-sided Fisher’s exact test). Finally, we analyzed two datasets of brain organoids studies. By establishing and comparing cerebral organoids between humans, chimpanzees, and macaques, Pollen et al. identified 261 human-specific gene expression changes (Pollen et al., 2019). By fusing human and chimpanzee iPS cells and differentiating the hybrid iPS cells into hybrid cortical spheroids, Agoglia et al. generated a panel of tetraploid human-chimpanzee hybrid iPS cells and identified thousands of genes with divergent expression between humans and chimpanzees (Agoglia et al., 2021). We found that the 261 and 1102 genes in the two datasets are enriched with DBSs of HS lncRNAs compared with the genome-wide background (p=1.2e-16 and 3.4e-74).

Finally, we next examined the evolution of the impact of HS lncRNAs and their DBSs on gene expression using the GTEx data. For each DBS in a pair of HS lncRNA-target transcript that show correlated expression in a GTEx tissue, we computed the sequence distances of this DBS from modern humans to archaic humans. Then, we compared the distribution of DBS sequence distances in each tissue with the distribution of DBS sequence distances in all tissues (the control) (one-sided two-sample Kolmogorov-Smirnov test). DBSs in HS lncRNA-target transcript pairs correlated in brain regions have significantly changed sequence distances since the Altai Neanderthals and Denisovans (Figure 3B), suggesting that gene expression regulation by HS lncRNAs in the brain has experienced significant evolution and that HS lncRNAs contributed greatly to recent brain evolution (Ponce de Leon et al., 2021). Results from other GTEx tissues, including Kidney-Cortex, Ovary, and Skin-Sun Expressed (Figure 3B), also suggest continuous influences of rewired gene expression on human evolution.

HS lncRNAs mediate human-specific correlated gene expression in the brain

Species-specificity of expression correlation highlights species-specific transcriptional regulation. To further explore the evidence in this aspect, we used the GTEx data from the human frontal cortex (BA9) and anterior cingulate cortex (BA24) (n=101 and 83, respectively) (GTEx Consortium, 2017), the transcriptomic data for the same brain regions in macaques (n=22 and 25, respectively) (Zhu et al., 2018), and the eGRAM program we developed recently, to identify co-expressed gene modules. HS lncRNAs’ target genes form distinct modules in the human brain, in which genes have rich correlation relationships (Pearson’s r>0.8) and are enriched in neurodevelopment-related KEGG pathways (hypergeometric distribution test, FDR<0.05). In contrast, their orthologues in the same macaque brain regions do not have these features (Figure 4; Supplementary Figure 24; Supplementary Table 17). The distinct differences in gene modules in these brain regions further support that gene expression has been substantially rewired by HS lncRNAs in the human brain.

Human-specifically rewired gene expression by HS lncRNAs in the frontal cortex (BA9).

(A) Genes expressed in the frontal cortex are enriched for HS lncRNAs’ target genes and neurodevelopment-related pathways. Squares indicate HS lncRNAs; dots indicate gene modules (Module_1 and Module_2 are illustrated); colors indicate enriched KEGG pathways. (B) Comparison of modules of the same genes in humans and macaques (indicated by H and M). In each pair of modules, green dots, and blue dots denote human genes and their orthologues in the corresponding positions, and lines indicate correlation. Correlated expression is more prominent in humans than in macaques.

Discussion

Modern humans differ from archaic humans and great apes trivially in genomes but greatly in phenotypes, making what genomic differences critically determine modern humans a profound question. This question has been addressed by many researchers along multiple approaches. Previous attention has been paid to new protein-coding genes, HARs, and TF-TFBS interactions, and findings include that HS protein-coding genes promote rapid neocortex expansion (Fiddes et al., 2018; Florio et al., 2018; Pinson et al., 2022), that HARs are significantly enriched in domains important for human-specific 3D genome organization (Keough et al., 2023), and that gene expression is rewired by TFs and TFBSs (Johnson, 2017). These findings make us postulate that HS lncRNAs have greatly rewired gene expression. To verify this postulation, this study identified HS lncRNAs, their DBSs, and these DBSs’ counterparts in ancient humans and multiple apes. Our analyses of these lncRNAs, DBSs, and target genes, including their evolution and interaction, indicate that HS lncRNAs have greatly promoted human evolution by distinctly rewiring gene expression. This conclusion is supported by our analyses of multiple experimental datasets. The rewiring has greatly evolved from humans’ ancestors to archaic humans and modern humans and still influences modern human populations.

As DBS prediction is the basis of the study, we performed CRISPR/Cas9 experiments to validate predicted DBSs. When running Infernal, LongTarget, MAFFT7, Phylip, and eGRAM, default parameters that suit most situations were used. When analyzing gene sets, the whole genome or the chimpanzee genome were used as controls. When identifying HS lncRNAs’ target genes that probably critically characterize human evolution, the top 20% of genes having the largest DBS distances from humans to chimpanzees were used because this percentage of genes may well characterize the human-chimpanzee differences. Accordingly, the DBS distance of 0.034 was used as the threshold to identify genes that may well characterize the differences between modern and archaic humans. The numbers of critical genes agree with the phylogenetic relationships between modern humans, archaic humans, and chimpanzees. Importantly, we differentiated strong/weak DBSs and old/young DBSs. We found that different classes of DBSs have different evolutionary and population genetics features and genes with these DBSs are enriched in distinct functions. Other values of parameters and thresholds may generate slightly different results, but the core of the results and the main conclusion probably still hold. Finally, not all detected signals reliably indicate positive selection.

Several findings are noteworthy. First, Neanderthal-inherited sequences have measurable impacts on gene expression in modern humans, and the impacts are least detectable in the brain (McCoy et al., 2017). In comparison, we find that HS lncRNAs, which may have evolved quite recently, have more significantly influenced gene expression in the brain than in other tissues. Second, the regulation of many genes by HS lncRNAs has undergone great evolution, notable examples including those making humans adapt to high sugar intake. Third, intensively regulated genes are enriched in different functions at different periods of human evolution. Specifically, “old” SNPs in DBSs are in genes regulating neural development, and “new” SNPs in DBSs are in genes regulating such processes as glucose metabolism. Fourth, recently emerged regulations (i.e., target genes with young weak DBSs) have significantly promoted recent human evolution. These findings reveal diverse, continuous, and subtle contributions of HS lncRNA to human evolution.

The above findings also raise new questions. First, the protein-coding regions in the mouse and human genomes are about 85% identical; however, many mouse lncRNAs are rodent-specific (Yue et al., 2014). Thus, two questions are how mouse-specific lncRNAs specifically rewire gene expression in mice and how human- and mouse-specific rewiring influences the cross-species transcriptional differences (Breschi et al., 2017; Hodge et al., 2019; Zhu et al., 2018). Second, whether mice and other mammals (especially model animals) have species-specific lncRNAs such as RP11-423H2.3 that substantially rewires gene expression species-specifically. Third, whether and how clade- and species-specific lncRNAs would make evolution preordained. These questions are theoretically interesting and important for translational medicine.

Material and methods

Data sources

The sequences of human (GRCh38/hg38, GRCh37/hg19) and chimpanzee genomes (panTro5) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). Three high-quality archaic human genomes were obtained from the Max Planck Institute for Evolutionary Anthropology (https://www.eva.mpg.de/genetics/index.html), which include an Altai Neanderthal which lived approximately 122 thousand years ago (kya) (Prüfer et al., 2013), a Denisovan (an Asian relative of Neanderthals) which lived approximately 72 kya (Meyer et al., 2012), and a Vindija Neanderthal which lived approximately 52 kya (Prufer et al., 2017). The ancestral sequences for the human genome (GRCh37) (which were generated upon the EPO alignments of six primate species human, chimpanzee, gorilla, orangutan, macaque, and marmoset) were downloaded from the EBI website (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/). The SNP data of modern humans were obtained from the 1000 Genomes Project (phase I) (1000 Genomes Project Consortium et al., 2012). The three modern human populations are CEU (Utah residents with Northern and Western European ancestry), CHB (Han Chinese in Beijing, China), and YRI (Yoruba in Ibadan, Nigeria), which contain 100, 99, and 97 individuals, respectively. The genetic map files of modern humans were downloaded from the 1000 Genomes Project website (ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/technical/working/20110106_recombination_hotspots/). The 1000 Genomes phased genotype data in VCF format were downloaded from the International Genome Sample Resource website (http://www.internationalgenome.org/). Annotated human genes and transcripts were obtained from the Ensembl website (http://www.Ensembl.org). FASTA-format nucleotide sequences were converted from VCFs using the vcf-consensus program in the VCFtools package when necessary (Danecek et al., 2011). Data in the GeneCards Human Gene Database (www.genecards.org) were used to annotate genes and transcripts.

RNA-sequencing (RNA-seq) data (transcripts per million [TPM]) and eQTL data of human tissues were obtained from the Genotype-Tissue Expression Project (GTEx, v8) website (https://gtexportal.org/) (GTEx Consortium, 2017). Histone modification and DNA methylation signals in multiple cell lines and ENCODE Candidate Cis-Regulatory Elements (cCRE) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). The genome-wide DNA binding sites of lncRNA NEAT1, MALAT1, and MEG3 were obtained from experimental studies (Mondal et al., 2015; West et al., 2014). The predicted and experimentally detected DBSs in target genes of MALAT1, NEAT1, and MEG3 are given in supplementary tables (Supplementary Table 15).

Identifying HS lncRNA genes

We used the Infernal program (Nawrocki et al., 2009), which searches orthologous RNA sequences upon sequence and structure alignment, to identify orthologous exons in 16 mammalian genomes for each exon in each of the 13562 GENCODE (v18)-annotated human lncRNA genes (Lin et al., 2019). The 16 mammals were chimpanzee (CSAC 2.1.4/panTro4), macaque (BGI CR_1.0/rheMac3), marmoset (WUGSC 3.2/calJac3), tarsier (Broad/tarSyr1), mouse lemur (Broad/micMur1), tree shrew (Broad/tupBel1), mouse (GRCm38/mm10), rat (Baylor3.4/rn4, RGSC6.0/rn6), guinea pig (Broad/cavPor3), rabbit (Broad/oryCun2), dog (Broad CanFam3.1/canFam3), cow (Baylor Btau_4.6.1/bosTau7), elephant(Broad/loxAfr3), hedgehog (EriEur2.0/eriEur2), opossum (Broad/monDom5), and platypus (WUGSC 5.0.1/ornAna1) (http://genome.UCSC.edu). If the number of orthologous exons of a human lncRNA gene in a genome exceeded half the exon number of the human lncRNA gene, these orthologous exons were assumed to form an orthologous lncRNA gene. If a human lncRNA gene had no orthologous gene in all of the 16 mammals, it was assumed to be human-specific.

Identifying DBSs of HS lncRNAs

LncRNAs bind to DNA sequences by forming RNA:DNA triplexes. Each triplex comprises triplex-forming oligonucleotides (TFO) in the lncRNA and a triplex-targeting site (TTS) in the DNA sequence. We used the LongTarget program to predict HS lncRNAs’ DNA binding domains (DBD) and binding sites (DBS) with the default parameters (Ruleset = all, T.T. penalty = -1000, CC penalty = 0, Offset = 15, Identity ≥ 60, Nt ≥ 50) (Lin et al., 2019; Wen et al., 2022). LongTarget program predicts DBDs and DBSs simultaneously, with a DBD comprising a set of densely overlapping TFOs and a DBS comprising a set of densely overlapping TTSs. For each HS lncRNA, we predicted its DBSs in the 5000 bp promoter regions (-3500 bp upstream and +1500 bp downstream the transcription start site) of the 179128 Ensembl-annotated transcripts (release 79). For each DBS, a binding affinity was calculated by multiplexing the DBS length and the averaged Identity score of all TTSs (the Identity score is the percentage of paired nucleotides). Strong and weak DBSs were classified upon binding affinity ≥ 60 and < 60. A transcript whose promoter region contains a strong DBS of an HS lncRNA was assumed to be a target transcript of the HS lncRNA, and the gene containing this transcript was assumed to be a target gene of the HS lncRNA. As the 1000 Genomes Project (phase I) data and the archaic human genomes are based on GRCh37/hg19, the coordinates of DBSs were converted from GRCh38/hg38 to GRCh37/hg19 using the liftover program from the UCSC Genome Browser (Kuhn et al., 2013).

Experimentally validating DBS prediction

A 157 bp sequence (chr17:80252565-80252721, hg19) containing the DBD of RP13-516M14.1, a 202 bp sequence (chr1:113392603-113392804, hg19) containing the DBD of RP11-426L16.8, and a 198 bp sequence (chr17:19460524-19460721, hg19) containing the DBD of SNORA59B, were knocked out in the HeLa cell line, RKO cell line, and SK-MES-1 cell line, respectively. Two sequences (chr1:156643524-156643684, chr10:52445649-52445740, hg38) containing the DBD of two wrongly transcribed noncoding sequences were knocked out in the HCT-116 and A549 cell lines, respectively. The seven knockouts were performed by UBIGENE, Guangzhou, China (http://www.ubigene.com) using CRISPR-UTM, a revised version of CRISPR/Cas9 technology. Before and after the seven DBD knockouts, RNA-sequencing (RNA-seq) was performed by Novogene, Beijing, China (https://cn.novogene.com) and HaploX, Shenzhen, China ( https://www.haplox.cn/). The reads were aligned to the human GRCh38 genome using the Hiasat2 program (Kim et al., 2019), and the resulting Sam files were converted to bam files using Samtools (Li et al., 2009). The Stringtie program was used to quantify gene expression levels (Pertea et al., 2015). Fold change of gene expression was computed using the edgeR package (Robinson et al., 2010), and significant up- and down-regulation of target genes after DBD knockout was determined upon |log2(fold change)| > 1 with FDR < 0.1.

Genome-wide DBSs of NEAT1, MALAT1, and MEG3 were experimentally detected (Mondal et al., 2015; West et al., 2014). We also used these data to validate DBS prediction by predicting DBSs of the three lncRNAs and checking the overlap between predicted and experimentally detected DBSs (Supplementary Table 15).

Mapping DBSs of HS lncRNAs in the chimpanzee and archaic human genomes

We used the liftover program from the UCSC Genome Browser (http://genome.UCSC.edu) to map loci of DBSs in the human genome hg38 to the chimpanzee genome Pan_tro 3.0 (panTro5). The mapping results were examined by checking the human-chimpanzee pairwise alignment (hg38 vs panTro5) in the UCSC Genome Browser (http://genome.UCSC.edu). 97.81% (100403/102651) of DBSs were mapped, and the 2248 unmapped DBSs in 429 genes were considered human-specific. We used vcf-consensus in the VCFtools package to extract the DBSs of HS lncRNAs from the VCF files of Altai Neanderthals, Denisovans, and Vindija Neanderthals. The variant with the highest quality score was selected whenever multiple variant calls were observed at a given locus. The obtained DBS sequences in the chimpanzees and three archaic humans are called counterparts of DBSs in these genomes.

Estimating sequence distances of DBSs between different genomes

We first aligned DBS sequences in the genomes of humans, chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals using the MAFFT7 program for measuring sequence distances from modern humans to chimpanzees and archaic humans (Katoh and Standley, 2013). We then computed sequence distances using the dnadist program with the Kimura 2-parameter model in the PHYLIP (3.6) package (http://evolution.genetics.washington.edu/phylip.html) and using the Tamura-Nei model in the MEGA7 package (Kumar et al., 2016). The two methods generated equivalent results. The largest distance between DBSs in humans and their counterparts in chimpanzees is 5.383. Since 2248 DBSs in 429 human genes do not have counterparts in chimpanzees, we assumed that the sequence distances of these DBSs between humans and chimpanzees were 10.0. Since the top 20% of DBSs in chimpanzees (including these 2248 DBSs) have distances ≥ 0.034, we used 0.034 as the threshold to identify the "mostly changed" DBSs in the chimpanzees and three archaic humans.

We determined human ancestral sequences of DBSs upon the human ancestor sequences from the EBI website, which were generated upon the EPO alignments of six primate species. We used the above-mentioned methods to compute DBSs’ sequence distances from the human ancestor to chimpanzees, archaic humans, and modern humans. We found that when the human-chimpanzee ancestral sequence has the ancestral sequence (which means the inferring of ancestral allele is of high confidence), DBS distances from the human ancestor to modern humans are larger than to archaic humans, but this situation only accounts for about 63.8%. For many DBSs, the distances from the human ancestor to modern humans are smaller than to archaic humans (especially Neanderthals and Denisovans) and even to chimpanzees. This defect may be caused by the absence of archaic humans in building the human ancestral sequence.

Detecting positive selection signals in HS lncRNA genes and DBSs

We used multiple tests to detect positive selection signals in HS lncRNA genes. First, we used the XP-CLR test (parameters = -w1 0.001 300 100 -p0 0.95, window size = 0.1 cM, grid size = 100 bp) to perform six pairwise genome-wide scans (i.e., CEU-CHB, CEU-YRI, CHB-CEU, CHB-YRI, YRI-CEU, and YRI-CHB) (Chen et al., 2010). The upper 1% of scores across the entire genome in each pairwise scan were 34.6 in the CEU-YRI scan, 16.8 in the CEU-CHB scan, 45.0 in the CHB-YRI scan, 26.9 in the CHB-CEU scan, 14.1 in the YRI-CEU scan, and 14.1 in the YRI-CHB scan. These scores were used as the thresholds of positive selection signals in these populations. Second, we used the iSAFE program to scan each genomic region that contains a HS lncRNA gene and its 500-kb upstream and downstream sequences (Akbari et al., 2018). Strongly selected loci were detected only in CEU and CHB. Third, we used the VCFtools program to calculate Tajima’s D values for each HS lncRNA gene in CEU, CHB, and YRI (Danecek et al., 2011). The calculation was performed using a 1500-bp non-overlapping sliding window because the lengths of these genes exceed 1500 bp. To generate the background reference for judging the significant increase or decrease of Tajima’s D for HS lncRNA genes in a population, we calculated Tajima’s D values across the whole genome using a sliding window of 1500 bp. As values of Tajima’s D were compared with the background reference, a significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively, instead of population demography dynamics (Tajima, 1989). Fourth, we used the integrated Fst to detect positive selection signals in HS lncRNA genes. The Fst of each HS lncRNA gene was computed for three comparisons, i.e., CEU-YRI, CHB-YRI, and CHB-CEU. Extreme Fst values of SNPs were detected in HS lncRNA genes in the comparisons of CEU-YRI and CHB-YRI. Since allele frequencies for different loci vary in a genome and genetic drift may have different effects at different loci, we used a sliding window of 1500 for comparing Fst values of HS lncRNA genes with the genome-wide background. Extreme Fst values indicate positive selection. Finally, we applied linkage disequilibrium (LD) analysis to each HS lncRNA gene. We computed the pairwise LD (r²) in CEU, CHB, and YRI for common SNPs (with minimal allele frequency (MAF) ≥ 0.05 in at least one population) in HS lncRNA genes and DBSs using the PLINK program (Purcell et al., 2007). Significantly increased LD was detected in SNPs in HS lncRNA genes in CEU and CHB. The LD patterns were portrayed using the Haploview program (Barrett et al., 2005).

Next, we used the above tests to detect positive selection signals in DBSs. First, the 100 bp grid size of the XP-CLR test also allowed the detection of selection signals in DBSs. Second, we performed the Tajima’s D and Fay-Wu’s H tests (Fay and Wu, 2000; Tajima, 1989). We calculated Tajima’s D values for each DBS in CEU, CHB, and YRI using the VCFtools program (Danecek et al., 2011), with a sliding window of 147 bp (the mean length of strong DBSs). To generate the background reference for judging the significant increase or decrease of Tajima’s D in a population, we calculated Tajima’s D values across the whole genome using the same sliding window. As values of Tajima’s D were compared with the background reference, a significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively. Fay-Wu’s H values were calculated similarly using the VariScan program (Vilella et al., 2005). Calculating Fay-Wu’s H demands the ancestral sequences as the outgroup. We extracted the ancestral sequences of DBSs from the human ancestral sequence, which was generated upon the EPO alignments of six primate species (human, chimpanzee, gorilla, orangutan, macaque, and marmoset) (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/). Third, we computed the Fst to measure the frequency differences of alleles in DBSs between populations. For the CEU-YRI, CHB-YRI, and CHB-CEU pairwise comparisons, we used the revised VCFtools program to compute the weighted Fst for all SNPs in each DBS (Weir and Cockerham, 1984). Fourth, we integrated the weighted Fst values in the three populations into an "integrated Fst" which indicated whether the DBS locus was under selection in a certain population (Nielsen and Slatkin, 2013). We used sliding windows of 147 bp for comparing Fst values of DBSs with the genome-wide background. We empirically denoted the values of integrated Fst in the upper 10% of scores across the entire genome as statistically significant. To determine positive selection more reliably, we used Tajima’s D and the integrated Fst to jointly determine if a DBS was under positive selection in a population. The thresholds that determined the upper 10% of Tajima’s D values across the entire genome in CEU, CHB, and YRI were -0.97, -0.96, and -0.97, respectively, and the threshold that determined the upper 10% of integrated Fst values across the entire genome in the three populations was 0.22. For example, a DBS was assumed to be under positive selection in CEU if (a) the DBS had a Tajima’s D<-0.97 in CEU and Tajima’s D>0.0 in the two other populations and (b) the DBS had an integrated Fst>0.22. Analogously, a DBS was assumed to be under positive selection in both CEU and CHB if the DBS had a Tajima’s D <-0.97 in CEU, <-0.96 in CHB, and >0.0 in YRI, and had an integrated Fst>0.22.

Functional enrichment analysis of genes

We used the g:Profiler program (with the parameter settings: Organism=Homo sapiens, Ordered query = No, Significance threshold = Benjamini-Hochberg FDR, User threshold = 0.05, 50 < terms size < 1000) and the Gene Ontology (GO) database to perform over-representation analysis (Raudvere et al., 2019). This analysis is used to determine what pre-defined gene sets (GO terms) are more present (over-represented) in a list of “interesting” genes than what would be expected by chance. The lists of genes included genes with strongest DBSs in humans, genes with weakest DBSs in humans, and genes with mostly changed DBSs in chimpanzees and three archaic humans. To choose genes with strong DBSs, we ranked genes in descending order based on binding affinity and chose the top 2000 genes. To choose genes with weak DBSs, we first ranked genes in ascending order based on binding affinity, then ranked genes in descending order based on the number of regulatory HS lncRNAs, and finally chose the top 2000 genes with binding affinity ≤40 and regulatory HS lncRNAs ≥5. We used the distance threshold of 0.034 to choose genes with mostly changed DBSs. Genes with mostly changed DBSs in chimpanzees include the 429 genes in which 2248 DBSs lack counterparts. 4248, 1256, 2514, and 134 genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals have DBSs with distance ≥0.034. To make the statistical significance comparable across chimpanzees, Altai Neanderthals, and Denisovans, we used the same number of genes (the top 1256 genes whose DBSs have the largest DBS distances) as the input to the g:Profiler program.

Analysing SNP frequencies in human populations

The frequency of common SNPs (MAF ≥0.05) in DBSs in the three modern human populations was computed using the VCFtools package (Danecek et al., 2011). The ancestral/derived states of SNPs were inferred upon the human ancestor sequences and were used to determine derived allele frequency.

Analysing the cis-effect of SNPs in DBSs on target gene expression

SNPs with MAF >0.1 in DBSs in any of the three modern human populations and absolute values of cis-effect size >0.5 (FDR <0.05) in any of the GTEx tissues were examined for an influence on the expression of the target genes (GTEx Consortium, 2017). SNPs that are eQTLs in the GTEx tissues and have biased derived allele frequencies (DAF) in the three modern human populations were examined to estimate whether the eQTL is population-specific.

Examining the tissue-specific impact of HS lncRNA-regulated gene expression

First, we examined the expression of HS lncRNA genes across the GTEx tissues. HS lncRNA genes with a median TPM value >0.1 in a tissue were considered robustly expressed in that tissue. Upon this criterion, 40 HS lncRNA genes were expressed in at least one tissue and were used to examine the impact of HS lncRNA regulation on gene and transcript expression (other HS lncRNAs may function in the cytoplasm). Since an HS lncRNA gene may have multiple transcripts, we adopted the transcript, which contains the predicted DNA binding domain (DBD) and has the largest TPM value as the representative transcript of the HS lncRNA. We calculated the pairwise Spearman’s correlation coefficient between the expression of an HS lncRNA (the representative transcript) and the expression of each of its target transcripts using the scipy.stats.spearmanr program in the scipy package. The expression of an HS lncRNA gene and a target transcript was considered to be significantly correlated if the |Spearman’s rho| > 0.3, with Benjamini-Hochberg FDR <0.05. We examined the percentage distribution of correlated HS lncRNA-target transcript pairs across GTEx tissues and organs (Figure 3A).

We examined all GTEx tissues to determine in which tissues HS lncRNA-regulated gene expression may have undergone significant changes from archaic to modern humans. If a pair of HS lncRNA and target transcript are robustly expressed in a tissue and their expression shows a significant correlation (|Spearman’s rho| >0.3, with Benjamini-Hochberg FDR <0.05) in the tissue, we computed the sequence distance of the HS lncRNA’s DBS in the transcript from the three archaic humans to modern humans. We compared the sequence distances of all DBSs in each tissue with the sequence distances of all DBSs in all GTEx tissues (i.e. the distribution of all DBS distances was taken as the background). A one-sided two-sample Kolmogorov-Smirnov test was used to examine whether the sequence distances of all DBSs in a specific tissue deviate from the background distribution (which reflects the "neutral evolution" of gene expression). For each tissue, if the test result was Benjamini-Hochberg FDR < 0.001, the tissue was considered to have significantly changed gene expression regulated by the HS lncRNA. We used different colors to mark tissues with significantly changed gene expression regulation since Altai Neanderthals, Denisovans, and Vindija Neanderthals, and used "D", "A.D.", and "ADV" to indicate changes since Denisovans, since Altai Neanderthals and Denisovans, and since Altai Neanderthals, Denisovans, and Vindija Neanderthals, respectively (Figure 3B).

Examining enrichment of favored and hitchhiking mutations in DBSs

Using the deep learning network DeepFavored we developed that integrates multiple statistical tests for identifying favored mutations (Tang et al., 2022), we identified 13339 favored mutations and 244098 hitchhiking mutations in 17 human populations (Tang et al., 2023). In this study, we classified DBSs in two ways - into strong ones (affinity>60) and weak ones (36<affinity<60), and into old ones (Human-Chimp distance>0.034 AND Human-Altai Neanderthals distance==0), young ones (Human-Altai Neanderthals distance>0.034 OR Human-Denisovan distance>0.034), and others. We then examined the numbers of favored and hitchhiking mutations in each class of DBSs. The weak young DBSs have the largest proportion of favored and hitchhiking mutations.

Identifying and analyzing transcriptional regulatory modules

Most clustering algorithms classify genes into disjoint modules based on expression correlation without considering regulatory relationships (Saelens et al., 2018). The GRAM program identifies gene regulatory modules based on correlation and TF-TFBS binding (Bar-Joseph et al., 2003). LncRNAs transcriptionally regulate genes based on lncRNA-DBS binding and correlate gene expression. We developed the eGRAM program to identify gene modules based on correlated expression, TF-TFBS interaction, and lncRNA-DBS interaction. In this study, we used eGRAM to identify gene modules in the same regions in the human and macaque brain and enriched KEGG pathways using reported RNS-seq datasets (n=101 and 83 for frontal cortex and n=22 and 25 for anterior cingulate cortex) (GTEx Consortium, 2017; Zhu et al., 2018). The default parameters of DBS affinity=60, Pearson correlation=0.5, module size=50, and FDR=0.01 (hypergeometric distribution test) were used. The key steps of program are as follows. (a1) Identify each lncRNA’s correlated lncRNAs, which may form a set of co-regulators. (a2, optional) Identify each TF’s correlated TFs, which may form a set of co-regulators. (b1) Compute the correlation between each lncRNA and all genes. (b2, optional) Compute the correlation between each TF and all genes. (c1) Identify each lncRNA’s target genes. (c2, optional) Identify each TF’s target genes. (d1) Identify each lncRNA set’s target module upon correlation and targeting relationships. (d2, optional) Identify each TF set’s target module upon correlation and targeting relationships. (e) Check whether TFs’ modules contain lncRNAs’ targets and whether lncRNAs’ modules contain TFs’ targets, which reveal genes co-regulated by TFs and lncRNAs and genes independently regulated by TFs and lncRNAs. (f) Performs pathway enrichment analysis for all modules.

eLife Assessment

This valuable study uses tools of population and functional genomics to examine long non-coding RNAs (lncRNAs) in the context of human evolution. Analyses of computationally predicted human-specific lncRNAs and their genomic targets lead to the development of hypotheses regarding the potential roles of these genetic elements in human biology. The conclusions regarding evolutionary acceleration and adaptation, however, only incompletely take data and literature on human/chimpanzee genetics and functional genomics into account.

Reviewer #2 (Public review):

In this valuable manuscript, Lin et al attempt to examine the role of long non coding RNAs (lncRNAs) in human evolution, through a set of population genetics and functional genomics analyses that leverage existing datasets and tools. Although the methods are incomplete and at times inadequate, the results nonetheless point towards a possible contribution of long non coding RNAs to shaping humans, and suggest clear directions for future, more rigorous study.

Comments on revisions:

I thank the authors for their revision and changes in response to previous rounds of comments. As it had been nearly two years since I last saw the manuscript, I reread the full text to familiarise myself again with the findings presented. While I appreciate the changes made and think they have strengthened the manuscript, I still find parts of it a bit too speculative or hyperbolic. In particular, I think claims of evolutionary acceleration and adaptation require more careful integration with existing human/chimpanzee genetics and functional genomics literature. For example:

Line 155: "About 5% of genes have significant sequence differences in humans and chimpanzees," This statement needs a citation, and a definition of what is meant by 'significant', especially as multiple lines below instead mention how it's not clear how many differences matter, or which of them, etc.

line 187: "Notably, 97.81% of the 105141 strong DBSs have counterparts in chimpanzees, suggesting that these DBSs are similar to HARs in evolution and have undergone human-specific evolution." I do not see any support for the inference here. Identifying HARs and acceleration relies on a far more thorough methodology than what's being presented here. Even generously, pairwise comparison between two taxa only cannot polarise the direction of differences; inferring human-specific change requires outgroups beyond chimpanzee.

line 210: "Based on a recent study that identified 5,984 genes differentially expressed between human-only and chimpanzee-only iPSC lines (Song et al., 2021), we estimated that the top 20% (4248) genes in chimpanzees may well characterize the human-chimpanzee differences" I do not agree with the rationale for this claim, and do not agree that it supports the cutoff of 0.034 used below. I also find that my previous concerns with the very disparate numbers of results across the three archaics have not been suitably addressed.

I also think that there is still too much of a tendency to assume that adaptive evolutionary change is the only driving force behind the observed results in the results. As I've stated before, I do not doubt that lncRNAs contribute in some way to evolutionary divergence between these species, as do other gene regulatory mechanisms; the manuscript leans down on it being the sole, or primary force, however, and that requires much stronger supporting evidence. Examples include, but are not limited to:

line 230: "These results reveal when and how HS lncRNA-mediated epigenetic regulation influences human evolution." This statement is too speculative.

Line 268: "yet the overall results agree well with features of human evolution." What does this mean? This section is too short and unclear.

Line 325: "and form 198876 HS lncRNA-DBS pairs with target transcripts in all tissues." This has not been shown in this paper - sequence based analyses simply identify the *potential* to form pairs.

Line 423: "Our analyses of these lncRNAs, DBSs, and target genes, including their evolution and interaction, indicate that HS lncRNAs have greatly promoted human evolution by distinctly rewiring gene expression." I do not agree that this conclusion is supported by the findings presented - this would require significant additional evidence in the form of orthogonal datasets.

I also return briefly to some of my comments before, in particular on the confounding effects of gene length and transcript/isoform number. In their rebuttal the authors argued that there was no need to control for this, but this does in fact matter. A gene with 10 transcripts that differ in the 5' end has 10 times as many chances of having a DBS than a gene with only 1 transcript, or a gene with 10 transcripts but a single annotated TSS. When the analyses are then performed at the gene level, without taking into account the number of transcripts, this could introduce a bias towards genes with more annotated isoforms. Similarly, line 246 focuses on genes with "SNP numbers in CEU, CHB, YRI are 5 times larger than the average." Is this controlled for length of the DBS? All else being equal a longer DBS will have more SNPs than a shorter one. It is therefore not surprising that the same genes that were highlighted above as having 'strong' DBS, where strength is impacted by length, show up here too.

Author response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #2 (Public review):

In this valuable manuscript, Lin et al attempt to examine the role of long non coding RNAs (lncRNAs) in human evolution, through a set of population genetics and functional genomics analyses that leverage existing datasets and tools. Although the methods are incomplete and at times inadequate, the results nonetheless point towards a possible contribution of long non coding RNAs to shaping humans, and suggest clear directions for future, more rigorous study.

Comments on revisions:

I thank the authors for their revision and changes in response to previous rounds of comments. As it had been nearly two years since I last saw the manuscript, I reread the full text to familiarise myself again with the findings presented. While I appreciate the changes made and think they have strengthened the manuscript, I still find parts of it a bit too speculative or hyperbolic. In particular, I think claims of evolutionary acceleration and adaptation require more careful integration with existing human/chimpanzee genetics and functional genomics literature.

We thank the reviewer heartfully for the great patience and valuable comments, which have helped us further improve the manuscript. Before responding to comments point by point, we provide a summary here.

(1) On parameters and cutoffs.

Parameters and cutoffs influence data analysis. The large number of Supplementary Notes, Supplementary Figures, and Supplementary Tables indicates that we paid great attention to the influence of parameters and robustness of analyses. Specifically, here we explain the DBS sequence distance cutoff of 0.034, which determines the top 20% genes that most differentiate humans from chimpanzees and influences the gene set enrichment analysis (Figure 2). As described in the revised manuscript, we estimated this cutoff based on Song et al., verified its rationality based on Prufer et al. (Song et al. 2021; Prufer et al. 2017), and measured its influence by examining slightly different cutoff values (e.g., 0.035).

(2) Analyses of HS TFs and HS TF DBSs.

It is desirable to compare the contribution of HS lncRNAs and HS TFs to human evolution. Identifying HS TFs faces the challenges that different institutions (e.g., NCBI and Ensembl) annotate orthologous genes using different criteria, and that multiple human TF lists have been published by different research groups. Recently, Kirilenko et al. identified orthologous genes in hundreds of placental mammals and birds and organized different types of genes into datasets of parewise comparison (e.g., hg38-panTro6) using humans and mice as references (Kirilenko et al. Integrating gene annotation with orthology inference at scale. Science 2023). Based on (a) the many2zero and one2zero gene lists in the “hg38-panTro6” dataset, (b) three human TF lists reported by two studies (Bahram et al. 2015; Lambert et al. 2018) and used in the SCENIC package, we identified HS TFs. The number of HS TFs and HS lncRNAs (5 vs 66) alone lends strong evidence suggesting that HS lncRNAs have contributed more significantly to human evolution than HS TFs (note that 5 is the union of three intersections between <many2zero + one2zero> and the three ).

TF DBS (i.e., TFBS) prediction has also been challenging because they are very short (mostly about 10 bp) and TF-DNA binding involves many cofactors (Bianchi et al. Zincore, an atypical coregulator, binds zinc finger transcription factors to control gene expression. Science 2025). We used two TF DBS prediction programs to predict HS TF DBSs, including the well-established FIMO program (whose results have been incorporated into the JASPAR database) (Rauluseviciute et al. JASPAR 2024: 20th anniversary of the open-access database of transcription factor binding profiles Open Access. NAR 2023) and the recently reported CellOracle program (Kamimoto et al. Dissecting cell identity via network inference and in silico gene perturbation. Nature 2023). Then, we performed downstream analyses and obtained two major results. One is that on average (per base), fewer selection signals are detected in HS TF DBSs (anyway, caution is needed because TF DBSs are very short); the other is that HS TFs and HS lncRNAs contribute to human evolution in quite different ways (Supplementary Figs. 25 and 26).

(3) On genes with more transcripts may appear as spurious targets of HS lncRNAs.

Now, the results of HS TF DBSs allow us to address the question of whether genes with more transcripts may appear as spurious targets of HS lncRNAs. We note that (a) we predicted HS lncRNA DBSs and HS TF DBSs in the same promoter regions before the same 179128 Ensembl-annotated transcripts (release 79), (b) we used the same GTEx transcript expression matrices in the analyses of HS TF DBSs and HS lncRNA DBSs (the GTEx database includes gene expression matrices and transcript expression matrices, the latter includes multiple transcripts of a gene). Thus, the analyses of HS TF DBSs provide an effective control for examining the question of whether genes with more transcripts may appear as spurious targets of HS lncRNAs, and consequently, cause the high percentages of HS lncRNA-target transcript pairs that show correlated expression in the brain (Figure 3). We find that the percentages of HS TF-target transcript pairs that show correlated expression are also high in the brain, but the whole profile in GTEx tissues is significantly different from that of HS lncRNA DBSs (Figure 3A; Supplementary Figure 25). On the other hand, on the distribution of significantly changed DBSs in GTEx tissues, the difference between HS lncRNA DBSs and HS TF DBSs is more apparent (Figure 3B; Supplementary Figure 26). Together, these suggest that the brain-enriched distribution of co-expressed HS lncRNA-target transcript pairs must arise from HS lncRNA-mediated transcriptional regulation rather than from the transcript number difference.

(4) Additional notes on HS TFs and HS TF DBSs.

First, the “many2zero” and “one2zero” gene lists in the “hg38-panTro6” dataset of Kirilenko et al. provide the most update, but not most complete, data on human-specific genes because “hg38-panTro6” is a pairwise comparison. On the other hand, the Ensembl database also annotates orthologous genes, but lacks such pairwise comparisons as “hg38-panTro6”. Therefore, not all HS genes based on “hg38-panTro6” agree with orthologous genes in the Ensembl database. Second, if HS genes are identified based on both Ensembl and Kirilenko et al., HS TFs will be fewer.

(5) On speculative or hyperbolic claims.

First, the title “Human-specific lncRNAs contributed critically to human evolution by distinctly regulating gene expression” is now further supported by HS TF DBSs analyses. Second, we have carefully revised the entire manuscript, trying to make it more readable, accurate, logically reasonable, and biologically acceptable. Third, specifically, in the revision, we avoid speculative or hyperbolic claims in results, interpretations, and discussions as possible as we can. This includes the tone-down of statements and claims, for example, using “reshape” to replace “rewire” and using “suggest” to replace “indicate”. Since the revisions are pervasive, we do not mark all of them, except those that are directly relevant to the reviewer’s comments.

(1) Line 155: "About 5% of genes have significant sequence differences in humans and chimpanzees," This statement needs a citation, and a definition of what is meant by 'significant', especially as multiple lines below instead mention how it's not clear how many differences matter, or which of them, etc.

Different studies give different estimates, from 1.24% (Ebersberger et al. Genomewide Comparison of DNA Sequences between Humans and Chimpanzees. Am J Hum Genet. 2002) to 5% (Britten RJ. Divergence between samples of chimpanzee and human DNA sequences is 5%, counting indels. PNAS 2002). The 5% for significant gene sequence differences arises when considering a broader range of genetic variations, particularly insertions and deletions of genetic material (indels). To provide more accurate information, we have replaced this simple statement with a more comprehensive one and cited the above two papers.

(2) line 187: "Notably, 97.81% of the 105141 strong DBSs have counterparts in chimpanzees, suggesting that these DBSs are similar to HARs in evolution and have undergone human-specific evolution." I do not see any support for the inference here. Identifying HARs and acceleration relies on a far more thorough methodology than what's being presented here. Even generously, pairwise comparison between two taxa only cannot polarise the direction of differences; inferring human-specific change requires outgroups beyond chimpanzee.

Here, we actually made an analogy but not an inference; therefore, we used such words as “suggesting” and “similar” instead of using more confirmatory words. We have revised the latter half sentence, saying “raising the possibility that these sequences have evolved considerably during human evolution”.

(3) line 210: "Based on a recent study that identified 5,984 genes differentially expressed between human-only and chimpanzee-only iPSC lines (Song et al., 2021), we estimated that the top 20% (4248) genes in chimpanzees may well characterize the human-chimpanzee differences". I do not agree with the rationale for this claim, and do not agree that it supports the cutoff of 0.034 used below. I also find that my previous concerns with the very disparate numbers of results across the three archaics have not been suitably addressed.

(1) Indeed, “we estimated that the top 20% (4248) genes in chimpanzees may well characterize the human-chimpanzee differences” is an improper claim; we made this mistake due to the flawed use of English.

(2) What we need is a gene number, which (a) indicates genes that effectively differentiate humans from chimpanzees, (b) can be used to set a DBS sequence distance cutoff. Since this study is the first to systematically examine DBSs in humans and chimpanzees, we must estimate this gene number based on studies that identify differentially expressed genes in humans and chimpanzees. We choose Song et al. 2021 (Song et al. Genetic studies of human–chimpanzee divergence using stem cell fusions. PNAS 2021), which identified 5984 differentially expressed genes, including 4377 genes whose differential expression is due to trans-acting differences between humans and chimpanzeees. To the best of our knowledge, this is the only published data on trans-acting differences between humans and chimpanzeees, and most HS lncRNAs and their DBSs/targets have trans-acting relationships (see Supplementary Table 2). Based on these numbers, we chose a DBS sequence distance cutoff of 0.034, which corresponds to 4248 genes (the top 20%), slightly fewer than 4377.

(3) If we chose DBS sequence distance cutoff=0.033 or 0.035, slightly more or fewer genes would be determined, raising the question of whether they would significantly influence the downstream gene set enrichment analysis (Figure 2). We found that 91 genes have a DBS sequence distance of 0.034. Thus, if cutoff=0.035, 4248-91=4157 genes were determined, and the influence on gene set enrichment analysis was very limited.

(4) On the disparate numbers of results across the three archaics. Figure 1A is based on Figure 2 in Prufer et al. 2017. At first glance, our Figure 1A indicates that Altai Neanderthal is older than Denisovan (upon kya), making our result “identified 1256, 2514, and 134 genes in Altai Neanderthals, Denisovans, and Vindija Neanderthals” unreasonable. However, Prufer et al. (2017) reported that “It has been suggested that Denisovans received gene flow from a hominin lineage that diverged prior to the common ancestor of modern humans, Neandertals, and Denisovans……In agreement with these studies, we find that the Denisovan genome carries fewer derived alleles that are fixed in Africans, and thus tend to be older, than the Altai Neandertal genome”. This note by Prufer et al. provides an explanation for our result, which is that more genes with large DBS sequence distances were identified in Denisovans than in Altai Neanderthals. Of course, the 1256, 2514, and 134 depend on the cutoff of 0.034. If cutoff=0.035, these numbers change slightly, but their relationships remain (i.e., more genes in Denisovans). We examined multiple cutoff values and found that more genes in Denisovans have large DBS sequence distances than in Altai Neanderthals.

(4) I also think that there is still too much of a tendency to assume that adaptive evolutionary change is the only driving force behind the observed results in the results. As I've stated before, I do not doubt that lncRNAs contribute in some way to evolutionary divergence between these species, as do other gene regulatory mechanisms; the manuscript leans down on it being the sole, or primary force, however, and that requires much stronger supporting evidence. Examples include, but are not limited to:

(1) Indeed, the observed results are also caused by other genomic elements and mechanisms (but it is hardly feasible to identify and differentiate them in a single study), and we do not assume that adaptive evolutionary change is the only driving force. Careful revisions have been made to avoid leaving readers the impression that we have this tendency or hold the simple assumption.

(2) Comparing HS lncRNAs to HS TFs is critical, and we have done this.

(5) line 230: "These results reveal when and how HS lncRNA-mediated epigenetic regulation influences human evolution." This statement is too speculative.

We have toned down the statement, just saying “These results provide valuable insights into when and how HS lncRNA-mediated epigenetic regulation impacts human evolution”.

Line 268: "yet the overall results agree well with features of human evolution." What does this mean? This section is too short and unclear.

(1) First, the sentence “Selection signals in YRI may be underestimated due to fewer samples and smaller sample sizes (than CEU and CHB), yet the overall results agree well with features of human evolution” has been deleted, because CEU, CHB, and YRI samples are comparable (100, 99, and 97, respectively).

(2) Now the sentence has been changed to “These results agree well with findings reported in previous studies, including that fewer selection signals are detected in YRI (Sabeti et al., 2007; Voight et al., 2006)”.

(3) On “This section is too short and unclear” - To make the manuscript more readable, we adopt short sections instead of long ones. This section expresses that (a) our finding that more selection signals were detected in CEU and CHB than in YRI agrees with well-established findings (Voight et al. A Map of Recent Positive Selection in the Human Genome. PLoS Biology 2006; Sabeti et al. Genome-wide detection and characterization of positive selection in human populations. Nature 2007), (b) in considerable DBSs, selection signals were detected by multiple tests.

Line 325: "and form 198876 HS lncRNA-DBS pairs with target transcripts in all tissues." This has not been shown in this paper - sequence based analyses simply identify the “potential” to form pairs.

This section describes transcriptomic analysis using the GTEx data. Indeed, target transcripts of HS lncRNAs are results of sequence-based analysis, and a predicted target is not necessarily regulated by the HS lncRNA in a tissue. Here, “pair” means a pair of HS lncRNA-target transcript whose expression shows significant Pearson correlation in a GTEx tissue (by the way, we do not mean correlation equals regulation; actually, we identified HS lncRNA-mediated transcriptional regulation upon both DBS-targeting relationship and correlation relationship).

Line 423: "Our analyses of these lncRNAs, DBSs, and target genes, including their evolution and interaction, indicate that HS lncRNAs have greatly promoted human evolution by distinctly rewiring gene expression." I do not agree that this conclusion is supported by the findings presented - this would require significant additional evidence in the form of orthogonal datasets.

(1) As mentioned above, we have used “reshape” to replace “rewire” and used “suggest” to replace “indicate”. In addition, we have substantially revised the Discussion, in which this sentence is replaced by “our results suggest that HS lncRNAs have greatly reshaped (or even rewired) gene expression in humans”.

(2) Multiple citations have been added, including Voight et al. 2006 (Voight et al. A Map of Recent Positive Selection in the Human Genome. PLoS Biology 2006) and Sabeti et al. 2007 (Sabeti et al. Genome-wide detection and characterization of positive selection in human populations. Nature 2007).

(3) We have analyzed HS TF DBSs, and the obtained results also support the critical contribution of HS lncRNAs.

I also return briefly to some of my comments before, in particular on the confounding effects of gene length and transcript/isoform number. In their rebuttal the authors argued that there was no need to control for this, but this does in fact matter. A gene with 10 transcripts that differ in the 5' end has 10 times as many chances of having a DBS than a gene with only 1 transcript, or a gene with 10 transcripts but a single annotated TSS. When the analyses are then performed at the gene level, without taking into account the number of transcripts, this could introduce a bias towards genes with more annotated isoforms. Similarly, line 246 focuses on genes with "SNP numbers in CEU, CHB, YRI are 5 times larger than the average." Is this controlled for length of the DBS? All else being equal a longer DBS will have more SNPs than a shorter one. It is therefore not surprising that the same genes that were highlighted above as having 'strong' DBS, where strength is impacted by length, show up here too.

(1) In gene set enrichment analysis (Figure 2, which is a gene-level analysis), when determining genes differentiating humans from chimpanzees based on DBS sequence distance, if a gene has multiple transcripts/DBSs, we choose the DBS with the largest distance. That is, the input to g:Profiler is a non-redundant gene list.

(2) In GTEx data analysis (Figure 3, which is a transcriptome-level analysis), the analyses of HS TF DBSs using the GTEx data provide evidence suggesting that different DBS/transcript numbers of genes are unlikely to cause confounding effects. As explained above, we predicted HS TF DBSs in the same promoter regions of 179128 Ensembl-annotated transcripts (release 79), but Supplementary Figures 25 and 26 are distinctly different from Figure 3AB.

(3) In evolutionary analysis, a gene with 10 DBSs has a higher chance of having selection signals than a gene with 1 DBS. This is biologically plausible, because many conserved genes have novel transcripts whose expression is species-, tissue-, or developmental period-specific, and DBSs before these novel transcripts may differ from DBSs before conserved transcripts.

(4) “line 246 focuses on genes with "SNP numbers in CEU, CHB, YRI are 5 times larger than the average." Is this controlled for the length of the DBS?” - This is a defect. We have now computed SNP numbers per base and used the new table to replace the old Supplementary Table 8. After examining the new table, we find that the major results of SNP analysis remain.

(5) On “Is this controlled for length of the DBS? All else being equal a longer DBS will have more SNPs than a shorter one” - We do not think there are reasons to control for the length of DBSs; also, what “All else being equal” means matters. First, DBS sequences have specific features; thus, the feature of a long DBS is stronger than the feature of a short one, making a long DBS less likely to be generated by chance in the genome and less likely to be predicted wrongly than a short one. This means that longer DBSs are less likely to be false ones (note our explanation that the chance of a DBS of 147 bp, the mean length of DBSs, to be wrongly predicted is extremely low, p<8.2e-19 to 1.5e-48). Second, the difference in length suggests a difference in binding affinity, which in turn influences the regulation of the specific transcripts and influences the analysis of GTEx data. Third, it cannot be excluded that some SNPs may be selection signals (detecting selection signal is challenging, and many selection signals cannot be detected by statistical tests, see Grossman et al. A composite of multiple signals distinguishes causal variants in regions of positive selection. Science 2010).

(6) On “It is therefore not surprising that the same genes that were highlighted above as having 'strong' DBS, where strength is impacted by length” - Indeed, strength is influenced by length, see the above response.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

Finally, figure 1 panels D and F are not legible - the font is tiny! There's also a typo in panel A, where "Homo Sapien" should be "Homo sapiens".

(1) “Homo sapien” is changed to “Homo sapiens”.

(2) Even if we double the font size, they are still too small. Inserting a very large panel D into Figure 1 will make Figure 1 ugly, and converting Figure 1D into an independent figure is unnecessary. Actually, panels 1D and F are illustrative figures; the full Fig.1D is Supplementary Figure 6, and the full Fig.1F is Figure 3. We have revised Fig.1’s legend to explain these.

Version 4

Abstract

What genes and regulatory sequences differentiate humans from apes and from archaic humans, which share highly similar genomes but show distinct phenotypes, has puzzled researchers for decades. Previous studies have examined species-specific protein-coding genes and related regulatory sequences, revealing that birth, loss, and changes in these genes and sequences can drive speciation and evolution. However, investigations of species-specific lncRNA genes and the related regulatory sequences, which critically determine epigenetic regulation, remain very limited. We identified human-specific (HS) lncRNAs from GENCODE-annotated human lncRNAs, predicted their DNA binding domains (DBDs) and binding sites (DBSs) genome-wide, analyzed the DBS sequences in modern humans (CEU, CHB, and YRI), archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and chimpanzees, and analyzed how HS lncRNAs and DBSs influence gene expression in modern and archaic humans. Our results suggest that HS lncRNAs and their DBSs have substantially reshaped gene expression. This reshaping has evolved continuously from archaic to modern humans, enabling humans to adapt to new environments and lifestyles, promoting brain evolution, and also resulting in differences among modern humans. Further analyses of HS TFs and their DBSs obtained results suggesting that HS lncRNAs may contribute more significantly than HS TFs to human evolution.

Introduction

The limited genomic but substantial phenotypic and behavioral differences between humans and other hominids make the question of which sequence changes have critically driven human evolution an enduring puzzle (Anton et al., 2014; Pollen et al., 2023). Sequence changes include the birth and loss of genes, as well as the turnover and modifications of regulatory sequences (Albalat and Canestro, 2016; Kaessmann, 2010; Prud’homme et al., 2007). Waves of studies have identified genes that promote human brain development (e.g., NOTCH2NL and ASPM) (Evans et al., 2005; Fiddes et al., 2018; Mekel-Bobrov et al., 2005; Pinson et al., 2022; Suzuki et al., 2018). However, many conclusions from the gene birth-centric studies are controversial (Currat et al., 2006; Timpson et al., 2007; Yu et al., 2007). Since genes essential for brain development may not necessarily regulate other human traits (some of which preceded brain enlargement), these new brain-related genes do not fully answer the above question. Moreover, investigations into how species-specific lncRNAs, a major class of new genes in mammals, contribute to human evolution remain limited. Studies revealed regulatory sequence changes important for human evolution, including new sequences (Liu et al., 2021), lost sequences (McLean et al., 2011), human accelerated regions (HARs) (Dong et al., 2016; Mangan et al., 2022; Prabhakar et al., 2006; Whalen and Pollard, 2022), and turnover of transcription factor (TF) binding sites (Krieger et al., 2022; Otto et al., 2009; Zhang et al., 2023). Nevertheless, no systematic analysis of lncRNA binding sites, which also regulate transcription, has been reported.

Experimental studies over the past decades have generated substantial findings about lncRNAs and epigenetic regulation. (a) About one-third of human lncRNAs are primate-specific (Derrien et al., 2012). (b) Species-specific lncRNAs exhibit distinct expression in tissues and organs of different species (Sarropoulos et al., 2019). (c) Many lncRNAs can bind to DNA sequences by forming RNA:DNA triplexes (Abu Almakarem et al., 2012), thereby recruiting histone and DNA modification enzymes to binding sites and regulating transcription (Lee, 2009; Roadmap Epigenomics et al., 2015). (d) Approximately 40% of differentially expressed human genes result from interspecies epigenetic differences (Hernando-Herraez et al., 2015). Thus, not only human-specific (HS) TFs and their DNA binding sites (TF DBS), but also HS lncRNAs and their DNA binding sites (lncRNA DBS), critically regulate gene expression human-specifically.

To examine the latter, which have had few reports, the first step is to identify HS lncRNAs. Unlike DNA sequence search, RNA sequence search for identifying homologous and orthologous lncRNA genes needs combined structure and sequence alignment (Nawrocki, 2014; Nawrocki and Eddy, 2013), and human lncRNAs’ orthologues in other mammals determine HS lncRNAs. Many lncRNAs contain DNA binding domains (DBDs) that can bind to DBSs. Since RNA:DNA triplexes follow specific base-pairing rules (Abu Almakarem et al., 2012), these rules allow computational prediction of DBDs and DBSs, and accordingly, target genes and transcripts (Lin et al., 2019; Wen et al., 2022). Gene expression in organs and tissues, especially those examined by the GTEx project (GTEx Consortium, 2017), provides data to investigate whether the expression of lncRNAs and their targets is correlated, supporting their regulatory relationship. Finally, the sequenced genomes of modern humans, archaic humans, and multiple apes allow cross-species analysis of the genomes and transcriptomes (1000 Genomes Project Consortium et al., 2012; Chimpanzee Sequencing and Analysis Consortium, 2005; Meyer et al., 2012; Prufer et al., 2017; Prüfer et al., 2013). Thus, it is feasible to examine whether and how HS lncRNAs and their DBSs have contributed to human evolution by human-specifically regulating gene expression.

This study identified HS lncRNAs upon the orthologues of the GENCODE-annotated human lncRNA genes in 16 mammalian genomes, predicted HS lncRNAs’ DBDs and DBSs in the genomes of modern humans (CEU, CHB, and YRI), archaic humans (Altai Neanderthals, Denisovans, Vindija Neanderthals), as well as counterparts of DBSs in the genome of chimpanzees (Figure 1A). Joint prediction of DBDs and DBSs (which form RNA:DNA triplexes) using local alignment is more reasonable than (and outperforms) DBS prediction based on base-pairing rules (Lin et al., 2019). We validated the DBS prediction by knocking out (KO) predicted DBDs using CRISPR/Cas9, performing RNA-seq before and after KO, and examining whether the differentially expressed genes were predicted target genes. To accurately analyze DBSs with different lengths and identity values (the percentage of paired RNA and DNA nucleotides in a triplex), we defined binding affinity (length x identity) and classified DBSs into strong and weak ones based on binding affinity and also old and young ones based on when most of their sequence changes occurred. We then utilized multiple datasets and methods to perform cross-population and cross-species analyses, including gene set enrichment analyses, evolutionary/population genetics analyses, and transcriptomic analyses. The results suggest that HS lncRNAs and their DBSs have distinctly and continuously reshaped gene expression during human evolution, and this reshaped gene expression greatly determines human traits, from the rapidly evolved brain to the highly modified immune and metabolism systems.

Study Overview.

(A) The relationships between chimpanzees, the three archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and the three modern human populations. Dashed lines indicate the phylogenetic distances from modern humans to archaic humans and chimpanzees. Genes and DBSs have the following implications: the DBS in B2M lacks a counterpart in chimpanzees; the DBS in ABL2 has great differences between archaic and modern humans; the DBS in IRNAR1 is highly polymorphic in modern humans. Red letters in DBSs indicate tissue-specific eQTLs or population-specific favored mutations. (B) The mean length and binding affinity of strong DBSs. (C) Number of target genes and transcripts of the 66 HS lncRNAs. (D) The illustrative figure shows the targeting relationships between HS lncRNAs (see the full figure in Supplementary Note 2). (E) The sequence distances of DBSs (the top 40%) from modern humans to chimpanzees and archaic humans. (F) The illustrative figure shows the GTEx tissues used to investigate the impacts of HS lncRNA-target transcript pairs on gene expression (see the full figure in Figure 3).

Results

HS lncRNAs regulate diverse genes and transcripts

How many human lncRNAs are human-specific is a question without a precise estimate. After searching for orthologues of the 13562 GENCODE-annotated human lncRNA genes (v18) in 16 mammalian genomes and the genomes of archaic humans using the Infernal program (which performs a combined RNA sequence and structure alignment) (Figure 1A) (Lin et al., 2019; Nawrocki, 2014; Nawrocki et al., 2009), we identified 66 lncRNA genes that exist in humans but not in any other species (Supplementary Table 1). Then, we predicted DBSs of the 66 HS lncRNAs in the 5000 bp promoter regions of the 179128 Ensembl-annotated transcripts (release 79) using the LongTarget program (Supplementary Note 1) (Lin et al., 2019). Since DBD1’s DBSs vastly outnumber other DBDs’ DBSs, suggesting that DBD1 (hereafter simply called DBD) and its DBSs are more reliable, for each HS lncRNA, we analysed only DBD1 and its DBSs. We defined the binding affinity of a DBS as the product of its length and identity (the percentage of paired RNA and DNA nucleotides), and focused on DBSs with binding affinity>=60 (strong DBSs). The analysis of DBSs at known imprinted genes and the comparison of DBSs with CHART-seq-reported binding sequences (“peaks”) suggest that binding affinity characterizes DBSs better than length (Supplementary Fig. 2) (He et al., 2015). Since DBSs generated more recently may be weak yet play new and/or subtle functions, we also analyzed DBSs with 36<binding affinity<60 (weak DBSs). 105141 strong DBSs (mean length>147 bp) were identified in 96789 transcripts of 23396 genes, and 152836 weak DBSs were identified in 127898 transcripts of 33185 genes (Figure 1B; Supplementary Table 2, 3). Several HS lncRNAs (especially RP11-423H2.3) have abundant targets; many targets have DBSs of multiple HS lncRNAs; many HS lncRNAs have multiple DBSs in a target; only about 1.5% of target genes (0.6% of target transcripts) are human-specific (Figure 1C); and HS lncRNAs themselves show complex targeting relationships (Figure 1D; Supplementary Note 2). These suggest that the limited number of HS lncRNAs has significantly reshaped the expression of substantial genes.

We validated DBS prediction using multiple methods and datasets (Supplementary Note 1; Supplementary Fig. 1-5). (a) HS lncRNA DBSs overlap well with experimentally identified DNA methylation and histone modification signals in multiple cell lines (https://genome.UCSC.edu). (b) Many DBSs co-localize with annotated cis-regulatory elements (cCREs) in genes’ promoter regions (https://genome.UCSC.edu). (c) Predicted DBSs overlap well with experimentally detected DBSs of NEAT1, MALAT1, and MEG3 (Mondal et al., 2015; West et al., 2014). (d) We used CRISPR/Cas9 to knock out predicted DBDs (100-200 bp) in several lncRNAs across multiple cell lines, performed RNA-seq before and after DBD knockout, and analyzed the resulting differential gene expression. We found that the |fold change| of target genes was significantly larger than the |fold change| of non-target genes (one-sided Mann-Whitney test, p = 3.1e-72, 1.49e-114, 1.12e-206, 2.58e-09, 6.49e-41, 0.034, and 5.23e-07, respectively). (e) The LongTarget program uses a variant of the Smith-Waterman local alignment to identify DBD and DBS simultaneously; statistically, a local alignment (i.e., a triplex containing a pair of DBD and DBS) of 147 bp (the mean length of strong DBSs) is extremely unlikely to be generated by chance (p < 8.2e-19 to 1.5e-48).

HARs are conserved genomic loci yet show accelerated evolution in the human lineage. Multiple researchers postulate that HARs critically promote human evolution; however, results from the related studies are controversial (Levchenko et al., 2018). The Zoonomia Project recently identified 312 HARs important for 3D genome rewiring and neurodevelopment (Keough et al., 2023). We found that only 8 of these HARs overlap 26 DBSs of 14 HS lncRNAs, suggesting that DBSs and HARs are different kinds of sequences.

HS lncRNAs’ target genes characterize human evolution and traits

AS strong and weak DBSs may occur in different periods in human evolution, it is interesting and necessary to examine whether the related target genes are enriched in different functions. For genes with significant sequence differences between humans and chimpanzees, studies have generated different estimates, ranging from 1.24% to 5% (Britten, 2002; Ebersberger et al., 2002). More genes exhibit expression differences due to variations in regulatory sequences. To reliably differentiate genes with strong and weak DBSs, we sorted target genes by their DBS binding affinity and chose the top 2000 genes (DBS length>252 bp and binding affinity>151) and bottom 2000 genes (DBS length<60 bp but binding affinity>36) to conduct over-representation analysis using the g:Profiler program and Gene Ontology (GO) database (Supplementary Table 4). The top and bottom 2000 genes (with the whole genome as the background) are co-enriched in GO terms important for human evolution, including “neuron projection development” (GO:0031175), “learning” (GO:0007612), “regulation of anatomical structure size” (GO: GO:0090066), “head development” (GO:0060322), and “regulation of locomotion” (GO:0040012) (Supplementary Note 3; Supplementary Table 4). Genes with strongest DBSs (514 DBSs have length >500 bp) include IFNAR1 and NFATC1 (important for immune function), KIF21B and NTSR1 (critical for neural development), SLC2A11 and SLC2A1 (involved in glucose usage), BAIAP3 (a brain-specific angiogenesis inhibitor), TAS1R3 (a receptor for the sweet taste response), and several primate-specific noncoding RNA genes (e.g., CTD-3224I3.3 which is highly expressed in the cerebellum, lung, and testis) (Table 1; Supplementary Table 4, 5). On the other hand, genes with weak DBSs are enriched in such GO terms as “response to alcohol” (GO:0097305), “response to temperature stimulus” (GO:0009266), “DNA methylation or demethylation” (GO:0044728), and “female pregnancy” (GO:0007565) (Supplementary Note 3). Since weak DBSs are more likely to have occurred recently, these findings suggest that some regulatory functions of HS lncRNAs were obtained recently.

The reshaping of gene expression has greatly evolved

The human genome is approximately 99%, 99%, and 98% identical to the genomes of chimpanzees, bonobos, and gorillas (Chimpanzee Sequencing and Analysis Consortium, 2005; Prufer et al., 2012). However, the sequence variants that critically differentiate these species remain unclear. We therefore examined whether DBSs of HS lncRNAs are conserved in these species or are human-specific. Notably, 97.81% of the 105141 strong DBSs have counterparts in chimpanzees, raising the possibility that these sequences have evolved considerably during human evolution.

HS lncRNAs may be generated before, with, or after their DBSs. To further clarify these situations, we identified the counterparts of HS lncRNAs and their DBSs also in Altai Neanderthals, Denisovans, and Vindija Neanderthals (Meyer et al., 2012; Prufer et al., 2017; Prüfer et al., 2013). Counterparts of HS lncRNAs and strong DBSs were identified in these archaic humans, but the distances (distance per base) of HS lncRNAs are smaller than the distances of DBSs between archaic and modern humans, suggesting that the counterpart sequences of many DBSs existed before HS lncRNAs. In other words, these distances suggest that HS lncRNAs may have reshaped gene expression substantially in humans.

We computed DBS distances in two ways to examine DBS evolution. The first is from the reconstructed human ancestor (downloaded from the EBI website) to chimpanzees, archaic humans, and modern humans; the second is from chimpanzees and archaic humans to modern humans. In the first set of distances, many DBS distances from the human ancestor to modern humans are smaller than those to archaic humans. For example, DBSs in the top 20% genes in the human ancestor have distances >0.015 to modern humans, but more DBSs have distances >0.015 to Altai Neanderthals, Denisovans, and even chimpanzees (Supplementary Note 3). This anomaly is caused by the reconstructed human ancestor, which is built using six primates without archaic humans. On the other hand, the second set of distances agrees well with the phylogenetic distances between chimpanzees, archaic humans, and modern humans. DBSs with large distances account for a small portion (Figure 1E); we postulate that genes containing these DBSs may greatly characterize human evolution. To set a cutoff for “large distance”, we turned to a recent study that identified 5984 genes differentially expressed between human-only and chimpanzee-only iPSC lines, including 4377 genes whose differential expression is due to trans-acting differences between humans and chimpanzees (Song et al., 2021). A distance cutoff of 0.034 identifies 4248 (the top 20%) genes in chimpanzees, as well as 1256, 2514, and 134 genes in Altai Neanderthals, Denisovans, and Vindija Neanderthals, respectively. Since Prufer et al. reported that “it has been suggested that Denisovans received gene flow from a hominin lineage that diverged prior to the common ancestor of modern humans, Neandertals, and Denisovans” and that “we find that the Denisovan genome carries fewer derived alleles that are fixed in Africans, and thus tend to be older, than the Altai Neandertal genome” (Prufer et al., 2017), these gene numbers agree with the phylogenetic relationships between archaic and modern humans (Figure 1A; Supplementary Table 6).

To determine whether target genes with mostly changed DBSs are enriched in specific functions, we applied overrepresentation analysis to target genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals using the g:Profiler program and GO database. To make statistical significance comparable across chimpanzees, Altai Neanderthals, and Denisovans, we used just the top 1256 genes as the input for the three. The 1256 genes in chimpanzees and Altai Neanderthals are enriched in multiple common GO terms. However, “behavior” (GO:0007610) and “cognition” (GO:0050890) have much higher significance in Altai Neanderthals than in chimpanzees (Figure 2), and some GO terms such as “learning” (GO:0007612), “learning or memory” (GO:0007611), and “glucose homeostasis” (GO:0042593) are enriched only in Altai Neanderthals. Agreeing with the closer relationship between Vindija Neanderthals and modern humans, the 134 genes in Vindija Neanderthals are enriched only in two GO terms with low significance (“smoothened signaling pathway” (GO:0007224) and “regulation of smoothened signaling pathway” (GO:0008589)) (Supplementary Table 7). These results provide valuable insights into when and how HS lncRNA-mediated epigenetic regulation impacts human evolution.

1256 target genes whose DBSs have the largest distances from modern humans to chimpanzees and Altai Neanderthals are enriched in different Biological Processes GO terms.

Upon significance threshold = 0.05 (Benjamini-Hochberg FDR), the two gene sets in chimpanzees and Altai Neanderthals are enriched in 199 and 409 GO terms (50<terms size<1000), respectively. Shared GO terms are color-marked. Left: Top GO terms enriched in genes in chimpanzees. Right: Top GO terms enriched in genes in Altai Neanderthals.

Epigenetic regulation by HS lncRNAs shows differences in modern humans

To check whether the reshaped gene expression has been population-specifically revised in modern humans, we analyzed single-nucleotide polymorphisms (SNPs, with minimal allele frequency (MAF)≥0.1) in strong DBSs using the 1000 Human Genome data (1000 Genomes Project Consortium et al., 2012). We computed SNP number and SNP number per base for each DBS, identified highly polymorphic DBSs (SNP number per base in CEU, CHB, and YRI are 5 times larger than the average), and obtained the intersection of highly polymorphic DBSs and DBSs with mostly changed sequence distances (distance>0.034, from modern humans to chimpanzees, Altai Neanderthals, and Denisovans) (Supplementary Table 8). Most genes in the intersection are lncRNA genes, and protein-coding genes include those important for human evolution and adaptation (especially metabolism and immune), such as SCTR, TLR, NCR, IFNAR1, NFATC1, TAS1R3, INS, ST3GAL4, and FN3KRP. The rich SNPs, in addition to large sequence distances and Tajima’s D, indicate DBS differences in modern humans. Genes involved in sugar metabolism are especially notable. A key feature of modern human diets is high sugar intake, which can cause non-enzymatic oxidation of proteins (i.e., glycation) and deactivate protein functions. Among proteins encoded by genes in this intersection, TAS1R3 recognizes diverse natural and synthetic sweeteners, Insulin decreases blood glucose concentration, ST3GAL4 regulates protein glycosylation, and FN3KRP deglycates proteins. These DBSs and target genes indicate continuous evolution of transcriptional regulation with human and population specificity.

To examine whether these SNPS are neutral or indicate selection signals, we next used multiple statistical tests (with widely adopted parameters), including XP-CLR (Chen et al., 2010), iSAFE (Akbari et al., 2018), Tajima’s D (Tajima, 1989), Fay-Wu’s H (Fay and Wu, 2000), the fixation index (Fst) (Weir and Cockerham, 1984), and linkage disequilibrium (LD) (Slatkin, 2008), to detect selection signals in HS lncRNAs and strong DBSs in CEU, CHB, and YRI (the three representative populations of Europeans, Asians, and Africans) (Supplementary Note 4, 5; Supplementary Table 9, 10, 11). Selection signals were detected in several HS lncRNA genes in CEU and CHB but not in YRI, as well as in more DBSs in CEU and CHB than in YRI. These results agree well with findings reported in previous studies, including that fewer selection signals are detected in YRI (Sabeti et al., 2007; Voight et al., 2006). Selection signals in considerable DBSs were detected by multiple tests, indicating the reliability of the analysis. DBSs in the two pigmentation-related genes MC1R and MFSD12 and the two odor reception-related genes OR6C1 and TAS1R3 are notable examples. The dense SNPs, population-specific selection signals, and specific functions of these genes together indicate the ongoing reshaping of gene expression in modern humans.

Young weak DBSs may have greatly promoted recent human evolution

To further examine whether the reshaping has been revised recently, we classified strong and weak DBSs with mostly changed sequence distances into “old” ones, “young” ones, and “others”. DBSs with “Human-Chimp distance>0.034 AND Human-Altai Neanderthals distance=0” were defined as old ones, and DBSs with “Human-Altai Neanderthals distance>0.034 OR Human-Denisovan distance>0.034” were defined as young ones, which demark early and late sequence changes in human evolution. As we identified favored and hitchhiking mutations in 17 human populations (Tang et al., 2022; Tang et al., 2023), we further classified DBSs into six classes - strong old, strong young, strong others, weak old, weak young, weak others, and examined the numbers of favored and hitchhiking mutations in each class (Supplementary Table 16). Favored and hitchhiking mutations are most enriched in the weak young class; we interpret that these DBSs may have considerably promoted recent human evolution.

SNPs in DBSs have cis-effects on gene expression

To verify the above genomic and population genetics analysis, we further examined SNPs in DBSs and their impact on gene expression using transcriptomic data. Based on the Genotype-Tissue Expression (GTEx) data (GTEx Consortium, 2017), we identified expression quantitative trait loci (eQTLs) in DBSs using the widely adopted criteria of MAF≥0.1 and cis-effect size (ES)≥0.5. 1727 SNPs, with MAF ≥0.1 in at least one population and absolute values of ES≥0.5 in at least one tissue, were identified in DBSs in autosomal genes (Supplementary Table 12,13). These eQTLs include 372 “conserved” ones (i.e., also in DBSs in the three archaic humans) and 1020 “novel” ones (i.e., only in DBSs in modern humans). Many conserved eQTLs are in brain tissues and have high derived allele frequencies (DAFs) in all three modern populations. A notable eQTL with high DAFs in all three populations and positive ES≥0.5 in 44 out of 48 GTEx tissues is rs2246577 in the DBS in FN3KRP, whose encoded protein deglycates proteins to restore their function (Table 2). In contrast, many novel eQTLs are in a particular tissue and have population-specific DAFs (Supplementary Note 6).

Next, we performed two analyses to examine how eQTLs in DBSs influence gene expression. First, we computed the expression correlation between HS lncRNAs and target transcripts with DBSs having eQTLs in specific tissues. Most (94%) HS lncRNA-target transcript pairs showed significant expression correlation in tissues in which the eQTLs were identified (|Spearman’s rho| >0.3 and FDR <0.05). Second, we examined whether eQTLs are more enriched in DBSs than in Ensembl-annotated promoters by computing and comparing the density of eQTLs in the two classes of regions. In the comparison, we used promoters as the reference (background) because they contain DBSs. The results indicate that eQTLs are more enriched in DBSs than in promoters (one-sided Mann-Whitney test, p=0.0) (Supplementary Note 6). Thus, population-specific SNPs and tissue-specific eQTLs, on the levels of genome and transcriptome, respectively, suggest that DBSs enable HS lncRNAs to regulate transcription with population- and tissue-specific features.

HS lncRNAs promote brain evolution from archaic to modern humans

To further confirm that SNPs in DBSs have cis-effects on gene expression, we examined the impact of HS lncRNA-DBS on gene expression in the GTEx tissues (GTEx Consortium, 2017). 40 autosomal HS lncRNAs are expressed in at least one tissue (median TPM>0.1) and form 198876 HS lncRNA-target transcript pairs in all tissues. Of these pairs, 45% show a significant expression correlation in specific tissues (Spearman’s |rho|>0.3 and FDR<0.05). To assess the likelihood that these correlations could be obtained by chance, we randomly sampled 10000 pairs of lncRNAs and protein-coding transcripts genome-wide and found that only 2.3% of pairs showed significant expression correlation (Spearman’s |rho|>0.3 and FDR<0.05). Moreover, a higher percentage (56%) of HS lncRNA-target transcript pairs with significant correlation is detected in at least one brain region, indicating more extensive gene expression regulation by HS lncRNAs in the brain (Figure 3A).

The impact of HS lncRNA-DBS interaction on gene expression in GTEx tissues and organs.

(A) The distribution of the percentage of HS lncRNA-target transcript pairs with correlated expression across GTEx tissues and organs. Higher percentages of correlated pairs are in brain regions than in other tissues and organs. (B) The distribution of significantly changed DBSs (in terms of sequence distance) in HS lncRNA-target transcript pairs across GTEx tissues and organs between archaic and modern humans. Orange, red, and dark red indicate significant changes from Denisovans (D), Altai Neanderthals and Denisovans (AD), and all three archaic humans (ADV). DBSs in HS lncRNA-target transcript pairs with correlated expression in seven brain regions (in dark red) have changed significantly and consistently since the Altai Neanderthals, Denisovans, and Vindija Neanderthals (one-sided two-sample Kolmogorov-Smirnov test, significant changes determined by FDR <0.001).

To obtain supporting evidence on the transcriptome level, we analyzed nine experimental datasets (Supplementary Note 7). First, we analyzed two datasets of epigenetic studies: one examining H3K27ac and H3K4me2 profiling in human, macaque, and mouse corticogenesis, and the other examining gene expression and H3K27ac modification in eight brain regions in humans and four other primates (Reilly et al., 2015; Xu et al., 2018). Compared with the genome-wide background, 84% and 73% of genes in the two datasets have DBSs for HS lncRNAs, indicating significantly higher enrichment in HS lncRNA targets (p=1.21e-21 and 1.2e-56, two-sided Fisher’s exact test). When using gene expression in chimpanzees as the control, 1851 genes show human-specific transcriptomic differences in one or multiple brain regions, whereas only 240 genes show chimpanzee-specific transcriptomic differences. Second, we analyzed two datasets of PsychENCODE studies, one examining spatiotemporally differentially expressed genes and spatiotemporally differentially methylated sites in 16 brain regions, and the other examining the spatiotemporal transcriptomic divergence across human and macaque brain development (Li et al., 2018; Zhu et al., 2018). In the first dataset, 65 HS lncRNAs are expressed in all 16 brain regions, 109 transcripts have spatiotemporally differentially methylated sites in their promoters, and 56 of the 109 transcripts have DBSs for HS lncRNAs. In the second dataset, 8951 genes show differential expression between the human and macaque brains, and 72% of differentially expressed genes in the human brain have DBSs for HS lncRNAs. Thus, both datasets show significant enrichment for HS lncRNA regulation. Third, three studies identified genes critically regulating cortical expansion (Florio et al., 2018; Johnson et al., 2018; Suzuki et al., 2018). Of the 40 reported protein-coding genes, 29 have DBSs of HS lncRNAs in promoter regions (Supplementary Table 14). Thus, these genes are enriched with DBSs of HS lncRNAs compared to the genome-wide background (p<0.01, two-sided Fisher’s exact test). Fourth, we analyzed two datasets of brain organoid studies. By establishing and comparing cerebral organoids between humans, chimpanzees, and macaques, Pollen et al. identified 261 human-specific gene expression changes (Pollen et al., 2019). By fusing human and chimpanzee iPS cells and differentiating the hybrid iPS cells into hybrid cortical spheroids, Agoglia et al. generated a panel of tetraploid human-chimpanzee hybrid iPS cells and identified thousands of genes with divergent expression between humans and chimpanzees (Agoglia et al., 2021). We found that 261 and 1102 genes in the two datasets are enriched with DBSs of HS lncRNAs compared to the genome-wide background (p=1.2e-16 and 3.4e-74).

Finally, we examined the evolution of the impact of HS lncRNAs and their DBSs on gene expression further using the GTEx data (the transcript expression matrices). For each DBS in an HS lncRNA-target transcript pair that shows correlated expression in a GTEx tissue, we computed the sequence distances of this DBS from the HS lncRNA-target transcript pair in modern humans to the pairs in the three archaic humans. Then, we compared the distribution of DBS sequence distances in each tissue with the distribution of DBS sequence distances in all tissues as the background (one-sided two-sample Kolmogorov-Smirnov test). DBSs in HS lncRNA-target transcript pairs with correlated expression in brain regions have significantly changed sequence distances since the Altai Neanderthals and Denisovans (Figure 3B), suggesting that gene expression regulation by HS lncRNAs in the brain has undergone significant evolution recently (Ponce de Leon et al., 2021). Results from other GTEx tissues, including Kidney-Cortex, Ovary, and Skin-Sun Expressed (Figure 3B), also suggest continuous influences of reshaped gene expression on human evolution.

The above results prompted us to examine whether HS lncRNAs may have contributed more significantly to human evolution than HS TFs (Supplementary Note 9). Based on the “hg38-panTro6” gene sets reported by Kirilenko et al. and the human TF lists reported by previous studies and used in the SCENIC package (Bahrami et al., 2015; Kirilenko et al., 2023; Lambert et al., 2018), we identified five potential HS TFs. Then, we predicted their DBSs in the 5000 bp promoter regions of the 179128

Ensembl-annotated transcripts (release 79) using FIMO and CellOracle (Grant et al., 2011; Kamimoto et al., 2023), identified counterparts of these DBSs in archaic humans and chimpanzees, computed sequence distances of DBSs from modern humans to archaic humans and chimpanzees (Supplementary Table 18), computed the Pearson correlation of HS TFs and target transcripts across GTEx tissues, and computed sequence distances of DBSs in HS TF-target transcript pairs across GTEx tissues from modern humans to archaic humans. Highly correlated HS TF-target transcript pairs are distributed across multiple tissues, rather than being confined to the brain; additionally, significantly changed DBSs do not densely occur in the brain (Supplementary Figs. 25 and 26). These results further support that HS lncRNAs have promoted human evolution by specifically regulating gene expression.

HS lncRNAs mediate human-specific correlated gene expression in the brain

To explore the evidence indicating that the high percentage of correlated gene expression in the brain (Figure 3A) is human-specific, we used the GTEx data from the human frontal cortex (BA9) and anterior cingulate cortex (BA24) (n=101 and 83, respectively) (Consortium et al., 2017), the transcriptomic data for the same brain regions in macaques (n=22 and 25, respectively) (Zhu et al., 2018), and the eGRAM program we developed recently for examining transcriptional regulation modules, to identify gene modules. In the human brain, HS lncRNAs’ target genes form distinct modules, characterized by rich correlation relationships (Pearson’s r>0.8) and enrichment in neurodevelopment-related KEGG pathways (hypergeometric distribution test, FDR<0.05). In contrast, in the same macaque brain regions, the orthologues of HS lncRNAs’ target genes do not have these features (Figure 4; Supplementary Figure 24; Supplementary Table 17). The distinct differences in gene modules in these brain regions support that the reshaped gene expression by HS lncRNAs is human-specific.

Human-specifically reshaped gene expression by HS lncRNAs in the frontal cortex (BA9).

(A) Genes expressed in the human frontal cortex are enriched for HS lncRNAs’ target genes and neurodevelopment-related pathways. Squares, dots, and colors indicate HS lncRNAs, gene modules (Module_1 and Module_2 are illustrated), and enriched KEGG pathways, respectively. (B) Comparison of modules and genes in humans (indicated by H) and macaques (indicated by M). In each pair of modules, green and blue dots denote human genes and their orthologues, and lines between dots indicate correlated expression. Note that many orthologous genes in macaques are not in the modules but displayed in the corresponding positions, and correlated expression is more prominent in humans than in macaques.

Discussion

The trivial genomic differences but substantial phenotypic and behavioral differences between modern humans, archaic humans, and apes make the question of what genomic differences critically determine modern humans a profound one. This question has been addressed by numerous studies employing various methods. Previous studies notably examined new protein-coding genes and HARs, reporting that HS protein-coding genes promote rapid neocortex expansion (Fiddes et al., 2018; Florio et al., 2018; Pinson et al., 2022) and that HARs are significantly enriched in domains important for human-specific 3D genome organization (Keough et al., 2023). Many studies also examined the functions of lncRNAs in humans and mice. However, HS lncRNAs and lncRNA DBSs have been overlooked. Since gene expression can be substantially rewired by TFs and TF DBSs (Johnson, 2017), it is sensible to postulate that HS lncRNAs and their DBSs may have greatly reshaped gene expression. Verifying this postulation needs examination of cross-species genomic and transcriptomic data. This study aimed to verify this postulation by identifying HS lncRNAs, their DBSs, and the counterparts of these DBSs in modern humans, ancient humans, and chimpanzees. We used multiple methods to analyze diverse data, and our results suggest that HS lncRNAs have greatly reshaped (or even rewired) gene expression in humans.

As DBS prediction is the basis of the study, we used multiple methods, including DBD KO using CRISPR/Cas9, following the analysis of differential expression of target genes, to validate predicted DBSs. When running programs including Infernal, LongTarget, and g:Profiler and performing statistical tests that detect selection signals, we used default parameters that suit most situations and widely used significance levels. When identifying those of HS lncRNAs’ target genes that probably most critically differentiate humans from chimpanzees, the top 20% (4248) of genes were used whose DBSs in humans or the counterparts of these DBSs in chimpanzees have the largest binding affinity or sequence distances. In support that these parameters, significance levels, and cutoffs are reasonable, the numbers of genes we obtained that differentiate modern humans from archaic humans and chimpanzees agree with the phylogenetic relationships revealed by previous studies (Prufer et al., 2017), and are approximate to the number (4377) of differentially expressed genes between human-only and chimpanzee-only iPSC lines due to trans-acting differences between humans and chimpanzees (Song et al., 2021). Importantly, since weak DBSs may be less reliable than strong ones, we classified DBSs into strong/weak ones and old/young ones. We found that different classes of DBSs have distinct evolutionary and functional features, which is generally sensible given that human evolution has undergone different big events. Due to the nature of data analysis, it is challenging to eliminate bias, false positives, false negatives, and the impact of varying parameter values. Other parameters and cutoff values may generate somewhat (but not significantly) different results (e.g., a DBS distance cutoff=0.035 would determine 4157 top target genes). Not all detected signals reliably indicate positive selection, but the main conclusions are based on binding affinity, sequence distance, and SNP number, instead of selection signals in DBSs (Tables 1, 2). Genes have different transcript/isoform numbers, but this may unlikely cause confounding effects in downstream analyses such as overrepresentation analysis (Figure 2) and transcriptomic data analysis (Figure 3) because the input to g:Profiler is a non-redundant gene list and each RNA-seq file normally has just one record for each gene, even if some genes may have one but others have multiple DBSs (before their one or multiple transcripts). While a gene with more DBSs/transcripts may have a higher chance to evolve a large DBS sequence distance, this more likely suggests that the transcript with a large DBS sequence distance has become more/less important than others.

We have a few notes on several findings. First, Neanderthal-inherited sequences have measurable impacts on gene expression in modern humans, but the impacts are least detectable in the brain (McCoy et al., 2017). Here, we find that HS lncRNAs and their DBSs have more significantly influenced gene expression in the brain than in other tissues. Second, the regulation of many genes by HS lncRNAs may have undergone rapid evolution, notable examples including those making humans adapt to high sugar intake. Third, intensively regulated genes are enriched in different functions at different periods of human evolution, as evidenced by genes with strong/weak and young/old DBSs and with DBSs containing “old” and “new” SNPs. Notable details include that DBSs containing “old” and “new” SNPs are in genes regulating neural development and glucose metabolism. Fourth, newly emerged regulations (i.e., target genes with young weak DBSs) may have significantly promoted recent human evolution.

The above findings also raise new questions. First, a high percentage (about 85%) of protein-coding regions in the mouse and human genomes is identical, but many mouse lncRNAs are rodent-specific (Yue et al., 2014). Thus, two questions arise: how these lncRNAs specifically rewire gene expression in mice and to what extent human- and mouse-specific lncRNAs cause the cross-species transcriptional regulation differences (Breschi et al., 2017; Hodge et al., 2019; Zhu et al., 2018). Second, whether mice and other mammals have such species-specific lncRNAs as RP11-423H2.3 that regulate the expression of substantial genes. Third, whether lineage- and species-specific lncRNAs would make many evolutionary novelties preordained.

Material and methods

Data sources

The sequences of human (GRCh38/hg38, GRCh37/hg19) and chimpanzee genomes (panTro5) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). Three high-quality archaic human genomes were obtained from the Max Planck Institute for Evolutionary Anthropology (https://www.eva.mpg.de/genetics/index.html), which include an Altai Neanderthal that lived approximately 122 thousand years ago (kya) (Prüfer et al., 2013), a Denisovan (an Asian relative of Neanderthals) that lived approximately 72 kya (Meyer et al., 2012), and a Vindija Neanderthal that lived approximately 52 kya (Prufer et al., 2017). The ancestral sequences for the human genome (GRCh37) (which were generated upon the EPO alignments of six primate species human, chimpanzee, gorilla, orangutan, macaque, and marmoset) were downloaded from the EBI website (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignm ents/). The SNP data of modern humans were obtained from the 1000 Genomes Project (phase I) (1000 Genomes Project Consortium et al., 2012). The three modern human populations are CEU (Utah residents with Northern and Western European ancestry), CHB (Han Chinese in Beijing, China), and YRI (Yoruba in Ibadan, Nigeria), which contain 100, 99, and 97 individuals, respectively. The genetic map files of modern humans were downloaded from the 1000 Genomes Project website (ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/technical/working/20110106_recombination_hotspots/).

The 1000 Genomes phased genotype data in VCF format were downloaded from the International Genome Sample Resource website (http://www.internationalgenome.org/). Annotated human genes and transcripts were obtained from the Ensembl website (http://www.Ensembl.org). FASTA-format nucleotide sequences were converted from VCFs using the vcf-consensus program in the VCFtools package when necessary (Danecek et al., 2011). Data in the GeneCards Human Gene Database (www.genecards.org) were used to annotate genes and transcripts.

RNA-sequencing (RNA-seq) data (transcripts per million [TPM]) and eQTL data of human tissues were obtained from the Genotype-Tissue Expression Project (GTEx, v8) website (https://gtexportal.org/) (GTEx Consortium, 2017). Histone modification and DNA methylation signals in multiple cell lines and ENCODE Candidate Cis-Regulatory Elements (cCRE) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). The genome-wide DNA binding sites of lncRNA NEAT1, MALAT1, and MEG3 were obtained from experimental studies (Mondal et al., 2015; West et al., 2014). The predicted and experimentally detected DBSs in target genes of MALAT1, NEAT1, and MEG3 are given in the supplementary table (Supplementary Table 15).

Identifying HS lncRNA genes

We used the Infernal program (Nawrocki et al., 2009), which searches orthologous RNA sequences upon sequence and structure alignment, to identify orthologous exons in 16 mammalian genomes for each exon in each of the 13562 GENCODE (v18)-annotated human lncRNA genes (Lin et al., 2019). The 16 mammals were chimpanzee (CSAC 2.1.4/panTro4), macaque (BGI CR_1.0/rheMac3), marmoset (WUGSC 3.2/calJac3), tarsier (Broad/tarSyr1), mouse lemur (Broad/micMur1), tree shrew (Broad/tupBel1), mouse (GRCm38/mm10), rat (Baylor3.4/rn4, RGSC6.0/rn6), guinea pig (Broad/cavPor3), rabbit (Broad/oryCun2), dog (Broad CanFam3.1/canFam3), cow (Baylor Btau_4.6.1/bosTau7), elephant(Broad/loxAfr3), hedgehog (EriEur2.0/eriEur2), opossum (Broad/monDom5), and platypus (WUGSC 5.0.1/ornAna1) (http://genome.UCSC.edu). If the number of orthologous exons of a human lncRNA gene in a genome exceeded half the exon number of the human lncRNA gene, these orthologous exons were assumed to form an orthologous lncRNA gene. If a human lncRNA gene had no orthologous gene in all of the 16 mammals, it was assumed to be human-specific.

Identifying DBSs of HS lncRNAs

LncRNAs bind to DNA sequences by forming RNA:DNA triplexes. Each triplex comprises triplex-forming oligonucleotides (TFO) in the lncRNA and a triplex-targeting site (TTS) in the DNA sequence. We used the LongTarget program to predict HS lncRNAs’ DNA binding domains (DBD) and binding sites (DBS) with the default parameters (Ruleset = all, T.T. penalty = -1000, CC penalty = 0, Offset = 15, Identity ≥ 60, Nt ≥ 50) (Lin et al., 2019; Wen et al., 2022). The LongTarget program predicts DBDs and DBSs simultaneously, where a DBD comprises a set of densely overlapping TFOs and a DBS comprises a set of densely overlapping TTSs. For each HS lncRNA, we predicted its DBSs in the 5000 bp promoter regions (−3500 bp upstream and +1500 bp downstream the transcription start site) of the 179128 Ensembl-annotated transcripts (release 79). For each DBS, its binding affinity is the product of DBS length and the averaged Identity score of all TTSs (the Identity score is the percentage of paired nucleotides). Strong and weak DBSs were classified based on binding affinity ≥60 and <60. A transcript whose promoter region contains a strong DBS of an HS lncRNA was assumed to be a target transcript of the HS lncRNA, and the gene containing this transcript was assumed to be a target gene of the HS lncRNA. As the 1000 Genomes Project (phase I) data and the archaic human genomes are based on GRCh37/hg19, the coordinates of DBSs were converted from GRCh38/hg38 to GRCh37/hg19 using the liftover program from the UCSC Genome Browser (Kuhn et al., 2013).

Experimentally validating DBS prediction

A 157 bp sequence (chr17:80252565-80252721, hg19) containing the DBD of RP13-516M14.1, a 202 bp sequence (chr1:113392603-113392804, hg19) containing the DBD of RP11-426L16.8, and a 198 bp sequence (chr17:19460524-19460721, hg19) containing the DBD of SNORA59B, were knocked out in the HeLa cell line, RKO cell line, and SK-MES-1 cell line, respectively. Two sequences (chr1:156643524-156643684, chr10:52445649-52445740, hg38) containing the DBD of two wrongly transcribed noncoding sequences were knocked out in the HCT-116 and A549 cell lines, respectively. The seven knockouts were performed by UBIGENE, Guangzhou, China (http://www.ubigene.com) using CRISPR-UTM, a revised version of CRISPR/Cas9 technology. Before and after the seven DBD knockouts, RNA sequencing (RNA-seq) was performed by Novogene, Beijing, China (https://cn.novogene.com) and HaploX, Shenzhen, China (https://www.haplox.cn/). The reads were aligned to the human GRCh38 genome using the Hiasat2 program (Kim et al., 2019), and the resulting SAM files were converted to BAM files using Samtools (Li et al., 2009). The Stringtie program was used to quantify gene expression levels (Pertea et al., 2015). Fold change of gene expression was computed using the edgeR package (Robinson et al., 2010), and significant up- and down-regulation of target genes after DBD knockout was determined upon |log2(fold change)| > 1 with FDR < 0.1.

Genome-wide DBSs of NEAT1, MALAT1, and MEG3 were experimentally detected (Mondal et al., 2015; West et al., 2014). We also used these data to validate DBS prediction by predicting DBSs of the three lncRNAs and checking the overlap between predicted and experimentally detected DBSs (Supplementary Table 15).

Mapping DBSs of HS lncRNAs in the chimpanzee and archaic human genomes

We used the liftover program from the UCSC Genome Browser (http://genome.UCSC.edu) to map loci of DBSs in the human genome hg38 to the chimpanzee genome Pan_tro 3.0 (panTro5). The mapping results were examined by checking the human-chimpanzee pairwise alignment (hg38 vs panTro5) in the UCSC Genome Browser (http://genome.UCSC.edu). 97.81% (100403/102651) of DBSs were mapped, and the 2248 unmapped DBSs in 429 genes were considered human-specific. We used vcf-consensus in the VCFtools package to extract the DBSs of HS lncRNAs from the VCF files of Altai Neanderthals, Denisovans, and Vindija Neanderthals. The variant with the highest quality score was selected whenever multiple variant calls were observed at a given locus. The obtained DBS sequences in chimpanzees and three archaic humans are called counterparts of DBSs in these genomes.

Estimating sequence distances of DBSs between different genomes

We first aligned DBS sequences in the genomes of humans, chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals using the MAFFT7 program for measuring sequence distances from modern humans to chimpanzees and archaic humans (Katoh and Standley, 2013). We then computed sequence distances using the dnadist program with the Kimura 2-parameter model in the PHYLIP (3.6) package (http://evolution.genetics.washington.edu/phylip.html) and the Tamura-Nei model in the MEGA7 package (Kumar et al., 2016). The two methods generated equivalent results. The largest distance between DBSs in humans and their counterparts in chimpanzees is 5.383. Since 2248 DBSs in 429 human genes do not have counterparts in chimpanzees, we assumed that these DBSs have sequence distances of 10.0 between humans and chimpanzees. The relationship between cross-species sequence distances and cross-species phenotypic distances is little known. Based on the report that 5984 genes were identified differentially expressed between human-only and chimpanzee-only iPSC lines (including 4377 whose differential expression was due to cross-species trans-acting differences) (Song et al., 2021) and on the report that “the Denisovan genome thus tend to be older, than the Altai Neandertal genome” (Prufer et al., 2017), we chose 0.034 as the DBS sequence distance cutoff for identifying “mostly changed DBSs” from modern humans to chimpanzees and archaic humans. 0.034 determined 4248 genes (the top 20%) in chimpanzees, and 1256, 2514, and 134 genes in Altai Neanderthals, Denisovans, and Vindija Neanderthals, respectively, which may critically characterize humans from chimpanzees and three archaic humans. If cutoff=0.035, since 91 genes have DBS sequence distance=0.034, 4248-91=4157 genes, this small difference in gene number (4248 vs 4157) would not significantly influence overrepresentation analysis (Figure 2).

We determined human ancestral sequences of DBSs based on the human ancestor sequences from the EBI website, which were generated based on the EPO alignments of six primate species. We used the above-mentioned methods to calculate the sequence distances DBSs from the human ancestor to chimpanzees, archaic humans, and modern humans. We found that when the human-chimpanzee ancestral sequence has the ancestral sequence (which means the inference of ancestral allele is of high confidence), DBS distances from the human ancestor to modern humans are larger than to archaic humans, but this situation only accounts for about 63.8%. For many DBSs, the distances from the human ancestor to modern humans are smaller than to archaic humans (especially Neanderthals and Denisovans) and even to chimpanzees. This defect may be caused by the absence of archaic humans in building the human ancestral sequence.

Detecting positive selection signals in HS lncRNA genes and DBSs

We used multiple tests to detect positive selection signals in HS lncRNA genes. First, we used the XP-CLR test (parameters = -w1 0.001 300 100 -p0 0.95, window size = 0.1 cM, grid size = 100 bp) to perform six pairwise genome-wide scans (i.e., CEU-CHB, CEU-YRI, CHB-CEU, CHB-YRI, YRI-CEU, and YRI-CHB) (Chen et al., 2010). The upper 1% of scores across the entire genome in each pairwise scan was 34.6 in the CEU-YRI scan, 16.8 in the CEU-CHB scan, 45.0 in the CHB-YRI scan, 26.9 in the CHB-CEU scan, 14.1 in the YRI-CEU scan, and 14.1 in the YRI-CHB scan. These scores were used as the thresholds of positive selection signals in these populations. Second, we used the iSAFE program to scan each genomic region containing an HS lncRNA gene and its 500-kb upstream and downstream sequences (Akbari et al., 2018). Strongly selected loci were detected only in CEU and CHB. Third, we used the VCFtools program to calculate Tajima’s D values for each HS lncRNA gene in CEU, CHB, and YRI (Danecek et al., 2011). The calculation was performed using a 1500-bp non-overlapping sliding window because the lengths of these genes exceed 1500 bp. To generate a background reference for assessing the significant increase or decrease of Tajima’s D for HS lncRNA genes in a population, we calculated Tajima’s D values across the whole genome using a sliding window of 1500 bp. As the values of Tajima’s D were compared with the background reference, significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively, rather than population demography dynamics (Tajima, 1989). Fourth, we used the integrated Fst to detect positive selection signals in HS lncRNA genes. The Fst of each HS lncRNA gene was computed for three comparisons, i.e., CEU-YRI, CHB-YRI, and CHB-CEU. Extreme Fst values of SNPs were detected in HS lncRNA genes in the comparisons of CEU-YRI and CHB-YRI. Since allele frequencies for different loci vary across a genome and genetic drift may have different effects at different loci, we used a sliding window of 1500 to compare Fst values of HS lncRNA genes with the genome-wide background. Extreme Fst values indicate positive selection. Finally, we applied linkage disequilibrium (LD) analysis to each HS lncRNA gene. We computed the pairwise LD (r²) in CEU, CHB, and YRI for common SNPs (with minimal allele frequency (MAF) ≥ 0.05 in at least one population) in HS lncRNA genes and DBSs using the PLINK program (Purcell et al., 2007). Significantly increased LD was detected in SNPs in HS lncRNA genes in CEU and CHB. The LD patterns were portrayed using the Haploview program (Barrett et al., 2005).

Next, we used the above tests to detect positive selection signals in DBSs. First, the 100 bp grid size of the XP-CLR test also allowed the detection of selection signals in DBSs. Second, we performed Tajima’s D and Fay-Wu’s H tests (Fay and Wu, 2000; Tajima, 1989). We calculated Tajima’s D values for each DBS in CEU, CHB, and YRI using the VCFtools program (Danecek et al., 2011), with a sliding window of 147 bp (the mean length of strong DBSs). To generate a background reference for judging the significant increase or decrease of Tajima’s D in a population, we calculated Tajima’s D values across the whole genome using the same sliding window. As values of Tajima’s D were compared with the background reference, a significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively. Fay-Wu’s H values were calculated similarly using the VariScan program (Vilella et al., 2005). Calculating Fay-Wu’s H demands the ancestral sequences as the outgroup. We extracted the ancestral sequences of DBSs from the human ancestral sequence, which was generated upon the EPO alignments of six primate species (human, chimpanzee, gorilla, orangutan, macaque, and marmoset) (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignm ents/). Third, we computed the Fst to measure the frequency differences of alleles in DBSs between populations. For the CEU-YRI, CHB-YRI, and CHB-CEU pairwise comparisons, we used the revised VCFtools program to compute the weighted Fst for all SNPs in each DBS (Weir and Cockerham, 1984). Fourth, we integrated the weighted Fst values in the three populations into an “integrated Fst” which indicated whether the DBS locus was under selection in a certain population (Nielsen and Slatkin, 2013). We used sliding windows of 147 bp for comparing Fst values of DBSs with the genome-wide background. We empirically denoted the values of integrated Fst in the upper 10% of scores across the entire genome as statistically significant. To determine positive selection more reliably, we used Tajima’s D and the integrated Fst to jointly determine if a DBS was under positive selection in a population. The thresholds that determined the upper 10% of Tajima’s D values across the entire genome in CEU, CHB, and YRI were -0.97, -0.96, and -0.97, respectively, and the threshold that determined the upper 10% of integrated Fst values across the entire genome in the three populations was 0.22. For example, a DBS was assumed to be under positive selection in CEU if (a) the DBS had a Tajima’s D<-0.97 in CEU and Tajima’s D>0.0 in the two other populations and (b) the DBS had an integrated Fst>0.22. Analogously, a DBS was assumed to be under positive selection in both CEU and CHB if the DBS had a Tajima’s D <-0.97 in CEU, <-0.96 in CHB, and >0.0 in YRI, and had an integrated Fst>0.22.

Functional enrichment analysis of genes

We used the g:Profiler program (with the parameter settings: Organism=Homo sapiens, Ordered query = No, Significance threshold = Benjamini-Hochberg FDR, User threshold = 0.05, 50 < terms size < 1000) and the Gene Ontology (GO) database to perform over-representation analysis (Raudvere et al., 2019). This analysis is used to determine which pre-defined gene sets (GO terms) are more prevalent (over-represented) in a list of “interesting” genes than would be expected by chance. The lists of genes included genes with the strongest DBSs in humans, genes with the weakest DBSs in humans, and genes with mostly changed DBSs in chimpanzees and three archaic humans. To choose genes with strong DBSs, we ranked genes in descending order based on binding affinity and chose the top 2000 genes. To choose genes with weak DBSs, we first ranked genes in ascending order based on binding affinity, then ranked genes in descending order based on the number of regulatory HS lncRNAs, and finally chose the top 2000 genes with binding affinity ≤40 and regulatory HS lncRNAs ≥5. We used the distance threshold of 0.034 to select genes with mostly changed DBSs. Genes with mostly changed DBSs in chimpanzees include the 429 genes in which 2248 DBSs lack counterparts. 4248, 1256, 2514, and 134 genes in chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals have DBSs with distance ≥0.034. To make the statistical significance comparable across chimpanzees, Altai Neanderthals, and Denisovans, we used the same number of genes (the top 1256 genes whose DBSs have the largest DBS distances) as the input to the g:Profiler program.

Analysing SNP frequencies in human populations

The frequency of common SNPs (MAF ≥0.05) in DBSs in the three modern human populations was computed using the VCFtools package (Danecek et al., 2011). The ancestral/derived states of SNPs were inferred from the human ancestor sequences and were used to determine derived allele frequency.

Analysing the cis-effect of SNPs in DBSs on target gene expression

SNPs with MAF >0.1 in DBSs in any of the three modern human populations and absolute values of cis-effect size >0.5 (FDR <0.05) in any of the GTEx tissues were examined for an influence on the expression of the target genes (GTEx Consortium, 2017). SNPs that are eQTLs in the GTEx tissues and have biased derived allele frequencies (DAF) in the three modern human populations were examined to estimate whether the eQTL is population-specific.

Examining the tissue-specific impact of HS lncRNA-regulated gene expression

First, we examined the expression of HS lncRNA genes across the GTEx tissues. HS lncRNA genes with a median TPM value >0.1 in a tissue were considered robustly expressed in that tissue. Upon this criterion, 40 HS lncRNA genes were expressed in at least one tissue and were used to examine the impact of HS lncRNA regulation on gene and transcript expression (other HS lncRNAs may function in the cytoplasm). Since an HS lncRNA gene may have multiple transcripts, we adopted the transcript that contains the predicted DBD and has the largest TPM value as the representative transcript of the HS lncRNA. We calculated the pairwise Spearman’s correlation coefficient between the expression of an HS lncRNA (the representative transcript) and the expression of each of its target transcripts using the scipy.stats.spearmanr program in the scipy package. The expression of an HS lncRNA gene and a target transcript was considered to be significantly correlated if the |Spearman’s rho|>0.3, with Benjamini-Hochberg FDR<0.05. We examined the percentage distribution of correlated HS lncRNA-target transcript pairs across GTEx tissues and organs (Figure 3A).

We examined all GTEx tissues to determine in which tissues HS lncRNA-regulated gene expression may have undergone significant changes from archaic to modern humans. GTEx data have both gene expression matrices and transcript expression matrices; we used the latter to examine the change of HS lncRNA-target transcript pairs from modern humans to archaic humans. If a pair of HS lncRNA and target transcript is robustly expressed in a tissue and their expression shows a significant correlation (|Spearman’s rho|>0.3, with Benjamini-Hochberg FDR<0.05) in the tissue, we computed the sequence distance of the HS lncRNA’s DBS in the transcript from the three archaic humans to modern humans. We compared the sequence distances of all DBSs in each tissue with the sequence distances of all DBSs in all GTEx tissues (as the background). A one-sided two-sample Kolmogorov-Smirnov test was used to examine whether the sequence distances of all DBSs in a specific tissue deviate from the background distribution (which reflects the “neutral evolution” of gene expression). For each tissue, if the test result was Benjamini-Hochberg FDR < 0.001, the tissue was considered to have significantly changed gene expression regulated by the HS lncRNA. We used different colors to mark tissues with significantly changed gene expression regulation since Altai Neanderthals, Denisovans, and Vindija Neanderthals, and used “D”, “A.D.”, and “ADV” to indicate changes since Denisovans, since Altai Neanderthals and Denisovans, and since Altai Neanderthals, Denisovans, and Vindija Neanderthals, respectively (Figure 3B).

Examining enrichment of favored and hitchhiking mutations in DBSs

Using the deep learning network DeepFavored, which integrates multiple statistical tests for identifying favored mutations (Tang et al., 2022), we identified 13339 favored mutations and 244098 hitchhiking mutations in 17 human populations (Tang et al., 2023). In this study, we classified DBSs in two ways: into strong ones (affinity>60) and weak ones (36<affinity<60), and into old ones (Human-Chimp distance>0.034 AND Human-Altai Neanderthals distance=0), young ones (Human-Altai Neanderthals distance>0.034 OR Human-Denisovan distance>0.034), and others. We then examined the number of favored and hitchhiking mutations in each class of DBSs. The weak young DBSs have the largest proportion of favored and hitchhiking mutations.

Identifying and analyzing transcriptional regulatory modules

Most clustering algorithms classify genes into disjoint modules based on expression correlation without considering regulatory relationships (Saelens et al., 2018). The GRAM program identifies gene regulatory modules based on correlation and TF-TFBS binding (Bar-Joseph et al., 2003). LncRNAs transcriptionally regulate genes based on lncRNA-DBS binding and correlate gene expression. We developed the eGRAM program to identify gene modules based on correlated expression, TF-TFBS interaction, and lncRNA-DBS interaction. In this study, we used eGRAM to identify gene modules in the same regions of the human and macaque brains and enriched KEGG pathways using reported RNA-seq datasets (n=101 and 83 for frontal cortex and n=22 and 25 for anterior cingulate cortex) (GTEx Consortium, 2017; Zhu et al., 2018). The default parameters, DBS binding affinity=60, Pearson correlation=0.5, module size=50, and FDR=0.01 (hypergeometric distribution test), were used. The key steps of the program are as follows. (a1) Identify each lncRNA’s correlated lncRNAs, which may form a set of co-regulators. (a2, optional) Identify each TF’s correlated TFs, which may form a set of co-regulators. (b1) Compute the correlation between each lncRNA and all genes. (b2, optional) Compute the correlation between each TF and all genes. (c1) Identify each lncRNA’s target genes. (c2, optional) Identify each TF’s target genes. (d1) Identify each lncRNA set’s target module upon correlation and targeting relationships. (d2, optional) Identify each TF set’s target module upon correlation and targeting relationships. (e) Check whether TFs’ modules contain lncRNAs’ targets and whether lncRNAs’ modules contain TFs’ targets, which reveal genes co-regulated by TFs and lncRNAs and genes independently regulated by TFs and ncRNAs. (f) Performs pathway enrichment analysis for all modules.

The analysis of HS TFs and their DBSs

Kirilenko et al recently identified orthologous genes in hundreds of placental mammals and birds and organized genes into pairwise datasets using humans and mice as the references (e.g., “hg38-panTro6”, “hg38-mm10”, and “mm10-hg38”) (Kirilenko et al., 2023). Based on the “many2zero” and “one2zero” gene lists (which contain 0 and 147 genes) in the hg38-panTro6 dataset, which indicate the multiple human genes and the one human gene that have no orthologues in chimpanzees, we identified HS protein-coding genes. Further, based on three human TF lists reported by two studies and used in the SCENIC package (Bahrami et al., 2015; Lambert et al., 2018), we identified HS TFs. Based on the JASPAR database and using the CellOracle program (with default parameters), we predicted DBSs of HS TFs. Then, we repeated the steps of our HS lncRNA analyses (Supplementary Note 9), including computing sequence distances of DBSs and examining the impact of HF TF-target transcript pairs on gene expression in GTEx tissues and organs.

eLife Assessment

This valuable study uses tools of population and functional genomics to examine long non-coding RNAs (lncRNAs) in the context of human evolution. Analyses of computationally predicted human-specific lncRNAs and their genomic targets lead to the development of hypotheses regarding the potential roles of these genetic elements in human biology. The conclusions regarding evolutionary acceleration and adaptation, however, only incompletely take data and literature on human/chimpanzee genetics and functional genomics into account.

Reviewer #2 (Public review):

In this valuable manuscript, Lin et al attempt to examine the role of long non coding RNAs (lncRNAs) in human evolution, through a set of population genetics and functional genomics analyses that leverage existing datasets and tools. Although the methods are incomplete and at times inadequate, the results nonetheless point towards a possible contribution of long non coding RNAs to shaping humans, and suggest clear directions for future, more rigorous study.

Comments on revisions:

I thank the authors for their revision and changes in response to previous rounds of comments. As before, I appreciate the changes made in response to my comments, and I think everyone is approaching this in the spirit of arriving at the best possible manuscript, but we still have some deep disagreements on the nature of the relevant statistical approach and defining adequate controls. I highlight a couple of places that I think are particularly relevant, but note that given the authors disagree with my interpretation, they should feel free to not respond!

(1) On the subject of the 0.034 threshold, I had previously stated:

"I do not agree with the rationale for this claim, and do not agree that it supports the cutoff of 0.034 used below."

In their reply to me, the authors state:

"What we need is a gene number, which (a) indicates genes that effectively differentiate humans from chimpanzees, (b) can be used to set a DBS sequence distance cutoff. Since this study is the first to systematically examine DBSs in humans and chimpanzees, we must estimate this gene number based on studies that identify differentially expressed genes in humans and chimpanzees. We choose Song et al. 2021 (Song et al. Genetic studies of human-chimpanzee divergence using stem cell fusions. PNAS 2021), which identified 5984 differentially expressed genes, including 4377 genes whose differential expression is due to trans-acting differences between humans and chimpanzeees. To the best of our knowledge, this is the only published data on trans-acting differences between humans and chimpanzeees, and most HS lncRNAs and their DBSs/targets have trans-acting relationships (see Supplementary Table 2). Based on these numbers, we chose a DBS sequence distance cutoff of 0.034, which corresponds to 4248 genes (the top 20%), slightly fewer than 4377."

I have some notes here. First, Agoglia et al, Nature, 2021, also examined the nature of cis vs trans regulatory differences between human and chimps using a very similar set up to Song et al; their Supplementary Table 4 enables the discovery of genes with cis vs trans effects although admittedly this is less straightforward than the Song et al data. Second, I can't actually tell how the 4377 number is arrived at. From Song et al, "Of 4,671 genes with regulatory changes between human-only and chimpanzee-only iPSC lines, 44.4% (2,073 genes) were regulated primarily in cis, 31.4% (1,465 genes) were regulated primarily in trans, and the remaining 1,133 genes were regulated both in cis and in trans (Fig. 2C). This final category was further broken down into a cis+trans category (cis- and trans-regulatory changes acting in the same direction) and a cis-trans category (cis- and trans-regulatory changes acting in opposite directions)." Even when combining trans-only and cis&trans genes that gives 2,598 genes with evidence for some trans regulation. I cannot find 4,377 in the main text of the Song et al paper.

Elsewhere in their response, the authors respond to my comment that 0.034 is an arbitrary threshold by repeating the analyses using a cutoff of 0.035. I appreciate the sentiment here, but I would not expect this to make any great difference, given how similar those numbers are! A better approach, and what I had in mind when I mentioned this, would be to test multiple thresholds, ranging from, eg, 0.05 to 0.01 at some well-defined step size.

(2) The authors have introduced a new TFBS section, as a control for their lncRNAs - this is welcome, though again I would ask for caution when interpreting results. For instance, in their reply to me the authors state:

"The number of HS TFs and HS lncRNAs (5 vs 66) alone lends strong evidence suggesting that HS lncRNAs have contributed more significantly to human evolution than HS TFs (note that 5 is the union of three intersections between and the three )."

But this assumes the denominator is the same! There are 35899 lncRNAs according to the current GENCOVE build; 66/35899 = 0.0018, so, 0.18% of lncRNAs are HS. The authors compare this to 5 TFs. There are 19433 protein coding genes in the current GENCOVE build, which naively (5/19433) gives a big depletion (0.026%) relative to the lnc number. However, this assumes all protein coding genes are TFs, which is not the case. A quick search suggests that ~2000 protein coding genes are TFs (see, eg, https://pubmed.ncbi.nlm.nih.gov/34755879/); which gives an enrichment (although I doubt it is a statistically significant one!) of HS TFs over HS lncRNAs (5/2000 = 0.0025). Hence my emphasis on needing to be sure the controls are robust and valid throughout!

(3) In my original review I said:

line 187: "Notably, 97.81% of the 105141 strong DBSs have counterparts in chimpanzees, suggesting that these DBSs are similar to HARs in evolution and have undergone human-specific evolution." I do not see any support for the inference here. Identifying HARs and acceleration relies on a far more thorough methodology than what's being presented here. Even generously, pairwise comparison between two taxa only cannot polarise the direction of differences; inferring human-specific change requires outgroups beyond chimpanzee.

In their reply to me, the authors state:

Here, we actually made an analogy but not an inference; therefore, we used such words as "suggesting" and "similar" instead of using more confirmatory words. We have revised the latter half sentence, saying "raising the possibility that these sequences have evolved considerably during human evolution".

Is the aim here to draw attention to the ~2.2% of DBS that do not have a counterpart? In that case, it would be better to rewrite the sentence to emphasise those, not the ones that are shared between the two species? I do appreciate the revised wording, though.

(4) Finally, Line 408: "Ensembl-annotated transcripts (release 79)" Release 79 is dated to March 2015, which is quite a few releases and genome builds ago. Is this a typo? Both the human and the chimpanzee genome have been significantly improved since then!

Author response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #2 (Public review):

In this valuable manuscript, Lin et al attempt to examine the role of long non coding RNAs (lncRNAs) in human evolution, through a set of population genetics and functional genomics analyses that leverage existing datasets and tools. Although the methods are incomplete and at times inadequate, the results nonetheless point towards a possible contribution of long non coding RNAs to shaping humans, and suggest clear directions for future, more rigorous study.

Comments on revisions:

I thank the authors for their revision and changes in response to previous rounds of comments. As before, I appreciate the changes made in response to my comments, and I think everyone is approaching this in the spirit of arriving at the best possible manuscript, but we still have some deep disagreements on the nature of the relevant statistical approach and defining adequate controls. I highlight a couple of places that I think are particularly relevant, but note that given the authors disagree with my interpretation, they should feel free to not respond!

(1) On the subject of the 0.034 threshold, I had previously stated: "I do not agree with the rationale for this claim, and do not agree that it supports the cutoff of 0.034 used below."

In their reply to me, the authors state:

"What we need is a gene number, which (a) indicates genes that effectively differentiate humans from chimpanzees, (b) can be used to set a DBS sequence distance cutoff. Since this study is the first to systematically examine DBSs in humans and chimpanzees, we must estimate this gene number based on studies that identify differentially expressed genes in humans and chimpanzees. We choose Song et al. 2021 (Song et al. Genetic studies of human-chimpanzee divergence using stem cell fusions. PNAS 2021), which identified 5984 differentially expressed genes, including 4377 genes whose differential expression is due to trans-acting differences between humans and chimpanzees. To the best of our knowledge, this is the only published data on trans-acting differences between humans and chimpanzees, and most HS lncRNAs and their DBSs/targets have trans-acting relationships (see Supplementary Table 2). Based on these numbers, we chose a DBS sequence distance cutoff of 0.034, which corresponds to 4248 genes (the top 20%), slightly fewer than 4377."

I have some notes here. First, Agoglia et al, Nature, 2021, also examined the nature of cis vs trans regulatory differences between human and chimps using a very similar set up to Song et al; their Supplementary Table 4 enables the discovery of genes with cis vs trans effects although admittedly this is less straightforward than the Song et al data. Second, I can't actually tell how the 4377 number is arrived at. From Song et al, "Of 4,671 genes with regulatory changes between human-only and chimpanzee-only iPSC lines, 44.4% (2,073 genes) were regulated primarily in cis, 31.4% (1,465 genes) were regulated primarily in trans, and the remaining 1,133 genes were regulated both in cis and in trans (Fig. 2C). This final category was further broken down into a cis+trans category (cis- and transregulatory changes acting in the same direction) and a cis-trans category (cis- and trans-regulatory changes acting in opposite directions)." Even when combining trans-only and cis&trans genes that gives 2,598 genes with evidence for some trans regulation. I cannot find 4,377 in the main text of the Song et al paper.

Elsewhere in their response, the authors respond to my comment that 0.034 is an arbitrary threshold by repeating the analyses using a cutoff of 0.035. I appreciate the sentiment here, but I would not expect this to make any great difference, given how similar those numbers are! A better approach, and what I had in mind when I mentioned this, would be to test multiple thresholds, ranging from, eg,0.05 to 0.01 <DBS dist =0.01 -> 0.034 -> 0.05> at some well-defined step size.

(1) We sincerely thank the reviewer for this critical point. Our initial purpose, based on DBS distances from the human genome to chimpanzee genome and archaic genomes, was that genes with large DBS distances may have contributed more to human evolution. However, our ORA (overrepresentation analysis) explored only genes with large DBS distances (the legend of old Figure 2 was “1256 target genes whose DBSs have the largest distances from modern humans to chimpanzees and Altai Neanderthals are enriched in different Biological Processes GO terms”), with the use of the cutoff (threshold) of 0.034 for defining large distance. The cutoff is not totally unreasonable (as our new results and the following sensitivity analysis indicate), but this approach was indirect and flawed.

(2) We have now performed ORA using two methods. The first uses only DBS distances. Instead of using a cutoff, we now sort genes by DBS distance (human-chimpanzee distances and human-Altai Neanderthal distance, respectively, see Supplementary Table 5) and use the top 25% and bottom 25% of genes to perform ORA. This directly examines whether DBS distances along indicate that genes with large DBS distances contribute more to human evolution than genes with small DBS distances. The second also explores the ASE genes (allele-specific expression, genes undergoing human/chimpanzee-specific regulation in the tetraploid human–chimpanzee hybrid iPS) reported by Agoglia et al. 2021. We select the top 50% and bottom 50% of genes with large and small DBS distances, intersect them with ASE genes from Agoglia et al. 2021 (their Supplementary Table 4), and apply ORA to the intersections. Both the results are that: (a) more GO terms are obtained from genes with large DBS distances, (b) more human evolution-related GO terms are obtained from genes with large DBS distances (Supplementary Table 5,6,7; Figure 2; Supplementary Fig. 15). These results directly suggest that genes with large DBS distances contribute more to human evolution than genes with small DBS distances, which is a key theme of the study.

(3) Regarding Song et al 2021, the statement of “we differentiated…allotetraploid (H1C1a, H1C1b, H2C2a, H2C2b) lines into ectoderm, mesoderm, and endoderm” made us assume that their differentiated hybrid cell lines cover more tissue types than those of Agoglia et al. 2021. Now, upon re-examining Supplementary Table 5 of Song et al. and Supplementary Table 4 of Agoglia et al. 2021, we find that the latter more clearly indicates significant ASE genes (p-adj<0.01 and |LFC>0.5| in GRCh38 and PanTro5).

(4) We have also performed two additional analyses in response to the suggestion of “test multiple thresholds, ranging from, eg, 0.05 to 0.01 <DBS dist =0.01 -> 0.034 -> 0.05> at some well-defined step size”. First, we performed a multi-threshold sensitivity analysis using a spectrum of cutoffs (0.03, 0.034, 0.04, 0.05), and tracked the number of genes identified and the enrichment significance of key GO terms (e.g., "neuron projection development," "behavior") across these thresholds. The result confirms that while the absolute number of genes varies with the cutoffs, the core biological conclusion (specifically, the significant enrichment of target genes in neurodevelopmental and cognitive functions) remains stable and significant. For instance, "behavior" maintains strong statistical significance (FDR<0.01) in both the human-chimpanzee and human-Altai Neanderthal comparisons across all tested cutoffs, and "Neuron projection development" also remains significant across three (0.03, 0.034, 0.04) of the four cutoffs in the Altai comparison. This pattern suggests that our core findings regarding neurodevelopmental functions are robust across a range of cutoffs. Nevertheless, we did not extend the analysis to smaller cutoffs (e.g., 0.01 or 0.02) because such values would identify an excessively large number of genes (>10000) for ORA, which would render the GOterm enrichment analysis less meaningful due to a loss of specificity.

Second, we have performed an additional validation to directly evaluate whether the 0.034 cutoff itself represents a stringent and biologically meaningful value. We sought to empirically determine how often a DBS sequence distance of 0.034 or greater might occur by chance in promoter regions, thereby testing its significance as a marker of potential evolutionary divergence. We randomly sampled 10,000 windows from annotated promoter regions across the hg38 genome, each with a size matching the average length of DBSs (147 bp). We then calculated the per-base sequence distances for these random windows between modern humans and chimpanzees, as well as between modern humans and the three archaic humans (Altai, Denisovan, Vindija). The analysis reveals that a distance of ≥0.034 is a rare event in random promoter sequences: for Human-Chimp, Human-Altai, HumanDenisovan, and Human-Vindija, 5.49% (549/10000), 0.31% (31/10000), 4.47% (447/10000), and0.03% (3/10000) of random windows reach this distance. This empirical evidence suggests that 0.034 is a sufficiently strong cutoff for defining large DBS distance, it would occur very unlikely in a random genomic background (P<0.1 for Chimpanzee and P<0.05 for the archaic humans), and DBSs exceeding this cutoff are significantly enriched for sequences that have undergone substantial evolutionary change instead of being random neutral variations.

(5) We present new Figure 2, Supplementary Table 5,6,7, and Supplementary Fig. 15. We have substantially revised section 2.3, related sections in Results, Supplementary Note 3, and Supplementary Table 8. We have removed related descriptions and explanations in the main text and Supplementary Notes. The results of the above two analyses are presented here as two Author response images.

Distribution of per-base sequence distances for DBS size-matched random genomic windows in Ensembl-annotated promoter regions, calculated between modern humans and (A) chimpanzee, (B) Altai Neanderthal, (C) Denisovan, and (D) Vindija Neanderthal genomes.

(2) The authors have introduced a new TFBS section, as a control for their lncRNAs - this is welcome, though again I would ask for caution when interpreting results. For instance, in their reply to me the authors state: "The number of HS TFs and HS lncRNAs (5 vs 66) alone lends strong evidence suggesting that HS lncRNAs have contributed more significantly to human evolution than HS TFs (note that 5 is the union of three intersections between <many2zero + one2zero> and the three )."

But this assumes the denominator is the same! There are 35899 lncRNAs according to the current GENCOVE build; 66/35899 = 0.0018, so, 0.18% of lncRNAs are HS. The authors compare this to 5 TFs. There are 19433 protein coding genes in the current GENCOVE build, which naively (5/19433) gives a big depletion (0.026%) relative to the lnc number. However, this assumes all protein coding genes are TFs, which is not the case. A quick search suggests that ~2000 protein coding genes are TFs (see, eg, https://pubmed.ncbi.nlm.nih.gov/34755879/); which gives an enrichment (although I doubt it is a statistically significant one!) of HS TFs over HS lncRNAs (5/2000 = 0.0025). Hence my emphasis on needing to be sure the controls are robust and valid throughout!

We thank the reviewer for this comment. While 5 vs 66 reveals a difference, a direct comparison is too simplified. The real take-home message of the new TFBS section is not the numbers but the distributions of HS TFs’ targets and HS lncRNAs’ targets across GTEx organs and tissues (Figure 3 and Supplementary Figures 24, 25) - correlated HS lncRNA-target transcript pairs are highly enriched in brain regions, but correlated HS TF-target transcript pairs are distributed broadly across GTEx tissues and organs. We have now removed the simple comparison of “5 vs 66” and more carefully explained our comparison in section 2.6.

(3) In my original review I said: line 187: "Notably, 97.81% of the 105141 strong DBSs have counterparts in chimpanzees, suggesting that these DBSs are similar to HARs in evolution and have undergone human-specific evolution." I do not see any support for the inference here. Identifying HARs and acceleration relies on a far more thorough methodology than what's being presented here. Even generously, pairwise comparison between two taxa only cannot polarise the direction of differences; inferring human-specific change requires outgroups beyond chimpanzee.

In their reply to me, the authors state:

Here, we actually made an analogy but not an inference; therefore, we used such words as "suggesting" and "similar" instead of using more confirmatory words. We have revised the latter half sentence, saying "raising the possibility that these sequences have evolved considerably during human evolution".

Is the aim here to draw attention to the ~2.2% of DBS that do not have a counterpart? In that case, it would be better to rewrite the sentence to emphasise those, not the ones that are shared between the two species? I do appreciate the revised wording, though.

(1) Our original phrasing may be misleading, and we agree entirely that “pairwise comparison between two taxa only cannot polarise the direction of differences; inferring human-specific change requires outgroups beyond chimpanzee”. As explained in that reply, we know and think that DBSs and HARs are two different classes of sequences, and indeed, identifying HARs and acceleration relies on a far more thorough methodology. Yet, three factors prompted us to compare them. First, both suggest the importance of sequences outside genes. Second, both are quite “old” sequences and have undergone considerable evolution recently (although the references are different). Third, both have contributed greatly to human brain evolution.

(2) Here, our stress is 97.81% but not 2.2%, and we have made this analogy more clearly and cautiously. Relevant revisions have been made in the Results, Discussion, and Methods sections.

(3) We also have further determined whether the 2.2% DBSs are human-specific gains by analyzing them using the UCSC Multiz Alignments of 100 Vertebrates. The result confirms that all 2248 DBSs are present in the human genome but are absent from the chimpanzee genome and all other aligned vertebrate genomes. We add this result into the manuscript.

(4) Finally, Line 408: "Ensembl-annotated transcripts (release 79)" Release 79 is dated to March 2015, which is quite a few releases and genome builds ago. Is this a typo? Both the human and the chimpanzee genome have been significantly improved since then!

(1) We thank the reviewer for this comment, which prompts us to provide further explanation and additional data. First, we began predicting HS lncRNAs’ DBSs when Ensembl release 79 was available, but did not re-predict DBSs when new Ensembl releases were published because (a) these new Ensembl releases are based also on hg38, (b) we did not find any fault in the LongTarget program during our use, nor received any one from users, (c) predicting lncRNAs’ DBSs using the LongTarget program is highly time-consuming.

(2) Second, to assess the influence of newer Ensembl releases, we compared the promoters annotated in release 79 and in release 115. We found that the vast majority (87.3%) of promoters newly annotated in release 115 belong to non-coding genes. Thus, using release 115 may predict more DBSs in non-coding genes, but downstream analyses based on protein-coding genes would be essentially the same (meaning that all figures and tables would be the same).

(3) Third, a key element of this study is GTEx data analysis, and these data were also published years ago.

(4) Finally, some lncRNA genes have new gene symbols in new Ensembl releases. To allow researchers to use our data conveniently, we have added a new column titled "Gene symbol (Ensembl release115)" to Supplementary Tables 2A and 2B.

Summary:

Major changes based on Reviewer’s comments:

(1) The following revisions are made to address the comment on “the 0.034 threshold”: (a) Section 2.3, section 2.4, Supplementary Note 3, and related contents in Discussion and Methods are revised, (b) new Figure 2, Supplementary Figure 15, new Supplementary Table 5,6,7, (c) Table 2 and Supplementary Table 8 are revised.

(2) To address the comment on “new TFBS section”, section 2.6 and section 4.13 are revised.

(3) To address the comment on “97.81% and 2.2% of DBSs”, section 2.3 is revised.

(4) The following revisions are made to address the comment on “release 79”: (a) the old Supplementary Table 2, 3 are merged to Supplementary Table 2AB, and the new column "Gene symbol (Ensembl release115)" is added to Supplementary Table 2AB, (b) accordingly, Supplementary Table 4,5 are renamed to Supplementary Table 3,4.

Additional revisions:

(1) Section 2.5 “Young weak DBSs may have greatly promoted recent human evolution” is moved into Supplementary Note 3 (which now has the subtitle “Target genes with specific DBS features are enriched in specific functions”), because this section is short and lacking sufficient cross-validation.

(2) Considerable minor revisions of sentences have been made.

(3) Since there are many supplementary figures, the main text now cites only Supplementary Notes, as the reader can easily access supplementary figures in Supplementary Notes.

Version 5

Abstract

What genes and regulatory sequences critically differentiate modern humans from apes and archaic humans, which share highly similar genomes but show distinct phenotypes, has puzzled researchers for decades. Previous studies examined species-specific protein-coding genes and related regulatory sequences, revealing that birth, loss, and changes in these genes and sequences can drive speciation and evolution. However, investigations of species-specific lncRNA genes and related regulatory sequences, which regulate substantial genes, remain limited. We identified human-specific (HS) lncRNAs from GENCODE-annotated human lncRNAs, predicted their DNA binding domains (DBDs) and binding sites (DBSs), analyzed DBS sequences in modern humans (CEU, CHB, and YRI), archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and chimpanzees, and investigated how HS lncRNAs and their DBSs have influenced gene expression in archaic and modern humans. Our results suggest that these lncRNAs and DBSs have substantially reshaped gene expression. This reshaping has evolved continuously from archaic to modern humans, enabling humans to adapt to new environments and lifestyles, promoting brain evolution, and resulting in cross-population differences. The parallel analysis of gene expression in GTEx tissues by HS TFs and their DBSs indicates that HS lncRNAs have reshaped gene expression in the brain more significantly than HS TFs.

Introduction

The limited genomic but substantial phenotypic and behavioral differences between humans and other hominids make which sequence changes have critically driven human evolution an enduring puzzle (Anton et al., 2014; Pollen et al., 2023). Sequence changes include the birth and loss of genes, as well as the turnover and change of regulatory sequences (Albalat and Canestro, 2016; Kaessmann, 2010; Prud’homme et al., 2007). Studies have identified human-specific (HS) genes important for promoting human brain development (e.g., NOTCH2NL and ASPM) (Evans et al., 2005; Fiddes et al., 2018; Mekel-Bobrov et al., 2005; Pinson et al., 2022; Suzuki et al., 2018). However, gene-centric studies have limitations (Currat et al., 2006; Timpson et al., 2007; Yu et al., 2007), including that genes promoting brain enlargement may not critically determine other traits (e.g., bipedal walking); meanwhile, studies on species-specific lncRNA genes remain limited. LncRNA genes are a major class of new genes in mammals, and HS lncRNAs may have greatly influenced human evolution. On the other hand, multiple regulatory sequence changes important for human evolution have been identified, including new sequences (Liu et al., 2021), lost sequences (McLean et al., 2011), human accelerated regions (HARs) (Dong et al., 2016; Mangan et al., 2022; Prabhakar et al., 2006; Whalen and Pollard, 2022), and turnover of transcription factor (TF) DNA-binding domains (TF DBD, commonly called DNA-binding motifs) and DNA-binding sites (DBS) (Krieger et al., 2022; Otto et al., 2009; Zhang et al., 2023). However, no systematic investigation has been reported on lncRNA DBD and DBS.

Studies have generated multiple important findings about lncRNAs and epigenetic regulation. (a) About one-third of human lncRNAs are primate-specific (Derrien et al., 2012). (b) Species-specific lncRNAs exhibit distinct expression in tissues and organs of different species (Sarropoulos et al., 2019). (c) Many lncRNAs can bind to DNA sequences by forming RNA:DNA triplexes (Abu Almakarem et al., 2012), recruit histone and DNA modification enzymes to DBSs, and regulate transcription (Lee, 2009; Roadmap Epigenomics et al., 2015). (d) Approximately 40% of differentially expressed human genes result from interspecies epigenetic differences (Hernando-Herraez et al., 2015). Thus, besides HS TFs and their DBSs, HS lncRNAs and their DBSs also critically regulate gene expression human-specifically.

This study focuses on HS lncRNAs, HS lncRNA DBSs, and their impacts on human evolution by exploring multiple methods and resources. The first is RNA sequence search based on structure and sequence alignment to identify orthologues of lncRNA genes in genomes (Nawrocki, 2014; Nawrocki and Eddy, 2013). The second is the specific base-pairing rules that RNA and DNA sequences follow to form RNA:DNA triplexes (Abu Almakarem et al., 2012); these rules allow computational prediction of lncRNA DBDs and DBSs (Lin et al., 2019; Wen et al., 2022). The third is gene expression in organs and tissues, especially the Genotype-Tissue Expression (GTEx) project (GTEx Consortium, 2017), which provides data for examining and comparing gene expression regulation by HS lncRNAs and HS TFs. Finally, the genomes of modern humans, archaic humans, and multiple apes (especially chimpanzees) allow cross-species genomic and transcriptomic analysis (1000 Genomes Project Consortium et al., 2012; Chimpanzee Sequencing and Analysis Consortium, 2005; Meyer et al., 2012; Prufer et al., 2017; Pruüfer et al., 2013).

This study identified HS lncRNAs upon the orthologues of the GENCODE-annotated human lncRNA genes in 16 mammalian genomes, predicted their DBDs and DBSs in modern humans (CEU, CHB, and YRI, the three representative populations of Europeans, Asians, and Africans), archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and identified the counterparts of HS lncRNA DBSs in chimpanzees (Figure 1A). Our DBS prediction method combines a local alignment algorithm with RNA:DNA base-pairing rules to identify RNA:DNA triplexes, thus predicting DBDs and DBSs simultaneously. This outperforms DBS prediction using RNA:DNA base-pairing rules alone (Lin et al., 2019). Based on reliable recent data and methods, we also predicted HS TFs and their DBDs and DBSs. Cross-population and cross-species analyses based on HS TFs and their DBSs, HS lncRNAs and their DBSs, and substantial genomic and transcriptomic data suggest that HS lncRNAs and their DBSs have distinctly and continuously reshaped gene expression for adaptive human evolution.

Study overview.

(A) The relationships between chimpanzees, the three archaic humans (Altai Neanderthals, Denisovans, and Vindija Neanderthals), and the three modern human populations, with dashed lines indicating the phylogenetic distances from modern humans based on related studies. Based on the left-top icons, the DBS in B2M lacks a counterpart in chimpanzees; the DBS in ABL2 has great differences between archaic and modern humans; the DBS in IRNAR1 is polymorphic in modern humans (red letters indicate tissue-specific eQTLs or population-specific mutations). (B) The mean length and affinity of strong DBSs. (C) Numbers of target genes and target transcripts of HS lncRNAs. (D) An illustrative figure showing the targeting relationships between HS lncRNAs. (E) The sequence distances of DBSs (the top 40%) from modern humans to chimpanzees and archaic humans. (F) An illustrative figure showing the impacts of HS lncRNA-target transcript on gene expression in GTEx tissues (see Supplementary Note 2 and Figure 3).

Results

HS lncRNAs regulate diverse genes and transcripts

How many human lncRNAs are human-specific still lacks a precise estimate. Based on the orthologues of 13562 GENCODE-annotated human lncRNA genes (v18) in 16 mammalian genomes we searched using the Infernal program (which features in a combined sequence and structure alignment) (Lin et al., 2019; Nawrocki, 2014; Nawrocki et al., 2009), we identified 66 HS lncRNA genes that exist in humans but not in any other species (Supplementary Table 1). Using the LongTarget program, we predicted DBDs in exons of HS lncRNAs and their DBSs in the 5000 bp promoter regions of the 179128 Ensembl-annotated transcripts (release 79) (Lin et al., 2019). DBS prediction was validated using multiple methods and datasets (Supplementary Note 1). Predicted DBSs overlap well with experimentally identified DNA methylation and histone modification signals in multiple cell lines (https://genome.UCSC.edu), and overlap well with experimentally detected DNA binding sites of NEAT1, MALAT1, and MEG3 (Mondal et al., 2015; West et al., 2014). Many DBSs also co-localize with annotated cis-regulatory elements (cCREs) in promoter regions (https://genome.UCSC.edu). Specifically, we used CRISPR/Cas9 to delete multiple DBDs (100-200 bp) in several cell lines, performed RNA-seq before and after DBD knockout (KO), and analyzed the resulting differential gene expression. The seven DBD KO cases demonstrate that the |fold change| of target genes was significantly larger than the |fold change| of non-target genes (one-sided Mann-Whitney test, p=3.1e-72, 1.49e-114, 1.12e-206, 2.58e-09, 6.49e-41, 0.034, and 5.23e-07). Because DBD1’s DBSs vastly outnumber any other DBD’s DBSs, DBD1s not only regulate more targets than other DBDs but also are more reliable. Thus, this study analysed only DBD1s (simply called DBDs) and their DBSs.

To perform quantitative and comparative analysis, we identified HS lncRNAs and their DBSs in archaic humans and chimpanzees (Figure 1A), defined the binding affinity (simply called affinity) of a pair of DBD and DBS as the product of their length and identity (the percentage of paired RNA and DNA nucleotides), and computed DBSs’ sequence distances from humans to archaic humans and chimpanzees. LncRNA:DNA binding analysis at known imprinted genes suggests that affinity better characterizes DBSs than length (He et al., 2015). With affinity and distance, DBSs were classified into strong (affinity>=60) and weak (36<affinity<60) ones (as well as old and young ones based on when large sequence changes occurred). 105141 strong DBSs (mean length>147 bp) were identified in 96789 transcripts of 23396 genes, and 152836 weak DBSs were identified in 127898 transcripts of 33185 genes (Supplementary Table 2). Several HS lncRNAs (especially RP11-423H2.3) have abundant DBSs; only about 1.5% of target genes (0.6% of target transcripts) are human-specific; many targets have DBSs of multiple HS lncRNAs; many HS lncRNAs have multiple DBSs in a target gene; and HS lncRNAs themselves show complex targeting relationships (Figure 1BCD; Supplementary Note 2). These suggest that HS lncRNAs have significantly reshaped gene expression regulation.

Target genes with strong and weak DBS characterize adaptive evolution and human traits

Genes with significant sequence differences between humans and chimpanzees range from 1.24% to 5% (Britten, 2002; Ebersberger et al., 2002). However, substantial genes exhibit expression differences due to variations in regulatory sequences. It is therefore interesting to examine whether genes with strong and weak DBSs are enriched in different functions. We performed over-representation analysis (ORA) using the g:Profiler program, Gene Ontology (GO) database, and genes sorted by affinity (Supplementary Table 2). When choosing a cutoff to distinguish strong from weak DBSs, we found that the GO terms of the top 2000 protein-coding genes contain, but those of the top 1500 protein-coding genes lack, terms such as “hair follicle development” and “skin epidermis development” that critically differentiate humans from chimpanzees. We therefore performed ORA using the top 2000 and bottom 2000 protein-coding genes, respectively. In addition to shared GO terms (e.g., “behaviour”), the two gene sets have many specific ones. GO terms specific to the top 2000 genes include “renal system development”, and GO terms specific to the bottom 2000 genes include “cellular response to alcohol”, “response to temperature stimulus”, and “female pregnancy” (Supplementary Table 3; Supplementary Note 3).

Genes with strongest DBSs (affinity>=300) include IFNAR1 and NFATC1 (important for immune function), KIF21B and NTSR1 (critical for neural development), SLC2A11 and SLC2A1 (involved in glucose usage), BAIAP3 (a brain-specific angiogenesis inhibitor), TAS1R3 (a receptor for the sweet taste response), and several primate-specific lncRNA genes (e.g., CTD-3224I3.3 with high expression in the cerebellum, lung, and testis). DBSs in these genes underwent large sequence changes from chimpanzees and archaic humans to modern humans (Table 1; Supplementary Table 4).

Genes intensively regulated by HS lncRNAs may have promoted human evolution

The human genome is approximately 99%, 99%, and 98% identical to the genomes of chimpanzees, bonobos, and gorillas (Chimpanzee Sequencing and Analysis Consortium, 2005; Prufer et al., 2012). However, the sequence variants that critically differentiate these species remain unclear. To assess whether regulation by HS lncRNAs is critical, we first examined whether DBSs of HS lncRNAs are human-specific or conserved across these species, and found that 97.81% of the 105141 strong DBSs have counterparts in chimpanzees. To further determine whether the remaining 2.2% are human-specific gains, we checked them using the UCSC Multiz Alignments of 100 Vertebrates, and found that they are present in the human genome but absent from the chimpanzee genome and all other aligned vertebrate genomes. Since most DBSs have chimpanzee counterparts, yet have evolved considerably from chimpanzees to archaic and modern humans, they share features of HARs. However, of the 312 HARs identified by the Zoonomia Project as important for 3D genome rewiring and neurodevelopment (Keough et al., 2023), only eight overlap 26 DBSs of 14 HS lncRNAs, suggesting that DBSs and HARs may contribute differently to human evolution.

HS lncRNAs’ DBSs may be generated before, together with, or after HS lncRNAs, and the first situation suggests that HS lncRNAs reshape gene expression via DBSs. We therefore identified counterparts of HS lncRNAs and their DBSs in Altai Neanderthals, Denisovans, and Vindija Neanderthals (Meyer et al., 2012; Prufer et al., 2017; Pru fer et al., 2013). While both classes of counterparts were identified in these archaic humans, sequence distances (defined as the distance per base throughout the manuscript) of HS lncRNAs from humans to archaic humans are smaller than those of DBSs, suggesting that many DBS sequences were generated before HS lncRNAs.

We then computed DBS sequence distances using two methods. The first is from the reconstructed human ancestor (downloaded from the EBI website) to chimpanzees, archaic humans, and modern humans. In this result, many DBS distances from the human ancestor to modern humans are shorter than those to archaic humans (Supplementary Note 3). The second is from the human genome to chimpanzees and archaic humans. This set of distances agrees better with the phylogenetic distances between chimpanzees, archaic humans, and modern humans (Figure 1).

We postulate that genes with large DBS distances may have contributed more to human evolution than genes with small DBS distances. To test this postulation, we sorted genes by DBS distances from humans to chimpanzees and to Alai Neanderthals (Supplementary Table 5), and applied ORA to genes with large and small DBS distances using the g:Profiler program and GO database (Benjamini-Hochberg FDR, threshold=0.05, 50<terms size<1000). First, we examined the top 25% and bottom 25% of genes to determine whether they are enriched for different GO terms. The result indicates that the top 25% of genes generate more enriched GO terms and also more human evolution-related GO terms (Figure 2A; Supplementary Table 6). Second, we examined the top and bottom genes that also critically differentiate humans from chimpanzees. Agoglia et al. fused human and chimpanzee induced pluripotent stem cells to generate tetraploid hybrid stem cells (hybrid iPS), differentiated these cells into neural organoids, and measured allele-specific expression (ASE) of genes in hybrid iPS (Agoglia et al., 2021). We selected the top 50% and bottom 50% of genes, intersected them with the 2891 genes with significant ASE (p-adj<0.01 and |LFC|>0.5) (Supplementary Table 5), and applied ORA to the four gene sets. The four results indicate that more GO terms, especially more human evolution-related GO terms, were generated by the top 50% of genes (Figure 2B; Supplementary Table 7). Using ASE genes with less significance (just p-adj<0.01) yielded similar results (Supplementary Note 3). These results support the postulation that genes with large DBS distances contribute more to human evolution than genes with small DBS distances. Different GO terms, and GO terms with different significance levels, generated based on DBS distances to chimpanzees and archaic humans (e.g., “sensory perception of sound”), provide information on the timing and content of reshaped transcriptional regulation’s influence on human evolution.

GO terms generated by different gene sets with large and small DBS distances from humans to chimpanzees and Altai Neanderthals.

Shown are the presence and absence of GO terms highly related to human evolution. (A) Genes sorted by DBS distance from humans to chimpanzees and to Altai Neanderthals. Left: The top 25% of genes. Right: The bottom 25% of genes. (B) The intersections of the top 50% and bottom 50% of genes (with DBS distances from humans to chimpanzees and to Altai Neanderthals) and genes with significant ASE (p-adj<0.01 and |LFC|>0.5).

Regulation by HS lncRNAs shows cross-population differences

To reveal whether DBSs with large human-chimpanzee distances have further evolved in archaic and modern humans, we extracted those that (a) are in the top 20% of genes sorted by human-chimpanzee DBS distance (>0.037) and (b) also have a distance >0.037 to at least one archaic human. We label these DBSs using “A”, “D”, and “V” (e.g., “A,D” indicates the distances to both Altai Neanderthals and Denisovan >0.037), calculated SNP number therein (for SNPs with minimal allele frequency (MAF)≥0.1), and computed weighted Fst and Tajima’s D (1000 Genomes Project Consortium et al., 2012). SNP number per base, weighted Fst, and Tajima’s D reveal that many DBSs that undergo significant and continuous sequence change are also highly polymorphic and have been selected in specific modern humans (Supplementary Table 8). Many genes with these DBSs encode lncRNAs, and those encoding proteins include SCTR, NCR, IFNAR1, NFATC1, TAS1R3, INS, ST3GAL4, and FN3KRP, which are important for adaptive human evolution (Table 2). These suggest that the regulation of genes important for human evolution by HS lncRNAs has undergone continuous evolution.

Genes involved in sugar metabolism are notable. A key feature of modern human diets is high sugar intake, which can cause non-enzymatic oxidation of proteins (i.e., glycation) and deactivate protein functions. Among proteins encoded by genes in this intersection, TAS1R3 recognizes diverse natural and synthetic sweeteners, Insulin decreases blood glucose concentration, ST3GAL4 regulates protein glycosylation, and FN3KRP deglycates proteins. These DBSs and target genes indicate the continuous evolution of transcriptional regulation with human- and population-specificity.

To examine whether SNPs in DBSs are neutral or indicate selection, we used multiple statistical tests with widely adopted parameters, including XP-CLR (Chen et al., 2010), iSAFE (Akbari et al., 2018), Tajima’s D (Tajima, 1989), Fay-Wu’s H (Fay and Wu, 2000), the fixation index (Fst) (Weir and Cockerham, 1984), and linkage disequilibrium (LD) (Slatkin, 2008), to detect selection signals in HS lncRNAs and strong DBSs in CEU, CHB, and YRI (Supplementary Note 4, 5; Supplementary Table 9, 10, 11). Selection signals were detected in several HS lncRNA genes in CEU and CHB but not in YRI, and selection signals were detected in more DBSs in CEU and CHB than in YRI. The same signals were often detected by multiple tests. These results agree with the reports that fewer selection signals are detected in YRI (Sabeti et al., 2007; Voight et al., 2006).

SNPs in DBSs exhibit cis-effects on gene expression

To verify that DBSs and SNPs influence gene expression, we analysed the GTEx data (GTEx Consortium, 2017). We identified expression quantitative trait loci (eQTLs) in DBSs using the widely adopted criteria of MAF≥0.1 and cis-effect size (ES)≥0.5. 1727 SNPs, with MAF≥0.1 in at least one population and |ES|≥0.5 in at least one tissue, were identified in DBSs in autosomal genes (Supplementary Table 12,13). These eQTLs include 372 “conserved” ones (i.e., also in DBSs in the three archaic humans) and 1020 “novel” ones (i.e., only in DBSs in modern humans). A notable eQTL with high derived allele frequencies (DAFs) across all three populations and a positive ES≥0.5 in 44 of 48 GTEx tissues is rs2246577 in the DBS in FN3KRP, whose encoded protein deglycates proteins to restore their function (Table 2). Many conserved eQTLs are expressed in brain tissues and exhibit high DAFs across all three modern populations. In contrast, many novel eQTLs are tissue-specific and exhibit population-specific DAFs (Supplementary Note 6).

Next, we performed two analyses to examine how eQTLs in DBSs influence gene expression. First, we computed the expression correlation between HS lncRNAs and target transcripts with DBSs having eQTLs in specific tissues. Most (94%) HS lncRNA-target transcript pairs showed significant expression correlation in tissues in which the eQTLs were identified (|Spearman’s rho|>0.3 and FDR<0.05). Second, we examined whether eQTLs are more enriched in DBSs than in Ensembl-annotated promoters by computing and comparing the eQTL density in the two classes of regions (promoters were used as the reference because they contain DBSs). The results indicate that eQTLs are more enriched in DBSs than in promoters (one-sided Mann-Whitney test, p=0.0) (Supplementary Note 6). Thus, population-specific SNPs and tissue-specific eQTLs, at the genome and transcriptome levels, respectively, suggest that DBSs enable HS lncRNAs to regulate transcription with population- and tissue-specific features.

HS lncRNAs promote brain evolution from archaic to modern humans

To further verify that SNPs in DBSs have cis-effects on gene expression, we examined the impact of HS lncRNA DBSs on gene expression in the GTEx tissues (GTEx Consortium, 2017). 40 autosomal HS lncRNAs are expressed in at least one tissue (median TPM>0.1), these HS lncRNAs and their target transcripts form 198876 pairs in all tissues, and 45% of pairs show a significant expression correlation in specific tissues (Spearman’s |rho|>0.3 and FDR<0.05). To assess the likelihood that these correlations could be generated by chance, we randomly sampled 10000 pairs of lncRNAs and protein-coding transcripts genome-wide and found that only 2.3% of pairs showed significant expression correlation (Spearman’s |rho|>0.3 and FDR<0.05). Moreover, a higher percentage (56%) of HS lncRNA-target transcript pairs with significant correlation was detected in at least one brain region, indicating more extensive gene expression regulation in the brain (Figure 3A).

The impact of HS lncRNA-DBS interaction on gene expression in GTEx tissues and organs.

(A) The distribution of the percentage of HS lncRNA-target transcript pairs with correlated expression across GTEx tissues and organs. Higher percentages of correlated pairs are in brain regions than in other tissues and organs. (B) The distribution of significantly changed DBSs (in terms of sequence distance) in HS lncRNA-target transcript pairs across GTEx tissues and organs between archaic and modern humans. Orange, red, and dark red indicate significant changes from Denisovans (D), Altai Neanderthals and Denisovans (AD), and all three archaic humans (ADV). DBSs in HS lncRNA-target transcript pairs with correlated expression in seven brain regions (in dark red) have changed significantly and consistently since the Altai Neanderthals, Denisovans, and Vindija Neanderthals (one-sided two-sample Kolmogorov-Smirnov test, significant changes determined by FDR <0.001).

To obtain more supporting evidence on the transcriptome level, we analyzed nine experimental datasets (Supplementary Note 7). First, we analyzed two datasets of epigenetic studies: one examining H3K27ac and H3K4me2 profiling in human, macaque, and mouse corticogenesis, and the other examining gene expression and H3K27ac modification in eight brain regions in humans and four other primates (Reilly et al., 2015; Xu et al., 2018). Compared with the genome-wide background, 84% and 73% of genes in the two datasets have DBSs for HS lncRNAs, indicating significantly higher enrichment in HS lncRNA targets (p=1.21e-21 and 1.2e-56, two-sided Fisher’s exact test). When chimpanzee gene expression was used as the control, 1851 genes showed human-specific transcriptomic differences in one or multiple brain regions, whereas only 240 genes showed chimpanzee-specific transcriptomic differences. Second, we analyzed two datasets of PsychENCODE studies: one examining spatiotemporally differentially expressed genes and spatiotemporally differentially methylated sites across 16 brain regions, and the other examining the spatiotemporal transcriptomic divergence between human and macaque brain development (Li et al., 2018; Zhu et al., 2018). In the first dataset, 65 HS lncRNAs are expressed in all 16 brain regions, 109 transcripts have spatiotemporally differentially methylated sites in their promoters, and 56 of the 109 transcripts have DBSs for HS lncRNAs. In the second dataset, 8951 genes show differential expression between the human and macaque brains, and 72% of differentially expressed genes in the human brain have DBSs for HS lncRNAs. Thus, both datasets show significant enrichment for HS lncRNA regulation. Third, three studies identified genes critically regulating cortical expansion (Florio et al., 2018; Johnson et al., 2018; Suzuki et al., 2018). Of the 40 reported protein-coding genes, 29 have DBSs of HS lncRNAs in promoter regions (Supplementary Table 14). Thus, these genes are enriched with DBSs of HS lncRNAs compared to the genome-wide background (p<0.01, two-sided Fisher’s exact test). Fourth, we analyzed two datasets of brain organoid studies. By establishing and comparing cerebral organoids between humans, chimpanzees, and macaques, Pollen et al. identified 261 human-specific gene expression changes (Pollen et al., 2019). By fusing human and chimpanzee iPS cells and differentiating the hybrid iPS cells into hybrid cortical spheroids, Agoglia et al. generated a panel of tetraploid human-chimpanzee hybrid iPS cells and identified thousands of genes with divergent expression between humans and chimpanzees (Agoglia et al., 2021). We found that 261 and 1102 genes in the two datasets are enriched for DBSs of HS lncRNAs compared to the genome-wide background (p=1.2e-16 and 3.4e-74).

Next, we examined the evolution of the impact of HS lncRNAs and their DBSs on gene expression using the GTEx data. For each DBS in an HS lncRNA-target transcript pair that shows correlated expression in a GTEx tissue, we computed the sequence distances of this DBS from the modern humans to the three archaic humans. Then, we compared the distribution of DBS sequence distances in each tissue with that in all tissues as the background (one-sided two-sample Kolmogorov-Smirnov test). DBSs in HS lncRNA-target transcript pairs with correlated expression in brain regions have significantly changed sequence distances since the Altai Neanderthals and Denisovans (Figure 3B), supporting that gene expression regulation by HS lncRNAs in the brain has undergone more significant evolution (Ponce de Leon et al., 2021).

To substantiate the above conclusion, we further examine whether HS lncRNAs have contributed more to human evolution than HS TFs. Based on the “hg38-panTro6” gene sets reported by Kirilenko et al. and the human TF lists reported by previous studies (Bahrami et al., 2015; Kirilenko et al., 2023; Lambert et al., 2018), five HS TFs were identified. We predicted their DBSs in the 5000 bp promoter regions of the same 179128 Ensembl-annotated transcripts (release 79) using the FIMO and CellOracle programs (Grant et al., 2011; Kamimoto et al., 2023), identified counterparts of these HS TF DBSs in archaic humans and chimpanzees, computed sequence distances of these DBSs from modern humans to archaic humans and chimpanzees (Supplementary Table 15), computed the Pearson correlation of HS TFs and their target transcripts across GTEx tissues, and computed sequence distances of DBSs in HS TF-target transcript pairs across GTEx tissues from modern humans to archaic humans. Highly correlated HS TF-target transcript pairs are distributed across many GTEx tissues, rather than being confined to the brain; also, significantly changed HS TF DBSs do not occur densely in the brain (Supplementary Note 8). These results further support that HS lncRNAs have promoted human brain evolution more than HS TFs.

HS lncRNAs mediate human-specific correlated gene expression in the brain

Finally, we further examined whether the gene expression pattern in the brain is human-specific. Based on the GTEx data from the human frontal cortex (BA9) and anterior cingulate cortex (BA24) (n=101 and n=83, respectively) (Consortium et al., 2017), and the gene expression data for the same brain regions in macaques (n=22 and n=25, respectively) (Zhu et al., 2018), we identified transcriptional regulation modules used the eGRAM program. In the two human brain regions, HS lncRNAs’ target genes form distinct modules, characterized by highly correlated gene expression (Pearson’s r>0.8) and enriched for KEGG pathways related to neurodevelopment (hypergeometric distribution test, FDR<0.05). In the same macaque brain regions, the orthologues of these human genes lack these features (Figure 4; Supplementary Note 9; Supplementary Table 16). The differences in gene modules suggest that the gene expression pattern in the human brain is highly human-specific and, to a great extent, can be attributed to the regulation by HS lncRNAs.

Human-specifically reshaped gene expression by HS lncRNAs in the frontal cortex (BA9).

(A) Genes expressed in the human frontal cortex are enriched for HS lncRNAs’ target genes and neurodevelopment-related pathways. Squares, dots, and colors indicate HS lncRNAs, gene modules (Module_1 and Module_2 are illustrated), and enriched KEGG pathways, respectively. (B) Comparison of modules and genes in humans (indicated by H) and macaques (indicated by M). In each pair of modules, green and blue dots denote human genes and their orthologues, and lines between dots indicate correlated expression. Many orthologous genes in macaques (displayed at the corresponding positions) are not in the modules, and correlated expression is more prominent in humans than in macaques.

Discussion

The limited genomic but substantial phenotypic and behavioral differences between modern humans, archaic humans, and apes make “what genomic differences critically determine modern humans” a profound question. This question has been addressed by many studies using various methods. Especially, studies examining protein-coding genes and HARs reported that HS protein-coding genes promote rapid neocortex expansion (Fiddes et al., 2018; Florio et al., 2018; Pinson et al., 2022) and that HARs are significantly enriched in genomic regions important for human-specific 3D genome organization (Keough et al., 2023). Many studies also examined lncRNAs in humans and mice. However, reports on HS lncRNAs and their DBSs have been rare. Similar to gene expression reshaping or rewiring by lineage-specific TFs and their DBSs (Johnson, 2017), gene expression can also be substantially rewired by lineage-specific lncRNAs and their DBSs. This study examined the postulation that HS lncRNAs and their DBSs have significantly reshaped gene expression during human evolution.

The nature of large-scale multi-omics data analysis means that the main results are unlikely to be generated by chance, although different parameters, thresholds, and significance levels can yield somewhat different results. Given the importance of DBS prediction, we used multiple methods, including DBD KO following differential gene expression analysis, to validate it. When predicting and analysing DBSs, we examined DBS in every transcript; thus, transcripts (including those in the brain) whose DBSs have had significant human-specific evolution can be identified. When running programs including Infernal, LongTarget, and g:Profiler and performing statistical tests that detect selection signals, we used the default parameters that suit most situations and the widely used significance levels. To compare the target genes of HS lncRNAs and their counterparts and orthologues in archaic humans and chimpanzees, we used multiple ORA-based methods and explored the reported ASE genes in tetraploid hybrid human-chimpanzee stem cells (Agoglia et al., 2021). Since weak DBSs reflect recent evolution but may be less reliable, we classified DBSs into strong/weak and old/young classes and found that different classes have distinct features. Understandably, not all detected signals reliably indicate positive selection; therefore, we also examined and described SNPs.

A few notes on several findings. First, relevant to and in line with the finding that Neanderthal-inherited sequences have measurable effects on gene expression in modern humans, but these effects are least detectable in the brain (McCoy et al., 2017), we find that HS lncRNAs and their DBSs have influenced gene expression in the brain more significantly than in other tissues. Second, the evolution of gene regulation by HS lncRNAs is rapid, and notable examples include those that enable humans to adapt to high-sugar intake. Third, regulated genes are enriched for different functions at different periods of human evolution, as evidenced by genes with young/old DBSs and with DBSs that contain “old” and “new” SNPs (e.g., DBSs containing “old” and “new” SNPs are in genes regulating neural development and glucose metabolism). Fourth, newly emerged regulations (possibly indicated by young weak DBSs) may have promoted recent human evolution.

Our findings also raise new questions. First, since many mouse lncRNAs are rodent-specific (Yue et al., 2014), two questions are how these lncRNAs specifically rewire gene expression in mice, and to what extent human- and mouse-specific lncRNAs cause differences in cross-species transcriptional regulation (Breschi et al., 2017; Hodge et al., 2019; Zhu et al., 2018). Second, whether mice and other mammals have such species-specific lncRNAs as RP11-423H2.3 that regulate the expression of substantial genes. Third, whether lineage- and species-specific lncRNAs would make many evolutionary novelties preordained.

Material and methods

Data sources

The sequences of human (GRCh38/hg38, GRCh37/hg19) and chimpanzee genomes (panTro5) were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). Three high-quality archaic human genomes were obtained from the Max Planck Institute for Evolutionary Anthropology (https://www.eva.mpg.de/genetics/index.html), which include an Altai Neanderthal that lived approximately 122 thousand years ago (kya) (Pru fer et al., 2013), a Denisovan (an Asian relative of Neanderthals) that lived approximately 72 kya (Meyer et al., 2012), and a Vindija Neanderthal that lived approximately 52 kya (Prufer et al., 2017). The ancestral sequences for the human genome (GRCh37) (which were generated upon the EPO alignments of six primate species human, chimpanzee, gorilla, orangutan, macaque, and marmoset) were downloaded from the EBI website (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/). The SNP data of modern humans were obtained from the 1000 Genomes Project (phase I) (1000 Genomes Project Consortium et al., 2012). The three modern human populations are CEU (Utah residents with Northern and Western European ancestry), CHB (Han Chinese in Beijing, China), and YRI (Yoruba in Ibadan, Nigeria), which contain 100, 99, and 97 individuals, respectively. The genetic map files of modern humans were downloaded from the 1000 Genomes Project website (ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/technical/working/20110106_recombination_hotspots/). The 1000 Genomes phased genotype data in VCF format were downloaded from the International Genome Sample Resource website (http://www.internationalgenome.org/). Annotated human genes and transcripts were obtained from the Ensembl website (http://www.Ensembl.org). FASTA-format nucleotide sequences were converted from VCFs using the vcf-consensus program in the VCFtools package when necessary (Danecek et al., 2011). Data in the GeneCards Human Gene Database (www.genecards.org) were used to annotate genes and transcripts.

RNA-sequencing (RNA-seq) data (transcripts per million [TPM]) and eQTL data of human tissues were obtained from the Genotype-Tissue Expression Project (GTEx, v8) website (https://gtexportal.org/) (GTEx Consortium, 2017). Histone modification and DNA methylation signals in multiple cell lines, and ENCODE Candidate Cis-Regulatory Elements (cCRE), were obtained from the UCSC Genome Browser (http://genome.UCSC.edu). The genome-wide DNA-binding sites of lncRNA NEAT1, MALAT1, and MEG3 were obtained from experimental studies (Mondal et al., 2015; West et al., 2014). The predicted and experimentally detected DBSs in target genes of MALAT1, NEAT1, and MEG3 are given in the supplementary table (Supplementary Table 17).

Identifying HS lncRNA genes

We used the Infernal program (Nawrocki et al., 2009), which searches orthologous RNA sequences upon sequence and structure alignment, to identify orthologous exons in 16 mammalian genomes for each exon in each of the 13562 GENCODE (v18)-annotated human lncRNA genes (Lin et al., 2019). The 16 mammals were chimpanzee (CSAC 2.1.4/panTro4), macaque (BGI CR_1.0/rheMac3), marmoset (WUGSC 3.2/calJac3), tarsier (Broad/tarSyr1), mouse lemur (Broad/micMur1), tree shrew (Broad/tupBel1), mouse (GRCm38/mm10), rat (Baylor3.4/rn4, RGSC6.0/rn6), guinea pig (Broad/cavPor3), rabbit (Broad/oryCun2), dog (Broad CanFam3.1/canFam3), cow (Baylor Btau_4.6.1/bosTau7), elephant(Broad/loxAfr3), hedgehog (EriEur2.0/eriEur2), opossum (Broad/monDom5), and platypus (WUGSC 5.0.1/ornAna1) (http://genome.UCSC.edu). If the number of orthologous exons of a human lncRNA gene in a genome exceeded half the exon number of the human lncRNA gene, these orthologous exons were assumed to form an orthologous lncRNA gene. If a human lncRNA gene had no orthologous gene in all of the 16 mammals, it was assumed to be human-specific.

Identifying DBSs of HS lncRNAs

LncRNAs bind to DNA sequences by forming RNA:DNA triplexes. Each triplex comprises triplex-forming oligonucleotides (TFO) in the lncRNA and a triplex-targeting site (TTS) in the DNA sequence. We used the LongTarget program to predict HS lncRNAs’ DNA binding domains (DBD) and binding sites (DBS) with the default parameters (Ruleset = all, T.T. penalty = −1000, CC penalty = 0, Offset = 15, Identity ≥ 60, Nt ≥ 50) (Lin et al., 2019; Wen et al., 2022). The LongTarget program simultaneously predicts DBDs and DBSs, where a DBD comprises a set of densely overlapping TFOs and a DBS comprises a set of densely overlapping TTSs. For each HS lncRNA, we predicted its DBSs in the 5000 bp promoter regions (−3500 bp upstream and +1500 bp downstream the transcription start site) of the 179128 Ensembl-annotated transcripts (release 79). For each DBS, its binding affinity is the product of DBS length and the averaged Identity score of all TTSs (the Identity score is the percentage of paired nucleotides). Strong and weak DBSs were classified based on binding affinity ≥60 and <60. A transcript whose promoter region contains a strong DBS of an HS lncRNA was assumed to be a target transcript of the HS lncRNA, and the gene containing this transcript was assumed to be a target gene of the HS lncRNA. As the 1000 Genomes Project (phase I) data and the archaic human genomes are based on GRCh37/hg19, the DBS coordinates were converted from GRCh38/hg38 to GRCh37/hg19 using the liftover program from the UCSC Genome Browser (Kuhn et al., 2013).

Experimentally validating DBS prediction

A 157 bp sequence (chr17:80252565-80252721, hg19) containing the DBD of RP13-516M14.1, a 202 bp sequence (chr1:113392603-113392804, hg19) containing the DBD of RP11-426L16.8, and a 198 bp sequence (chr17:19460524-19460721, hg19) containing the DBD of SNORA59B, were knocked out in the HeLa cell line, RKO cell line, and SK-MES-1 cell line, respectively. Two sequences (chr1:156643524-156643684, chr10:52445649-52445740, hg38) containing the DBD of two wrongly transcribed noncoding sequences were knocked out in the HCT-116 and A549 cell lines, respectively. The seven knockouts were performed by UBIGENE, Guangzhou, China (http://www.ubigene.com) using CRISPR-UTM, a revised version of CRISPR/Cas9 technology. Before and after the seven DBD knockouts, RNA sequencing (RNA-seq) was performed by Novogene, Beijing, China (https://cn.novogene.com) and HaploX, Shenzhen, China ( https://www.haplox.cn/). The reads were aligned to the human GRCh38 genome using the Hiasat2 program (Kim et al., 2019), and the resulting SAM files were converted to BAM files using Samtools (Li et al., 2009). The Stringtie program was used to quantify gene expression levels (Pertea et al., 2015). Fold change of gene expression was computed using the edgeR package (Robinson et al., 2010), and significant up- and down-regulation of target genes after DBD knockout was determined upon |log2(fold change)| > 1 with FDR < 0.1.

Genome-wide DBSs of NEAT1, MALAT1, and MEG3 were experimentally detected (Mondal et al., 2015; West et al., 2014). We also used these data to validate DBS prediction by predicting DBSs of the three lncRNAs and checking the overlap between predicted and experimentally detected DBSs (Supplementary Table 17).

Mapping DBSs of HS lncRNAs in the chimpanzee and archaic human genomes

We used the liftover program from the UCSC Genome Browser to map DBS loci from the human genome (hg38) to the chimpanzee genome (Pan_tro 3.0, panTro5). The mapping results were verified by inspecting the human-chimpanzee pairwise alignment in the UCSC Genome Browser. This initial screening identified 2248 DBSs (residing in 429 genes) that could not be mapped to the chimpanzee genome. To definitively determine whether these unmapped DBSs represent human-specific gains or chimpanzee-specific losses, we analyzed their sequences using the UCSC Multiz Alignments of 100 Vertebrates. This comparative genomics analysis confirmed that all 2248 DBSs are present in the human genome but are absent from the chimpanzee genome and all other aligned vertebrate genomes. Therefore, we classified these DBSs as human-specific gains.

We used vcf-consensus in the VCFtools package to extract the DBSs of HS lncRNAs from the VCF files of Altai Neanderthals, Denisovans, and Vindija Neanderthals. The variant with the highest quality score was selected whenever multiple variant calls were observed at a given locus. The obtained DBS sequences in chimpanzees and three archaic humans are called counterparts of DBSs in these genomes.

Estimating sequence distances of DBSs between different genomes

We first aligned DBS sequences in the genomes of humans, chimpanzees, Altai Neanderthals, Denisovans, and Vindija Neanderthals using the MAFFT7 program to measure sequence distances from modern humans to chimpanzees and archaic humans (Katoh and Standley, 2013). We then computed sequence distances using the dnadist program with the Kimura 2-parameter model in the PHYLIP (3.6) package (http://evolution.genetics.washington.edu/phylip.html) and the Tamura-Nei model in the MEGA7 package (Kumar et al., 2016). The two methods generated equivalent results. The largest distance between DBSs in humans and their chimpanzee counterparts is 5.383. Since 2248 DBSs in 429 human genes lack counterparts in chimpanzees, we assumed that these DBSs have a sequence distance of 10.0 between humans and chimpanzees.

We determined human ancestral sequences of DBSs using the human ancestor sequences from the EBI website, which were generated from the EPO alignments of six primate species. We used the above-mentioned methods to calculate the sequence distances DBSs from the human ancestor to chimpanzees, archaic humans, and modern humans. We found that when the human-chimpanzee ancestral sequence has the ancestral sequence (which means the inference of ancestral allele is of high confidence), DBS distances from the human ancestor to modern humans are larger than to archaic humans, but this situation accounts for only about 63.8%. For many DBSs, the distances from the human ancestor to modern humans are smaller than to archaic humans (especially Neanderthals and Denisovans) and even to chimpanzees. This defect may be caused by the absence of archaic humans in building the human ancestral sequence.

Detecting positive selection signals in HS lncRNA genes and DBSs

We used multiple tests to detect positive selection signals in HS lncRNA genes. First, we used the XP-CLR test (parameters = −w1 0.001 300 100 −p0 0.95, window size = 0.1 cM, grid size = 100 bp) to perform six pairwise genome-wide scans (i.e., CEU-CHB, CEU-YRI, CHB-CEU, CHB-YRI, YRI-CEU, and YRI-CHB) (Chen et al., 2010). The upper 1% of scores across the entire genome in each pairwise scan was 34.6 in the CEU-YRI scan, 16.8 in the CEU-CHB scan, 45.0 in the CHB-YRI scan, 26.9 in the CHB-CEU scan, 14.1 in the YRI-CEU scan, and 14.1 in the YRI-CHB scan. These scores were used as the thresholds of positive selection signals in these populations. Second, we used the iSAFE program to scan each genomic region containing an HS lncRNA gene and its 500-kb upstream and downstream sequences (Akbari et al., 2018). Strongly selected loci were detected only in CEU and CHB. Third, we used the VCFtools program to calculate Tajima’s D values for each HS lncRNA gene in CEU, CHB, and YRI (Danecek et al., 2011). The calculation was performed using a 1500-bp non-overlapping sliding window because the lengths of these genes exceed 1500 bp. To generate a background reference for assessing significant increases or decreases in Tajima’s D for HS lncRNA genes in a population, we calculated Tajima’s D across the whole genome using a sliding window of 1500 bp. As the values of Tajima’s D were compared with the background reference, significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively, rather than population demography dynamics (Tajima, 1989). Fourth, we used the integrated Fst to detect positive selection signals in HS lncRNA genes. The Fst of each HS lncRNA gene was computed for three comparisons, i.e., CEU-YRI, CHB-YRI, and CHB-CEU. Extreme Fst values of SNPs were detected in HS lncRNA genes in the comparisons of CEU-YRI and CHB-YRI. Since allele frequencies for different loci vary across a genome and genetic drift may have different effects at different loci, we used a sliding window of 1500 to compare Fst values of HS lncRNA genes with the genome-wide background. Extreme Fst values indicate positive selection. Finally, we applied linkage disequilibrium (LD) analysis to each HS lncRNA gene. We computed the pairwise LD (r²) in CEU, CHB, and YRI for common SNPs (with minimal allele frequency (MAF) ≥ 0.05 in at least one population) in HS lncRNA genes and DBSs using the PLINK program (Purcell et al., 2007). Significantly increased LD was detected in SNPs in HS lncRNA genes in CEU and CHB. The LD patterns were portrayed using the Haploview program (Barrett et al., 2005).

Next, we used the above tests to detect positive selection signals in DBSs. First, the 100 bp grid size of the XP-CLR test also allowed the detection of selection signals in DBSs. Second, we performed Tajima’s D and Fay-Wu’s H tests (Fay and Wu, 2000; Tajima, 1989). We calculated Tajima’s D values for each DBS in CEU, CHB, and YRI using the VCFtools program (Danecek et al., 2011), with a sliding window of 147 bp (the mean length of strong DBSs). To generate a background reference for judging the significant increase or decrease of Tajima’s D in a population, we calculated Tajima’s D values across the whole genome using the same sliding window. When Tajima’s D values were compared with the background reference, significant D<0 and D>0 indicate positive (or directional) selection and balancing selection, respectively. Fay-Wu’s H values were calculated similarly using the VariScan program (Vilella et al., 2005). Calculating Fay-Wu’s H demands the ancestral sequences as the outgroup. We extracted the ancestral sequences of DBSs from the human ancestral sequence, which was generated upon the EPO alignments of six primate species (human, chimpanzee, gorilla, orangutan, macaque, and marmoset) (ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/). Third, we computed the Fst to measure the frequency differences of alleles in DBSs between populations. For the CEU-YRI, CHB-YRI, and CHB-CEU pairwise comparisons, we used the revised VCFtools program to compute the weighted Fst for all SNPs in each DBS (Weir and Cockerham, 1984). Fourth, we integrated the weighted Fst values in the three populations into an “integrated Fst” which indicated whether the DBS locus was under selection in a certain population (Nielsen and Slatkin, 2013). We used sliding windows of 147 bp for comparing Fst values of DBSs with the genome-wide background. We empirically defined the upper 10% of integrated Fst scores across the entire genome as statistically significant. To determine positive selection more reliably, we used Tajima’s D and the integrated Fst to jointly determine if a DBS was under positive selection in a population. The thresholds that determined the upper 10% of Tajima’s D values across the entire genome in CEU, CHB, and YRI were −0.97, −0.96, and −0.97, respectively, and the threshold that determined the upper 10% of integrated Fst values across the entire genome in the three populations was 0.22. For example, a DBS was assumed to be under positive selection in CEU if (a) the DBS had a Tajima’s D<-0.97 in CEU and Tajima’s D>0.0 in the two other populations and (b) the DBS had an integrated Fst>0.22. Analogously, a DBS was assumed to be under positive selection in both CEU and CHB if the DBS had a Tajima’s D <-0.97 in CEU, <-0.96 in CHB, and >0.0 in YRI, and had an integrated Fst>0.22.

Functional enrichment analysis of genes

We used the g:Profiler program (with the parameter settings: Organism=Homo sapiens, Ordered query = No, Significance threshold = Benjamini-Hochberg FDR, User threshold = 0.05, 50<terms size<1000) and the Gene Ontology (GO) database to perform over-representation analysis (Raudvere et al., 2019). This analysis determines which pre-defined gene sets (GO terms) are more prevalent (over-represented) in a list of “interesting” genes than would be expected by chance. The lists of genes included genes with strong and weak DBSs, genes with large and small DBS distances from humans to chimpanzees, and genes with large and small DBS distances from humans to archaic humans. Strong DBSs have top affinity, and genes with weak DBSs not only have DBS affinity≤40 but also have DBSs of >=5 HS lncRNAs to help ensure that these genes are likely HS lncRNAs’ targets.

Analysing SNP frequencies in human populations

The frequencies of common SNPs (MAF ≥ 0.05) in DBSs across the three modern human populations were computed using the VCFtools package (Danecek et al., 2011). The ancestral/derived states of SNPs were inferred from the human ancestor sequences and were used to determine derived allele frequencies.

Analysing the cis-effect of SNPs in DBSs on target gene expression

SNPs with MAF >0.1 in DBSs in any of the three modern human populations and absolute values of cis-effect size >0.5 (FDR <0.05) in any of the GTEx tissues were examined for an influence on the expression of the target genes (GTEx Consortium, 2017). SNPs that are eQTLs in the GTEx tissues and have biased derived allele frequencies (DAF) in the three modern human populations were examined to estimate whether the eQTL is population-specific.

Examining the tissue-specific impact of HS lncRNA-regulated gene expression

First, we examined the expression of HS lncRNA genes across the GTEx tissues. HS lncRNA genes with a median TPM value >0.1 in a tissue were considered robustly expressed in that tissue. Upon this criterion, 40 HS lncRNA genes were expressed in at least one tissue and were used to examine the impact of HS lncRNA regulation on gene and transcript expression (other HS lncRNAs may function in the cytoplasm). Since an HS lncRNA gene may have multiple transcripts, we selected the transcript containing the predicted DBD and with the highest TPM as the representative transcript of the HS lncRNA. We calculated the pairwise Spearman’s correlation coefficient between the expression of an HS lncRNA (the representative transcript) and the expression of each of its target transcripts using the scipy.stats.spearmanr program in the scipy package. The expression of an HS lncRNA and a target transcript was considered to be significantly correlated if the |Spearman’s rho|>0.3, with Benjamini-Hochberg FDR<0.05. We examined the percentage distribution of correlated HS lncRNA-target transcript pairs across GTEx tissues and organs (Figure 3A).

We examined all GTEx tissues to determine which tissues may have exhibited significant changes in HS lncRNA-regulated gene expression from archaic to modern humans. GTEx data include both gene expression matrices and transcript expression matrices; we used the latter to examine changes in HS lncRNA-target transcript pairs from modern humans to archaic humans. If a pair of HS lncRNA and target transcript is robustly expressed in a tissue and their expression shows a significant correlation (|Spearman’s rho|>0.3, with Benjamini-Hochberg FDR<0.05) in the tissue, we computed the sequence distance of the HS lncRNA’s DBS in the transcript from the three archaic humans to modern humans. We compared the sequence distances of all DBSs in each tissue with the sequence distances of all DBSs in all GTEx tissues (as the background). A one-sided two-sample Kolmogorov-Smirnov test was used to examine whether the sequence distances of all DBSs in a specific tissue deviate from the background distribution (which reflects the “neutral evolution” of gene expression). For each tissue, if the Benjamini-Hochberg FDR was <0.001, the tissue was considered to have significantly altered gene expression regulated by the HS lncRNA. We used different colors to mark tissues with significantly changed gene expression regulation since Altai Neanderthals, Denisovans, and Vindija Neanderthals, and used “D”, “A.D.”, and “ADV” to indicate changes since Denisovans, since Altai Neanderthals and Denisovans, and since Altai Neanderthals, Denisovans, and Vindija Neanderthals, respectively (Figure 3B).

Examining enrichment of favored and hitchhiking mutations in DBSs

Using the deep learning network DeepFavored, which integrates multiple statistical tests for identifying favored mutations (Tang et al., 2022), we identified 13339 favored mutations and 244098 hitchhiking mutations in 17 human populations (Tang et al., 2023). In this study, we classified DBSs in two ways: into strong ones (affinity>60) and weak ones (36<affinity<60), and into old ones (Human-Chimp distance>0.034 AND Human-Altai Neanderthals distance=0), young ones (Human-Altai Neanderthals distance>0.034 OR Human-Denisovan distance>0.034), and others. We then examined the number of favored and hitchhiking mutations in each class of DBSs. The weak young DBSs have the largest proportion of favored and hitchhiking mutations.

The analysis of HS TFs and their DBSs

Kirilenko et al. recently identified orthologous genes in hundreds of placental mammals and birds and organized genes into pairwise datasets using humans and mice as the references (e.g., “hg38-panTro6”, “hg38-mm10”, and “mm10-hg38”) (Kirilenko et al., 2023). Based on the “many2zero” and “one2zero” gene lists (which contain 0 and 147 genes, respectively) in the hg38-panTro6 dataset, which indicate multiple human genes and a single human gene that have no orthologues in chimpanzees, we identified HS protein-coding genes. Further, based on three human TF lists reported by two studies and used in the SCENIC package (Bahrami et al., 2015; Lambert et al., 2018), we identified HS TFs. Using the JASPAR database and the CellOracle program (with default parameters), we predicted DBSs for HS TFs. Then, we repeated the steps of our HS lncRNA analyses (Supplementary Note 8; Supplementary Table 15), including computing sequence distances of DBSs and examining the impact of HS TF-target transcript pairs on gene expression in GTEx tissues and organs.

Identifying and analyzing transcriptional regulatory modules

Most clustering algorithms classify genes into disjoint modules based on expression correlation without considering regulatory relationships (Saelens et al., 2018). The GRAM program identifies gene regulatory modules based on correlation and TF-TFBS binding (Bar-Joseph et al., 2003). LncRNAs transcriptionally regulate genes based on lncRNA-DBS binding and correlate gene expression. We developed the eGRAM program to identify gene modules based on correlated expression, TF-TFBS interaction, and lncRNA-DBS interaction. In this study, we used eGRAM to identify gene modules in the same regions of the human and macaque brains and enriched KEGG pathways using reported RNA-seq datasets (n=101 and 83 for frontal cortex and n=22 and 25 for anterior cingulate cortex) (GTEx Consortium, 2017; Zhu et al., 2018). The default parameters, DBS binding affinity=60, Pearson correlation=0.5, module size=50, and FDR=0.01 (hypergeometric distribution test), were used. The key steps of the program are as follows. (a1) Identify each lncRNA’s correlated lncRNAs, which may form a set of co-regulators. (a2, optional) Identify each TF’s correlated TFs, which may form a set of co-regulators. (b1) Compute the correlation between each lncRNA and all genes. (b2, optional) Compute the correlation between each TF and all genes. (c1) Identify each lncRNA’s target genes. (c2, optional) Identify each TF’s target genes. (d1) Identify each lncRNA set’s target module upon correlation and targeting relationships. (d2, optional) Identify each TF set’s target module upon correlation and targeting relationships. (e) Check whether TFs’ modules contain lncRNAs’ targets and whether lncRNAs’ modules contain TFs’ targets, which reveal genes co-regulated by TFs and lncRNAs and genes independently regulated by TFs and ncRNAs. (f) Performs pathway enrichment analysis for all modules.